---
title: "Mexicans in Chicago Report Committee"
author: ""
date: "Great Cities Institute"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    highlight: pygments
    extra_dependencies: ["float"]
    fig_caption: true
mainfont: "Arial"
header-includes:
  - \usepackage{titling}
  - \pretitle{\begin{flushleft}}
  - \posttitle{\end{flushleft}}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
editor_options: 
  chunk_output_type: console
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F, fig.pos="H")


```


```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}

library(ipumsr)
library(tidyverse)
library(purrr)
library(sf)
library(tidycensus)
library(tidyr)
library(readxl)
library(sf)
library(htmltools)
library(leaflet)
library(janitor)
library(data.table)
library(tinytex)
library(survey) # for survey analysis
library(srvyr)  # for tidy survey analysis
library(tmap)
library("kableExtra") 
library(tmaptools)
library("RColorBrewer") 
# library(OpenStreetMap)
# remotes::install_github('yihui/knitr')
# install.packages("knitr")
# library(knitr)
# options(tinytex.verbose = TRUE)
# tinytex::reinstall_tinytex()
# install.packages('tinytex')

# update.packages(ask = FALSE, checkBuilt = TRUE)
# tinytex::tlmgr_update()
# tinytex::latexmk("Housing_Visual.tex")
options(knitr.kable.NA = '-') 

# Read Data

usa_data <- read.csv("C:/Users/dsegovi2/Box/Great Cities Institute/Research/Mexican Report/final_Mexican_IL_2000_22.csv") %>% 
  mutate(race_ethnicity = case_when(hispan ==0 & race == 1 ~ "White (non-Hispanic or Latino)",
                                    hispan ==0 & race == 2 ~ "Black (non-Hispanic or Latino)",
                                    hispan %in% c(2,3,4) ~ "Other Hispanics or Latinos",
                                    hispan ==1 ~ "Mexican", 
                                    hispan ==0 & race %in%c(3:9) ~ "Other (non-Hispanic or Latino)",
                                    TRUE ~ NA_character_),
         
  race_ethnicity = factor(race_ethnicity, level = c("Mexican", "Other Hispanics or Latinos","White (non-Hispanic or Latino)", "Black (non-Hispanic or Latino)",  "Other (non-Hispanic or Latino)")),
  
  hispan_breakdown = case_when(
  hispand == 0 ~ "Not Hispanic",
  hispand == 100 ~ "Mexican", 
  hispand == 200 ~ "Puerto Rican",
  hispand == 300 ~ "Cuban",
  hispand == 411 ~ "Costa Rican",
  hispand == 412 ~ "Guatemalan",
  hispand == 413 ~ "Honduran",
  hispand == 414 ~ "Nicaraguan",
  hispand == 415 ~ "Panamanian",
  hispand == 416 ~ "Salvadoran",
  hispand == 417 ~ "Central American, not specified",
  hispand == 420 ~ "Argentinean",
  hispand == 421 ~ "Bolivian",
  hispand == 422 ~ "Chilean",
  hispand == 423 ~ "Colombian",
  hispand == 424 ~ "Ecuadorian",
  hispand == 425 ~ "Paraguayan",
  hispand == 426 ~ "Peruvian",
  hispand == 427 ~ "Uruguayan",
  hispand == 428 ~ "Venezuelan",
  hispand == 431 ~ "South American, not specified",
  hispand == 450 ~ "Spaniard",
  hispand == 460 ~ "Dominican",
  hispand == 498 ~ "Other, not specified",
  TRUE ~ "Other"
),
mex_foreign_born = case_when(
    race_ethnicity == "Mexican" & bpl %in% as.character(1:56) ~ "U.S. Born",
    race_ethnicity == "Mexican" & bpl == "200" ~ "Mexico Born",
    TRUE ~ NA_character_  # Optional: default case for other values or groups
  )) %>% ## filter for hispanic groups here ##
 mutate(hispan_breakdown = factor(hispan_breakdown, level = c("Mexican", "Puerto Rican","Ecuadorian", "Cuban", "Guatemalan", "Colombian"))
  )


# Construct the full file path
chicago_ca_pop <- read_sf("Data Tables/shp/comm_pop_chi_2022.shp") %>% as.data.frame() 

##### Chicago boarder
Chicago_boundary_shp <- read_sf("C:/Users/dsegovi2/Box/Great Cities Institute/Research/Mexican Report/Boundaries_Chicago/geo_export_8881adc5-2460-457b-892f-04f4d6028b72.shp") %>% 
  st_transform(crs = 4326)
# mapview(community_shp)

# Get the current bounding box of the data
bbox_data <- st_bbox(Chicago_boundary_shp)

# Calculate the range of x and y values in degrees
xrange <- bbox_data$xmax - bbox_data$xmin
yrange <- bbox_data$ymax - bbox_data$ymin

# Calculate the center of the bounding box
center_x <- (bbox_data$xmin + bbox_data$xmax) / 2
center_y <- (bbox_data$ymin + bbox_data$ymax) / 2
# Modify the bounding box to fit the desired width while keeping the height the same
bbox_new <- c(center_x - 0.24,
              center_y - 0.23,
              center_x + 0.23,
              center_y + 0.23)

# osm_chicago <- read_osm(bbox_new, 
#                         type = "bing")



theme_plots <-
  theme_minimal() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.line = element_line(color="#969696", linewidth = 1),
        plot.title = element_text(face ="bold", size = 10, hjust = 0.5),
      plot.subtitle = element_text(face ="italic", size = 8, color = "#636363"),
      axis.text = element_text(size = 9),
      axis.title = element_text(face ="bold", size = 9),
      plot.caption = element_text(hjust = 0, size = 8, color = "black"),
      legend.text = element_text(size = 8, color = "#636363"),
      legend.title = element_text(face = "bold", size = 8, color = "#636363")) 

color_race_ethnicity <- c("Mexican" = "#CE1126", "Other Latinos" = "#F9AA51",
                               "White (non-Hispanic or Latino)" = "#9CA168",
                               "Black (non-Hispanic or Latino)" = "#8CBEC0",
                               "Other (non-Hispanic or Latino)" = "#0B8489")

school_indicator <- read_csv("School/df_school.csv")


chicago_health <- read_csv("Health Data/Chicago Health Atlas Data Download - Community areas.csv") %>% clean_names() %>% rename(
  obesity_rate = hcsobp_2022_2023,
  diabetes_rate = hcsdiap_2022_2023,
  asthma_rate = hcsathp_2022_2023,
  lbw_rate = vrbwp_2017_2021,
  unmet_mental_health_rate = hcsumhap_2022_2023,
  cancer_diagnosis_rate = ccr_2017_2021,
  hyper_tension_rate = hcshytp_2022_2023,
  psych_distress_rate =   hcsspdp_2022_2023,
  lung_cancer_rate =ccg_2017_2021,
  loneliness_rate = chavqos_2022_2023
)


# merge chicago ca with health atlas
df_health <- chicago_ca_pop %>% select(commnty, prc_mxc, prc_nh_w, prc_nh_b) %>% mutate(commnty = str_to_title(tolower(commnty))) %>% left_join(chicago_health, by = c("commnty" = "name")) %>% select(-layer)  %>%
 mutate(across(matches("rate$"), ~ format(round(as.numeric(.), 1), nsmall = 1))) %>% mutate(across(ends_with("rate"), ~ as.numeric(as.character(.))))

comm_2000_2022 <- st_read ("Data Tables/shp/comm_pop_chi_2000_2022.shp" ) %>% 
  mutate(prc2 = paste0(prc, "%"))




```


\renewcommand{\arraystretch}{1.4}

\clearpage


\blandscape
 

# Demographic



## 1B(1): Population by Age for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates) 

```{r}
# analysis

# filter for year 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)

# 1B: Population pyramid of the percentage of Mexican population by age group compared to the rest of the population in Chicago and median age in Chicago (Data source: 2018-2022 ACS data)

df <-  data_chi_2018_22 %>%
  mutate(
    age_group = case_when(
      age < 20 ~ "Young (0-19)",
      age >= 20 & age < 40 ~ "Adults (20-39)",
      age >= 40 & age < 60 ~ "Middle-aged (40-59)",
      age >= 60 & age < 80 ~ "Older Adults (60-79)",
      age >= 80 ~ "Elderly (80+)"
    )
  )
# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)


# Calculate weighted population counts by age group and ethnicity

# Calculate weighted population by ethnicity and age group
pop_pyramid <- df %>%
  as_survey_design(weights = perwt) %>% 
  survey_count(race_ethnicity, age_group, name="total_weighted_count")  %>%
  group_by(race_ethnicity) %>%
  mutate(total_weighted_percentage = total_weighted_count / sum(total_weighted_count) * 100,
         total_weighted_percentage  = paste0(format(round(total_weighted_percentage, 1), nsmall = 1), "%"))



# get median age

weighted_age  <- df %>%
  group_by(race_ethnicity) %>%
  summarize(
    weighted_age = matrixStats::weightedMedian(age, perwt)
  ) 



# combined

df_final  <- pop_pyramid %>% left_join(weighted_age)

```




```{r}
# visual

table_1b.1 <- df_final %>% 
  mutate(age_group = factor(age_group, levels = c("Young (0-19)", "Adults (20-39)", "Middle-aged (40-59)", "Older Adults (60-79)", "Elderly (80+)"))) %>% 
  arrange(age_group) %>%  # Explicitly arrange by factor levels
  pivot_wider(id_cols = c("age_group"), names_from = race_ethnicity, values_from = c("total_weighted_count", "total_weighted_percentage"))  %>% 
  select(age_group, 
         total_weighted_count_Mexican, total_weighted_percentage_Mexican,
         'total_weighted_count_Other Hispanics or Latinos', 'total_weighted_percentage_Other Hispanics or Latinos',
         'total_weighted_count_White (non-Hispanic or Latino)', 'total_weighted_percentage_White (non-Hispanic or Latino)',
         'total_weighted_count_Black (non-Hispanic or Latino)', 'total_weighted_percentage_Black (non-Hispanic or Latino)',
         'total_weighted_count_Other (non-Hispanic or Latino)', 'total_weighted_percentage_Other (non-Hispanic or Latino)') %>%
  
  # using kable 
  kbl(booktabs = T,
      col.names = c("", "Number", "Percent", "Number", "Percent", "Number", "Percent", "Number", "Percent", "Number", "Percent"),
      align = c("l", rep("r", 10)),  # Align headers and columns
      caption = "Population by Age for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "") %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:11, width = "3.08em") %>%  # Align all columns with numbers and percentages to the right
  row_spec(0, align = "C") %>%  # Center column headers
 add_header_above(
  align = c("l", rep("c", 5)),  # Assuming 1 for Age Group and 10 divided into 5 groups
  c("Age Group", "Mexican" = 2, "Other Hispanics\n or Latinos" = 2, "White\n (non-Hispanic\n or Latino)" = 2, "Black\n (non-Hispanic\n or Latino)" = 2, "Other\n (non-Hispanic\n or Latino)" = 2),
  bold = TRUE, font_size = 8
) %>% 
  add_footnote(
    c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
    notation = "none",
    threeparttable = TRUE,
    escape = FALSE
  )
table_1b.1 
```


\elandscape



```{r fig.width=6, fig.height=5 ,fig.align='center'}
# visual

# Pivot data to long format and clean age_group names
df_long <- df_final %>%
  pivot_wider(id_cols = c("race_ethnicity"), 
              names_from = age_group, 
              values_from = c("total_weighted_count", "total_weighted_percentage")) %>% 
  select(race_ethnicity, 
         'total_weighted_percentage_Young (0-19)',
         'total_weighted_percentage_Adults (20-39)',
         'total_weighted_percentage_Middle-aged (40-59)',
         'total_weighted_percentage_Older Adults (60-79)',
         'total_weighted_percentage_Elderly (80+)') %>%
  pivot_longer(cols = starts_with("total_weighted_percentage_"), 
               names_to = "age_group", 
               values_to = "percentage") %>%
  mutate(age_group = gsub("total_weighted_percentage_", "", age_group),
         percentage = as.numeric(gsub("%", "", percentage))) # Convert percentage to numeric

# Reverse the order of age groups
df_long <- df_long %>%
  mutate(age_group = factor(age_group, 
                            levels = c("Young (0-19)", "Adults (20-39)", "Middle-aged (40-59)", "Older Adults (60-79)", "Elderly (80+)")))

# Define the custom colors for race_ethnicity
color_race_ethnicity <- c("Mexican" = "#CE1126", "Other Hispanics or Latinos" = "#F9AA51",
                               "White (non-Hispanic or Latino)" = "#9CA168",
                               "Black (non-Hispanic or Latino)" = "#8CBEC0",
                               "Other (non-Hispanic or Latino)" = "#0B8489")

# Create the bar chart with percentage labels
# Create the bar chart with percentage labels
ggplot(df_long, aes(x = age_group, y = percentage, fill = race_ethnicity)) +
  geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            position = position_dodge(width = 0.9), 
            vjust = -1, 
            size = 1.8, 
            fontface = "bold") +
  labs(title = "Population by Age for Mexicans and Other \nRacial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates) ", 
       caption = "Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year\n Estimates). Tabulations by Great Cities Institute.", fill = ''
) + 
  scale_y_continuous(breaks = seq(0, 40, by = 5)) +
  theme_minimal() +
  scale_fill_manual(values = color_race_ethnicity, name = "") +
  theme_plots +
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme_plots +   theme(axis.text.x = element_text(size = 8),
        panel.background = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
         axis.title.x = element_blank(),
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        legend.key.size = unit(0.5, "lines"),
        legend.position = "bottom",
        legend.text = element_text(size = 7, color = "#636363"), # Increase margins to fit legend
        plot.caption = element_text(size = 8))  + 
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) # Distribute legend items across 2 rows

```





## 1B(2) Population by Age for Hispanic or Latino Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis

# filter for year 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)

# 1B: Population pyramid of the percentage of Mexican population by age group compared to the rest of the population in Chicago and median age in Chicago (Data source: 2018-2022 ACS data)

df <-  data_chi_2018_22 %>%
  mutate(
    age_group = case_when(
      age < 20 ~ "Young (0-19)",
      age >= 20 & age < 40 ~ "Adults (20-39)",
      age >= 40 & age < 60 ~ "Middle-aged (40-59)",
      age >= 60 & age < 80 ~ "Older Adults (60-79)",
      age >= 80 ~ "Elderly (80+)"
    )
  )
# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)


# Calculate weighted population counts by age group and ethnicity

# Calculate weighted population by ethnicity and age group
pop_pyramid <- df %>%
  as_survey_design(weights = perwt) %>% 
  survey_count(hispan_breakdown, age_group, name="total_weighted_count")  %>%
  group_by(hispan_breakdown) %>%
  mutate(total_weighted_percentage = total_weighted_count / sum(total_weighted_count) * 100,
         total_weighted_percentage  = paste0(format(round(total_weighted_percentage, 1), nsmall = 1), "%"))




# get median age

weighted_age  <- df %>%
  group_by(hispan_breakdown) %>%
  summarize(
    weighted_age = matrixStats::weightedMedian(age, perwt)
  ) 



# combined

df_final  <- pop_pyramid %>% left_join(weighted_age) %>%   mutate(across(where(is.numeric), ~replace(., is.na(.), 0))) %>%
  mutate(across(where(~ !is.numeric(.)), ~replace(., is.na(.), "Missing Data"))) %>%
  filter(!is.na(hispan_breakdown))
```

```{r}
# visual

table_1b.2 <-  df_final %>%
  mutate(age_group = factor(age_group, levels = c("Young (0-19)", "Adults (20-39)", "Middle-aged (40-59)", "Older Adults (60-79)", "Elderly (80+)"))) %>%
  arrange(age_group) %>%  # Explicitly arrange by factor levels
  pivot_wider(id_cols = c("age_group"), names_from = hispan_breakdown, values_from = c("total_weighted_count", "total_weighted_percentage")) %>%
  select(age_group,
         total_weighted_count_Mexican, total_weighted_percentage_Mexican,
         'total_weighted_count_Puerto Rican', 'total_weighted_percentage_Puerto Rican',
         total_weighted_count_Ecuadorian, total_weighted_percentage_Ecuadorian,
         total_weighted_count_Cuban, total_weighted_percentage_Cuban,
         total_weighted_count_Guatemalan, total_weighted_percentage_Guatemalan,
         total_weighted_count_Colombian, total_weighted_percentage_Colombian) %>%

  # using kable
  kbl(, booktabs = T,
      col.names = c("", "Number", "Percent" , "Number", "Percent", "Number", "Percent", "Number", "Percent", "Number", "Percent", "Number", "Percent"),
      align = c("l", rep("r", 12)),
      caption = "Population by Age for Hispanic or Latino Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
     format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "12em") %>%
  column_spec(2:13, width = "2.75em") %>%
   row_spec(0, align = "C") %>%  # Center column headers
  add_header_above(align = c("l", rep("c", 6)),
                   c("Age Group", "Mexican" = 2, "Puerto Rican" = 2, "Ecuadorian" = 2, "Cuban" = 2, "Guatemalan" = 2, "Colombian" = 2),
                   bold = T, font_size = 8) %>%
  add_footnote(
  c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)

table_1b.2
```


```{r fig.width=6, fig.height=5 ,fig.align='center'}
# visual

# Pivot data to long format and clean age_group names
df_long <- df_final %>%
  pivot_wider(id_cols = c("hispan_breakdown"), 
              names_from = age_group, 
              values_from = c("total_weighted_count", "total_weighted_percentage")) %>% 
  select(hispan_breakdown, 
         'total_weighted_percentage_Young (0-19)',
         'total_weighted_percentage_Adults (20-39)',
         'total_weighted_percentage_Middle-aged (40-59)',
         'total_weighted_percentage_Older Adults (60-79)',
         'total_weighted_percentage_Elderly (80+)') %>%
  pivot_longer(cols = starts_with("total_weighted_percentage_"), 
               names_to = "age_group", 
               values_to = "percentage") %>%
  mutate(age_group = gsub("total_weighted_percentage_", "", age_group),
         percentage = as.numeric(gsub("%", "", percentage))) # Convert percentage to numeric

# Reverse the order of age groups
df_long <- df_long %>%
  mutate(age_group = factor(age_group, 
                            levels = c("Young (0-19)", "Adults (20-39)", "Middle-aged (40-59)", "Older Adults (60-79)", "Elderly (80+)")))

# Define the custom colors for race_ethnicity
color_hispan_breakdown <- c(
  "Mexican" = "#CE1126", 
  "Puerto Rican" = "#3F88C5",
  "Ecuadorian" = "#FFD700", 
  "Cuban" = "#FF6F61", 
  "Guatemalan" = "#2E8B57", 
  "Colombian" = "#8B4513"
)

# Create the bar chart with percentage labels
ggplot(df_long, aes(x = age_group, y = percentage, fill = hispan_breakdown)) +
  geom_bar(stat = "identity", position = "dodge") +
   geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            position = position_dodge(width = 0.9), 
            vjust = -1, 
            size = 1.8, 
            fontface = "bold") +
  labs(title = "Population by Age for Hispanic or Latino Groups \n  in Chicago, 2018-2022 (ACS 5-year Estimates)", 
      caption = "Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year\n Estimates). Tabulations by Great Cities Institute.", fill = ''
)  +
scale_y_continuous(breaks = seq(0, 40, by = 5)) +
  theme_minimal() +
  scale_fill_manual(values = color_hispan_breakdown) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme_plots +   theme(axis.text.x = element_text(size = 8),
        panel.background = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
         axis.title.x = element_blank(),
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        legend.key.size = unit(0.5, "lines"),
        legend.position = "bottom",
          legend.text = element_text(size = 7, color = "#636363"), # Increase margins to fit legend
        plot.caption = element_text(size = 8))  + 
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) # Distribute legend items across 2 rows
```



## 1B(3): Median Age for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)
```{r}
# analysis

# filter for year 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)

# 1B: Population pyramid of the percentage of Mexican population by age group compared to the rest of the population in Chicago and median age in Chicago (Data source: 2018-2022 ACS data)

df <-  data_chi_2018_22 %>%
  mutate(
    age_group = case_when(
      age < 20 ~ "Young (0-19)",
      age >= 20 & age < 40 ~ "Young Adults (20-39)",
      age >= 40 & age < 60 ~ "Middle-aged (40-59)",
      age >= 60 & age < 80 ~ "Older Adults (60-79)",
      age >= 80 ~ "Elderly (80+)"
    )
  )
# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)


# Calculate weighted population counts by age group and ethnicity

# Calculate weighted population by ethnicity and age group
pop_pyramid <- df %>%
  as_survey_design(weights = perwt) %>% 
  survey_count(race_ethnicity, age_group, name="total_weighted_count")  %>%
  group_by(race_ethnicity) %>%
  mutate(total_weighted_percentage = total_weighted_count / sum(total_weighted_count) * 100,
         total_weighted_percentage  = paste0(format(round(total_weighted_percentage, 1), nsmall = 1), "%"))




# get median age

weighted_age  <- df %>%
  group_by(race_ethnicity) %>%
  summarize(
    weighted_age = matrixStats::weightedMedian(age, perwt)
  )   %>% mutate(
    weighted_age = format(round(weighted_age, 1), nsmall = 1)
  ) 



# combined

df_final  <- pop_pyramid %>% left_join(weighted_age)
```


```{r}
# visuals
# visuals

table_1b.3 <- weighted_age  %>% select(race_ethnicity, weighted_age) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
       col.names = c("Race/Ethnicity", "Median Age"),
      align = c("l",rep("r",1)),
      caption = "Median Age for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2, width = "25.8em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%
add_footnote(
  c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)

table_1b.3


```



## 1B(4): Median Age for Hispanic or Latino Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis

# filter for year 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)

# 1B: Population pyramid of the percentage of Mexican population by age group compared to the rest of Latino Population and median age in Chicago (Data source: 2018-2022 ACS data)

df <-  data_chi_2018_22 %>%
  mutate(
    age_group = case_when(
      age < 20 ~ "Young (0-19)",
      age >= 20 & age < 40 ~ "Young Adults (20-39)",
      age >= 40 & age < 60 ~ "Middle-aged (40-59)",
      age >= 60 & age < 80 ~ "Older Adults (60-79)",
      age >= 80 ~ "Elderly (80+)"
    )
  )
# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)


# Calculate weighted population counts by age group and ethnicity

# Calculate weighted population by ethnicity and age group
pop_pyramid <- df %>%
  as_survey_design(weights = perwt) %>% 
  survey_count(hispan_breakdown, age_group, name="total_weighted_count")  %>%
  group_by(hispan_breakdown) %>%
  mutate(total_weighted_percentage = total_weighted_count / sum(total_weighted_count) * 100,
         total_weighted_percentage  = paste0(format(round(total_weighted_percentage, 1), nsmall = 1), "%"))




# get median age

weighted_age  <- df %>%
  group_by(hispan_breakdown) %>%
  summarize(
    weighted_age = matrixStats::weightedMedian(age, perwt)
  ) %>% mutate(
    weighted_age = format(round(weighted_age, 1), nsmall = 1)
  )  %>% filter(!is.na(hispan_breakdown))


# combined

df_final  <- pop_pyramid %>% left_join(weighted_age) %>% filter(!is.na(hispan_breakdown))
```



```{r}
# visuals
# visuals

table_1b.4 <- weighted_age  %>% select(hispan_breakdown, weighted_age) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
       col.names = c("Race/Ethnicity", "Median Age"),
      align = c("l",rep("r",1)),
      caption = "Median Age for Hispanic or Latino Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2, width = "25.8em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%
# row_spec(0, align = 'c') %>%
 add_footnote(
  c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)
  

table_1b.4


```


## 1C: Mean Household Size for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}

# Step 1: Filter data for the year 2022
data_chi_2018_22 <- usa_data %>% filter(year == 2022)

# Step 2: Calculate household size for each household
household_size <- data_chi_2018_22 %>%
   group_by(serial) %>%
   summarize(household_size = n()) %>%
   ungroup()

# Step 3: Merge household size back to the original dataset
data_with_household_size <- data_chi_2018_22 %>%
   left_join(household_size, by = "serial")

# Step 4: Define the survey design using srvyr with household weights
survey_design <- data_with_household_size %>%
   as_survey_design(weights = hhwt)

# Step 5: Calculate the weighted mean household size by race/ethnicity group
mean_household_size <- survey_design %>%
   group_by(race_ethnicity) %>%
   summarize(
     mean_household_size = survey_mean(household_size, na.rm = TRUE)
   ) %>%
   mutate(
     mean_household_size = format(round(mean_household_size, 1), nsmall = 1)
   )



```

```{r}
# visuals

table_1c <- mean_household_size %>% select(race_ethnicity, mean_household_size) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
       col.names = c("Race/Ethnicity", "Mean Household Size"),
      align = c("l",rep("r",1)),
      caption = "Mean Household Size for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2, width = "25.8em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%
 row_spec(0, align = "C") %>%  # Center column headers
 add_footnote(
  c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)

table_1c


```





## 1D(1): Mean and Median Household Income for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis
# filter for 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)

# 1D Income Levels and poverty rates for Mexicans, other Latinos, Black and White Populations in Chicago.

# Filter the data to remove negative and extreme values
filtered_data <- data_chi_2018_22 %>%
  filter(hhincome >= 0 & hhincome < 9999999, pernum == 1)
  
# Define the survey design using srvyr with only weights
survey_design <- filtered_data  %>%
  as_survey_design(weights = hhwt)

# Calculate the weighted Mean household income by group
weighted_avg_income <- survey_design %>% filter(pernum == 1) %>%
  group_by(race_ethnicity) %>%
  summarize(
    avg_hhincome = survey_mean(hhincome, vartype = "se", na.rm = TRUE)
  )   %>% mutate(avg_hhincome = round(avg_hhincome))



# weighted median: 

weighted_median_income <- data_chi_2018_22 %>%  filter(pernum == 1) %>%
  group_by(race_ethnicity) %>%
  summarize(
    weighted_median_hhincome = matrixStats::weightedMedian(hhincome, hhwt)
  ) %>%  
  ungroup() %>% mutate(weighted_median_hhincome = round(weighted_median_hhincome))



# final df with income levels
income_levels <- weighted_avg_income %>% left_join(weighted_median_income)  %>%
  mutate(
    avg_hhincome = paste0("$", format(avg_hhincome, big.mark = ",", scientific = FALSE)),
    weighted_median_hhincome = paste0("$", format(weighted_median_hhincome, big.mark = ",", scientific = FALSE))
  )


```


```{r}
# visuals



table_1d.1 <- income_levels %>% select(race_ethnicity, avg_hhincome, weighted_median_hhincome) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Mean Household Income", "Median Household Income"),
      align = c("l",rep("r",2)),
      caption = "Mean and Median Household Income for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

 add_footnote(
  c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)

table_1d.1
```


## 1D(2.1): Mean and Median Income of Full-time Workers for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)
```{r}
# analysis
# filter for 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)


# 1D Income Levels and poverty rates for Mexicans, other Latinos, Black and White Populations in Chicago.

# Filter the data to remove negative and extreme values
filtered_data <- data_chi_2018_22 %>%
  filter(incwage  > 0 & incwage  < 999999, empstat == 1, uhrswork >= 35, wkswork1 >= 48)



# Define the survey design using srvyr with only weights
survey_design <- filtered_data  %>%
  as_survey_design(weights = perwt)

# Calculate the weighted Mean household income by group
weighted_avg_income <- survey_design %>%
  group_by(race_ethnicity) %>%
  summarize(
    avg_incwage  = survey_mean(incwage , vartype = "se", na.rm = TRUE)
  ) %>% mutate(avg_incwage = round(avg_incwage))




# weighted median: 

weighted_median_income <- filtered_data %>% 
  group_by(race_ethnicity) %>%
  summarize(
    weighted_median_incwage  = matrixStats::weightedMedian(incwage , perwt)
  ) %>%
  ungroup() %>% mutate(weighted_median_incwage = round(weighted_median_incwage))



# final df with income levels
income_levels <- weighted_avg_income %>% left_join(weighted_median_income)  %>%
  mutate(
    avg_incwage  = paste0("$", format(avg_incwage , big.mark = ",", scientific = FALSE)),
    weighted_median_incwage  = paste0("$", format(weighted_median_incwage , big.mark = ",", scientific = FALSE))
  )

```


```{r}
# visuals



table_1d.2.1 <- income_levels %>% select(race_ethnicity, avg_incwage , weighted_median_incwage) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Mean Income", "Median Income"),
      align = c("l",rep("r",2)),
      caption = "Mean and Median Income of Full-time Workers for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%
  
   add_footnote(
  c(
    "\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
    "\\footnotesize{Note:Data covers full-time employees working a minimum of 35 hours per week for at least 48 weeks each year.}"
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)



table_1d.2.1
```


## 1D(2.2): Mean and Median Income for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis
# filter for 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)


# 1D Income Levels and poverty rates for Mexicans, other Latinos, Black and White Populations in Chicago.

# Filter the data to remove negative and extreme values
filtered_data <- data_chi_2018_22 %>%
  filter(incwage  > 0 & incwage  < 999999,  empstat == 1)



# Define the survey design using srvyr with only weights
survey_design <- filtered_data  %>%
  as_survey_design(weights = perwt)

# Calculate the weighted Mean household income by group
weighted_avg_income <- survey_design %>%
  group_by(race_ethnicity) %>%
  summarize(
    avg_incwage  = survey_mean(incwage , vartype = "se", na.rm = TRUE)
  ) %>% mutate(avg_incwage = round(avg_incwage))


# weighted median: 

weighted_median_income <- filtered_data %>% 
  group_by(race_ethnicity) %>%
  summarize(
    weighted_median_incwage  = matrixStats::weightedMedian(incwage , perwt)
  ) %>%
  ungroup() %>% mutate(weighted_median_incwage = round(weighted_median_incwage))


# final df with income levels
income_levels <- weighted_avg_income %>% left_join(weighted_median_income)  %>%
  mutate(
    avg_incwage  = paste0("$", format(avg_incwage , big.mark = ",", scientific = FALSE)),
    weighted_median_incwage  = paste0("$", format(weighted_median_incwage , big.mark = ",", scientific = FALSE))
  )

```


```{r}
# visuals



table_1d.2.2 <- income_levels %>% select(race_ethnicity, avg_incwage , weighted_median_incwage) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Mean Income", "Median Income"),
      align = c("l",rep("r",2)),
      caption = "Mean and Median Income for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%
  
     add_footnote(
  c(
    "\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
    "\\footnotesize{Note: The data includes all employed individuals, regardless of whether they work full-time or part-time.}"
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)

table_1d.2.2
```



## 1D(3.1): Mean and Median Income of Full-time Workers for Hispanic or Latino Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis
# filter for 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)

# 1D Income Levels and poverty rates for Mexicans, other Latinos, Black and White Populations in Chicago.

# Filter the data to remove negative and extreme values
filtered_data <- data_chi_2018_22 %>%
  filter(incwage  > 0 & incwage  < 999999, empstat == 1, uhrswork >= 35, wkswork1 >= 48)

  
# Define the survey design using srvyr with only weights
survey_design <- filtered_data  %>%
  as_survey_design(weights = perwt)

# Calculate the weighted Mean household income by group
weighted_avg_income <- survey_design %>% 
  group_by(hispan_breakdown) %>%
  summarize(
    avg_incwage  = survey_mean(incwage , vartype = "se", na.rm = TRUE)
  ) %>% mutate(avg_incwage = round(avg_incwage))


# weighted median: 

weighted_median_income <- filtered_data %>% 
  group_by(hispan_breakdown) %>%
  summarize(
    weighted_median_incwage  = matrixStats::weightedMedian(incwage , perwt)
  ) %>%
  ungroup() %>% mutate(weighted_median_incwage = round(weighted_median_incwage))


# final df with income levels
income_levels <- weighted_avg_income %>% left_join(weighted_median_income)  %>%
  mutate(
    avg_incwage  = paste0("$", format(avg_incwage , big.mark = ",", scientific = FALSE)),
    weighted_median_incwage  = paste0("$", format(weighted_median_incwage , big.mark = ",", scientific = FALSE))
  ) %>% filter(!is.na( hispan_breakdown))


```


```{r}
# visuals


table_1d.3.1 <- income_levels %>% select(hispan_breakdown, avg_incwage , weighted_median_incwage ) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Mean Income", "Median Income"),
      align = c("l",rep("r",2)),
      caption = "Mean and Median Income of Full-time Workers for Hispanic or Latino Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

   add_footnote(
  c(
    "\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
    "\\footnotesize{Note: Data covers full-time employees working a minimum of 35 hours per week for at least 48 weeks each year.}"
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)


table_1d.3.1
```


## 1D(3.2): Mean and Median Income for Hispanic or Latino Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis
# filter for 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)

# 1D Income Levels and poverty rates for Mexicans, other Latinos, Black and White Populations in Chicago.

# Filter the data to remove negative and extreme values
filtered_data <- data_chi_2018_22 %>%
  filter(incwage  > 0 & incwage  < 999999, empstat == 1)

  
# Define the survey design using srvyr with only weights
survey_design <- filtered_data  %>%
  as_survey_design(weights = perwt)

# Calculate the weighted Mean household income by group
weighted_avg_income <- survey_design %>% 
  group_by(hispan_breakdown) %>%
  summarize(
    avg_incwage  = survey_mean(incwage , vartype = "se", na.rm = TRUE)
  ) %>% mutate(avg_incwage = round(avg_incwage))


# weighted median: 

weighted_median_income <- filtered_data %>% 
  group_by(hispan_breakdown) %>%
  summarize(
    weighted_median_incwage  = matrixStats::weightedMedian(incwage , perwt)
  ) %>%
  ungroup() %>% mutate(weighted_median_incwage = round(weighted_median_incwage))


# final df with income levels
income_levels <- weighted_avg_income %>% left_join(weighted_median_income)  %>%
  mutate(
    avg_incwage  = paste0("$", format(avg_incwage , big.mark = ",", scientific = FALSE)),
    weighted_median_incwage  = paste0("$", format(weighted_median_incwage , big.mark = ",", scientific = FALSE))
  ) %>% filter(!is.na( hispan_breakdown))


```


```{r}
# visuals


table_1d.3.2 <- income_levels %>% select(hispan_breakdown, avg_incwage , weighted_median_incwage) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Mean Income", "Median Income"),
      align = c("l",rep("r",2)),
      caption = "Mean and Median Income for Hispanic or Latino Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%
  
     add_footnote(
  c(
    "\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
    "\\footnotesize{Note: The data includes all employed individuals, regardless of whether they work full-time or part-time.}"
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)

      

table_1d.3.2
```


## 1D(4.1): Mean and Median Income of Full-time Workers for Mexico Born and U.S. Born Mexicans in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis
# filter for 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022) %>% filter(!is.na(mex_foreign_born))

# 1D Income Levels and poverty rates for Mexicans born in Mexico vs. United States
# Filter the data to remove negative and extreme values
filtered_data <- data_chi_2018_22 %>%
  filter(incwage  > 0 & incwage  < 999999, empstat == 1, uhrswork >= 35, wkswork1 >= 48)

  
# Define the survey design using srvyr with only weights
survey_design <- filtered_data  %>%
  as_survey_design(weights = perwt)

# Calculate the weighted Mean household income by group
weighted_avg_income <- survey_design %>%
  group_by(mex_foreign_born) %>%
  summarize(
    avg_incwage  = survey_mean(incwage , vartype = "se", na.rm = TRUE)
  ) %>% mutate(avg_incwage = round(avg_incwage))


# weighted median: 

weighted_median_income <- filtered_data %>% 
  group_by(mex_foreign_born) %>%
  summarize(
    weighted_median_incwage  = matrixStats::weightedMedian(incwage ,perwt)
  ) %>%
  ungroup() %>% mutate(weighted_median_incwage = round(weighted_median_incwage))


# final df with income levels
income_levels <- weighted_avg_income %>% left_join(weighted_median_income)  %>%
  mutate(
    avg_incwage  = paste0("$", format(avg_incwage , big.mark = ",", scientific = FALSE)),
    weighted_median_incwage  = paste0("$", format(weighted_median_incwage , big.mark = ",", scientific = FALSE))
  ) 

```

```{r}

# visuals



table_1d.4.1 <- income_levels %>% select(mex_foreign_born, avg_incwage , weighted_median_incwage ) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Mexican Nativity", "Mean Income", "Median Income"),
      align = c("l",rep("r",2)),
      caption = "Mean and Median Income of Full-time Workers for Mexico Born and U.S. Born Mexicans in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

       add_footnote(
  c(
    "\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
    "\\footnotesize{Note: Data covers full-time employees working a minimum of 35 hours per week for at least 48 weeks each year.}"
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)

           
table_1d.4.1
```


## 1D(4.2): Mean and Median Income for Mexico Born and U.S. Born Mexicans in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis
# filter for 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022) %>% filter(!is.na(mex_foreign_born))

# 1D Income Levels and poverty rates for Mexicans born in Mexico vs. United States
# Filter the data to remove negative and extreme values
filtered_data <- data_chi_2018_22 %>%
  filter(incwage  > 0 & incwage  < 999999, empstat == 1)

  
# Define the survey design using srvyr with only weights
survey_design <- filtered_data  %>%
  as_survey_design(weights = perwt)

# Calculate the weighted Mean household income by group
weighted_avg_income <- survey_design %>%
  group_by(mex_foreign_born) %>%
  summarize(
    avg_incwage  = survey_mean(incwage , vartype = "se", na.rm = TRUE)
  ) %>% mutate(avg_incwage = round(avg_incwage))


# weighted median: 

weighted_median_income <- filtered_data %>% 
  group_by(mex_foreign_born) %>%
  summarize(
    weighted_median_incwage  = matrixStats::weightedMedian(incwage ,perwt)
  ) %>%
  ungroup() %>% mutate(weighted_median_incwage = round(weighted_median_incwage))


# final df with income levels
income_levels <- weighted_avg_income %>% left_join(weighted_median_income)  %>%
  mutate(
    avg_incwage  = paste0("$", format(avg_incwage , big.mark = ",", scientific = FALSE)),
    weighted_median_incwage  = paste0("$", format(weighted_median_incwage , big.mark = ",", scientific = FALSE))
  ) 

```

```{r}

# visuals



table_1d.4.2 <- income_levels %>% select(mex_foreign_born, avg_incwage , weighted_median_incwage ) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Mexican Nativity", "Mean Income", "Median Income"),
      align = c("l",rep("r",2)),
      caption = "Mean and Median Income for Mexico Born and U.S. Born Mexicans in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%
  
         add_footnote(
  c(
    "\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
    "\\footnotesize{Note: Data covers full-time employees working a minimum of 35 hours per week for at least 48 weeks each year.}"
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)


table_1d.4.2
```





## 1D(5): Population in Poverty and Poverty Rate for Mexicans and Other Racial/Ethnic Groups in Chicago, 2000 (Decennial Census), 2008-2012, and 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis

# poverty
# filter for 2022

data_chi_2018_22 <- usa_data 

# Define the survey design using srvyr with only weights
survey_design2 <- data_chi_2018_22  %>%
  as_survey_design(weights = perwt)
  

  
# Calculate poverty rates by group

# Calculate total weighted count by race_category and ethnicity
total_weighted_pop_count <- survey_design2 %>%
  group_by(race_ethnicity, year) %>%
  summarise(total_weighted_pop_count = survey_total())

# Calculate total weighted count in poverty by race_category and ethnicity
total_weighted_poverty_count <- survey_design2 %>%
  filter(poverty <= 100) %>%
  group_by(race_ethnicity, year) %>%
  summarise(total_weighted_poverty_count = survey_total())

# Combine total counts and poverty counts
combined <- left_join(total_weighted_pop_count, total_weighted_poverty_count)


poverty_rate_trends <- combined %>% mutate(poverty_rate = (total_weighted_poverty_count/total_weighted_pop_count)*100)

poverty_rate_trends_wide <- poverty_rate_trends %>%  pivot_wider(
  id_cols = race_ethnicity,
    names_from = year,
    values_from = c(
      total_weighted_poverty_count,
      poverty_rate
    ),
    names_sep = "_"
  ) %>% select(race_ethnicity, total_weighted_poverty_count_2000, poverty_rate_2000,  total_weighted_poverty_count_2012, poverty_rate_2012,  total_weighted_poverty_count_2022, poverty_rate_2022) %>% mutate(diff_2022_2000 = poverty_rate_2022-poverty_rate_2000) %>%  mutate(across(starts_with("poverty_rate") | starts_with("diff"), ~ paste0(format(round(.x, 1), nsmall = 1), "%")))


```


```{r}
# visuals


table_1d.5 <- poverty_rate_trends_wide  %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("", "Number", "Percent", "Number", "Percent", "Number", "Percent", "Difference"),
      align = c("l",rep("r",7)),
      caption = "Population in Poverty and Poverty Rate for Mexicans and Other Racial/Ethnic Groups in Chicago, 2000 (Decennial Census), 2008-2012, and 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:8, width = "4.4em") %>%  
 add_header_above(
    align = c("l",  rep("c", 3)),
    c(
      "Race/Ethnicity", 
      "Poverty in 2000 " = 2, 
      "Poverty in 2008-2012 " = 2, 
      "Poverty in 2018-2022" = 2,
     "Percentage Point Difference\n from 2000 to 2018-2022" = 1),
    bold = TRUE, 
    font_size = 8) %>%
  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8, align = "C") %>%
  
   add_footnote(
  c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2000 Decennial Census, 2018-2012, and 2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)



table_1d.5
```

## 1E: Total, Hispanic or Latino, and Mexican Population by 10 Community Areas with the Largest Share of Mexican Population, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis
top_10 <- chicago_ca_pop %>% as.data.frame() %>% select(commnty, totl_pp, ttl_hs_,ttl_mxc, prc_mx_, prc_mxc) %>%   mutate(commnty = str_to_title(commnty)) %>% arrange(desc(prc_mxc)) %>%  slice_max(order_by = prc_mxc, n = 10)


# top 10 row
totals <- top_10 %>%
  summarise(
    commnty = "10 Community Areas with the Largest Share of Mexican Population",
    totl_pp = sum(totl_pp, na.rm = TRUE),
    ttl_hs_ = sum(ttl_hs_, na.rm = TRUE),
    ttl_mxc = sum(ttl_mxc, na.rm = TRUE),
    prc_mx_ = (ttl_mxc/ttl_hs_) * 100,
    prc_mxc = (ttl_mxc/totl_pp) * 100
  )



# rest of community areas
min_67 <- chicago_ca_pop %>% as.data.frame() %>% select(commnty, totl_pp, ttl_hs_,ttl_mxc, prc_mx_, prc_mxc) %>%   mutate(commnty = str_to_title(commnty)) %>% slice_min(order_by = prc_mxc, n = 67)


# the rest
totals_min <- min_67 %>%
  summarise(
    commnty = "Rest of Community Areas",
    totl_pp = sum(totl_pp, na.rm = TRUE),
    ttl_hs_ = sum(ttl_hs_, na.rm = TRUE),
    ttl_mxc = sum(ttl_mxc, na.rm = TRUE),
    prc_mx_ = (ttl_mxc/ttl_hs_) * 100,
    prc_mxc = (ttl_mxc/totl_pp) * 100
  )



# all of chicago
df_all <- chicago_ca_pop %>% as.data.frame() %>% select(commnty, totl_pp, ttl_hs_,ttl_mxc, prc_mx_, prc_mxc) %>%   mutate(commnty = str_to_title(commnty)) %>% 
  summarise(
    commnty = "Chicago",
    totl_pp = sum(totl_pp, na.rm = TRUE),
    ttl_hs_ = sum(ttl_hs_, na.rm = TRUE),
    ttl_mxc = sum(ttl_mxc, na.rm = TRUE),
    prc_mx_ = (ttl_mxc/ttl_hs_) * 100,
    prc_mxc = (ttl_mxc/totl_pp) * 100
  )





df_final <- rbind(top_10, totals, totals_min, df_all) %>% mutate(
    prc_mx_ = paste0(format(round(prc_mx_, 1), nsmall = 1), "%"),
    prc_mxc = paste0(format(round(prc_mxc, 1), nsmall = 1), "%")
)


```


```{r}
# visuals


table_1e <- df_final   %>% select(commnty, totl_pp, ttl_hs_, ttl_mxc, prc_mx_ , prc_mxc) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("", "", "", "", "", ""),
      align = c("l",rep("c",5)),
      caption = "Total, Hispanic or Latino, and Mexican Population by 10 Community Areas with the Largest Share of Mexican\n Population, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:6, width = "6.16em") %>%  
  row_spec(0, align = "C") %>%  # Center column headers
  add_header_above(
    align = c("l", rep("c", 5)),
    c(
      "10 Community Areas\nwith the Largest Share\n of Mexican Population", 
      "Total\n Population" = 1, 
      "Hispanic or\n Latino Population " = 1, 
      "Mexican\n Population" = 1,
     "% Mexicans of \n Hispanic or Latino" = 1,
     "% Mexicans\n of Total Pop" = 1),
    bold = TRUE, 
    font_size = 8) %>% 
 row_spec(c(11), hline_after = T) %>%  
  row_spec(12:13, bold = T,
             background = "#f2f0f7",
            font_size = 8) %>%

     add_footnote(
  c("\\footnotesize{Data Source: IPUMS NHGIS, University of Minnesota, www.nhgis.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)


table_1e
```



# 2. Education

## 2.A: Number and Percent in Public and Private Schools for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

\blandscape


```{r}
# analysis
# select variables
data_chi_2018_22 <- usa_data  %>% filter(year == 2022) %>% select(year:gqtyped, perwt, hispan, race, race_ethnicity, school, schltype)



# Number and Percentage of Mexican population in public and private schools compared to other groups

# Step 1: Filter for individuals enrolled in school
enrolled_in_school <- data_chi_2018_22 %>%
  filter(school == 2)


# Define the survey design
survey_design <- enrolled_in_school %>% 
  as_survey_design(weights = perwt)

# Step 3: Calculate the total weighted count for each race/ethnicity group (denominator)
total_counts <- survey_design %>%
  group_by(race_ethnicity) %>%
  summarize(total_weighted_count = survey_total(vartype = "se"))

# Step 4: Calculate the weighted count of race/ethnicity in public and private schools (numerator)
school_counts <- survey_design %>%
  group_by(schltype, race_ethnicity) %>%
  summarize(weighted_count = survey_total(vartype = "se"))
  
# Step 5: Calculate the percentage of race/ethnicity groups in public and private schools
school_percentages <- school_counts %>%
  left_join(total_counts, by = "race_ethnicity") %>%
    mutate(percentage = paste0(format(round((weighted_count / total_weighted_count) * 100, 1), nsmall = 1), "%"))


# Step 6: Recode schltype for clarity
df_all <- school_percentages %>%
  mutate(schltype = case_when(
    schltype == 2 ~ "Public",
    schltype == 3 ~ "Private",
    TRUE ~ NA_character_
  ))
```


```{r}
# visual


table_2a <- df_all %>% 
  
pivot_wider(id_cols = schltype, names_from = "race_ethnicity",  values_from = c(weighted_count, total_weighted_count, percentage)) %>% 
  select(schltype,   weighted_count_Mexican,  percentage_Mexican, 
    'weighted_count_Other Hispanics or Latinos', 'percentage_Other Hispanics or Latinos',
    'weighted_count_White (non-Hispanic or Latino)',  'percentage_White (non-Hispanic or Latino)',
   'weighted_count_Black (non-Hispanic or Latino)',   'percentage_Black (non-Hispanic or Latino)',
    'weighted_count_Other (non-Hispanic or Latino)', 'percentage_Other (non-Hispanic or Latino)') %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("", "Number", "Percent" , "Number", "Percent", "Number", "Percent", "Number", "Percent", "Number", "Percent"),
      align = c("l",rep("r",10)),
      caption = "Number and Percent in Public and Private Schools for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:11, width = "3.08em")   %>%
   row_spec(0, align = "C") %>%  # Center column headers
 add_header_above(
  align = c("l", rep("c", 5)),  # Assuming 1 for Age Group and 10 divided into 5 groups
  c("Age Group", "Mexican" = 2, "Other Hispanics\n or Latinos" = 2, "White\n (non-Hispanic\n or Latino)" = 2, "Black\n (non-Hispanic\n or Latino)" = 2, "Other\n (non-Hispanic\n or Latino)" = 2),
  bold = TRUE, font_size = 8
) %>% 

   add_footnote(
  c(
    "\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)

table_2a 
```

\elandscape



## 2.B: Performance in Chicago Public Schools (CPS) Elementary Schools in Four Community Areas with Largest Share of Mexican Population and Performance Citywide by Race/Ethnicity, 2023

```{r}
# analysis
df_top_mex_schools <- chicago_ca_pop %>% arrange(desc(prc_mxc)) %>% top_n(4, prc_mxc) %>% left_join(school_indicator, by = c("commnty" = "commnty")) %>% select(school_name, chronic_absenteeism_hispanic_or_latino, ends_with("proficiency_hispanic_or_latino"))


# Calculate the mean for the specified columns, ignoring NA values
Mean_values <- colMeans(df_top_mex_schools[, c("chronic_absenteeism_hispanic_or_latino",
                                                  "percent_ela_proficiency_hispanic_or_latino",
                                                  "percent_math_proficiency_hispanic_or_latino",
                                                  "percent_science_proficiency_hispanic_or_latino")], na.rm = TRUE)

# Round the Mean to the nearest decimal digit
Mean_values_rounded <- round(Mean_values, 1)

# Create a new row with these Mean values
# Adding % sign to the appropriate columns
# Create a new row with these rounded Mean values and add the % sign
new_row <- c(
    "Mean Metrics for Hispanic or Latino Students in Four Community Areas with Largest Share of Mexican Population", 
    paste0(format(Mean_values_rounded[1], nsmall = 1), "%"), 
    paste0(format(Mean_values_rounded[2], nsmall = 1), "%"), 
    paste0(format(Mean_values_rounded[3], nsmall = 1), "%"), 
    paste0(format(Mean_values_rounded[4], nsmall = 1), "%"), 
    NA
)

# Set the column names of the new row to match the original data frame
names(new_row) <- colnames(df_top_mex_schools)

# Convert the new row to a data frame
df_Mean <- as.data.frame(t(new_row), stringsAsFactors = FALSE)

# Rename columns by removing "_hispanic_or_latino" suffix
names(df_Mean) <- gsub("_hispanic_or_latino", "", names(df_Mean))

# Mean schools 
df_Mean_mex <- df_Mean %>% select(school_name, 
chronic_absenteeism,
percent_ela_proficiency,
percent_math_proficiency,
percent_science_proficiency)
# Data Source: Hispanic/Latino District Total, 
#https://www.illinoisreportcard.com/District.aspx?source=trends&source2=proficiency&Districtid=15016299025 #https://www.illinoisreportcard.com/District.aspx?source=studentcharacteristics&source2=chronicabsenteeism&Districtid=15016299025

# Create a data frame
df_latino_state <- data.frame(
  school_name = "Total CPS Hispanic or Latino Students",
  chronic_absenteeism = 40.3,
  percent_ela_proficiency = 21.2,
  percent_math_proficiency = 13.6,
  percent_science_proficiency = 35.6
)

df_white_state <- data.frame(
  school_name = "Total CPS White Students",
  chronic_absenteeism = 27.1,
  percent_ela_proficiency = 54.3,
  percent_math_proficiency = 48.4,
  percent_science_proficiency = 65.0
)

df_black_state <- data.frame(
  school_name = "Total CPS Black Students",
  chronic_absenteeism = 45.8,
  percent_ela_proficiency = 16.5,
  percent_math_proficiency = 8.1,
  percent_science_proficiency = 24.4
)
# for Asian, AIAN, NHPI, Two or more
df_other_state <- data.frame(
  school_name = "Total CPS Other (non-Hispanic or Latino) Students",
  chronic_absenteeism = as.numeric(format((21.2 + 41 + 35.4 + 30.1) / 4, nsmall = 1)),
  percent_ela_proficiency = as.numeric(format((54.7 + 31.2 + 51.1 + 53.9) / 4, nsmall = 1)),
  percent_math_proficiency = as.numeric(format((53.5 + 23.9 + 40.7 + 45.7) / 4, nsmall = 1)),
  percent_science_proficiency = as.numeric(format((68.8 + 37.6 + 60 + 61.7) / 4, nsmall = 1))
)


df_state_avgs <- rbind(df_latino_state, df_white_state, df_black_state, df_other_state)  %>%
 mutate(
    chronic_absenteeism = paste0(format(round(chronic_absenteeism, 1), nsmall = 1), "%"),
    percent_ela_proficiency = paste0(format(round(percent_ela_proficiency, 1), nsmall = 1), "%"),
    percent_math_proficiency = paste0(format(round(percent_math_proficiency, 1), nsmall = 1), "%"),
    percent_science_proficiency = paste0(format(round(percent_science_proficiency, 1), nsmall = 1), "%")
)


df_final <- rbind(df_Mean_mex, df_state_avgs)
```


```{r}
# visuals
table_2b <-df_final  %>% select(school_name, percent_ela_proficiency, percent_math_proficiency, percent_science_proficiency) %>% 


 
   # using kable 
  
  kbl(
  , booktabs = TRUE,
  col.names = c("", "Percent", "Percent", "Percent"),
  align = c("l", rep("c", 3)),
  caption = "Performance in Chicago Public Schools (CPS) Elementary Schools in Four Community Areas with Largest Share of Mexican Population and Performance Citywide by Race/Ethnicity, 2023",
  format.args = list(big.mark = ","),
  linesep = ""
) %>%
  kable_classic(
    full_width = FALSE,
    font_size = 8,
    html_font = "Arial",
    position = "center",
    latex_options = "HOLD_position"
  ) %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "10.26em") %>%
   row_spec(0, align = "C") %>%  # Center column headers
  add_header_above(
    align = c("l", rep("c", 3)),
    c(
      "Aggregation", 
      "Mean English \nLanguage Arts Proficiency" = 1, 
      "Mean Math Proficiency" = 1, 
      "Mean Science Proficiency" = 1),
    bold = TRUE, 
    font_size = 8
  ) %>%
  
    add_footnote(
  c(
    "\\footnotesize{Data Source: Illinois Report Card (2023). Tabulations by Great Cities Institute.}",
    "\\footnotesize{Sample size: 31 Chicago Public Schools Elementary Schools in Gage Park, East Side, West Lawn, and South Lawndale.}"
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)

table_2b

```




## 2.C: Chronic Absenteeism in Chicago Public Schools (CPS) Elementary Schools in Four Community Areas with Largest Share of Mexican Population, 2023
```{r}
# visuals
# visuals
table_2c <-df_final  %>% select(school_name, chronic_absenteeism) %>% 


 
   # using kable 
  
  kbl(
  , booktabs = TRUE,
  col.names = c("", "Percent"),
  align = c("l", rep("c", 1)),
  caption = "Chronic Absenteeism in Chicago Public Schools (CPS) Elementary Schools in Four Community Areas with Largest Share of Mexican Population, 2023",
  format.args = list(big.mark = ","),
  linesep = ""
) %>%
  kable_classic(
    full_width = FALSE,
    font_size = 8,
    html_font = "Arial",
    position = "center",
    latex_options = "HOLD_position"
  ) %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2, width = "30.8em") %>%
  add_header_above(
    align = c("l", rep("c", 1)),
    c(
      "Aggregation", 
      "Mean Chronic Absenteeism" = 1),
    bold = TRUE, 
    font_size = 8
  ) %>%
 add_footnote(
  c(
    "\\footnotesize{Data Source: Illinois Report Card (2023). Tabulations by Great Cities Institute.}",
    "\\footnotesize{Sample size: 31 Chicago Public Schools Elementary Schools in Gage Park, East Side, West Lawn, and South Lawndale.}",
      "\\footnotesize{Note: Chronic Absenteeism is defined as students who miss 10 percent or more of school days per year.}"
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)




table_2c

```




## 2.D: Number and Percent of Population with Limited English Proficiency for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)


```{r}

# analysis

data_chi_2018_22 <- usa_data %>%  filter(year == 2022) %>% select(year:gqtyped, perwt, hispan, race, race_ethnicity, school, speakeng)

# Percentage of Limited English Proficient. 

# Step 1: Define Survey Design
survey_design <- data_chi_2018_22 %>%
  as_survey_design(weights = perwt)

# Step 2: Group by race_ethnicity and calculate weighted counts and totals
lep_rate_by_ethnicity <- survey_design %>%
  group_by(race_ethnicity) %>%
  summarise(
    lep_count = survey_total(speakeng %in% c(1, 6)),
    total_count = survey_total(),
    lep_percentage = 100 * survey_mean(speakeng %in% c(1, 6), vartype = "se")
  ) %>%
  mutate(lep_percentage = paste0(format(round(lep_percentage, 1), nsmall = 1), "%"))



```


```{r}
# visual



table_2d <- lep_rate_by_ethnicity %>% select(race_ethnicity, lep_count, lep_percentage) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number", "Percent"),
      align = c("l",rep("r",2)),
      caption = "Number and Percent of Population with Limited English Proficiency for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "15.4em") %>%  
  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%
  
  add_footnote(
  c(
    "\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
      "\\footnotesize{Note: English Proficiency is defined as respondents who does not speak English or speaks English poorly.}"
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)

  


table_2d

```



## 2.E: Number and Percent of Non-citizens enrolled in Public Schools for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis


# Step 1: Filter for individuals enrolled in school
enrolled_in_school <- usa_data  %>%  filter(year == 2022) %>%
  filter(school == 2)


# Step 2: Filter for non-citizen individuals enrolled in public schools
enrolled_in_public_schools <- enrolled_in_school %>%
  filter(schltype == 2 & citizen == 3)
  
# Step 3: Create survey design
  
survey_design_school <- enrolled_in_school %>%
  as_survey_design(weights = perwt)
  
survey_design_public_schools <- enrolled_in_public_schools %>%
  as_survey_design(weights = perwt)


# Step 4: Calculate denominator (total count enrolled in school) for each subgroup
denominator_counts <- survey_design_school %>%
  group_by(race_ethnicity) %>%
  summarise(num_enrolled_all_school = survey_total()) 
  
  
  
# Step 5: Calculate numerator (count of public school enrollment) for each subgroup
numerator_counts <- survey_design_public_schools %>%
  group_by(race_ethnicity) %>%
  summarise(num_enrolled_public = survey_total())
  
  
# Step 6: Calculate percentages and estimate SEs

percentages <- numerator_counts %>%
  left_join(denominator_counts, by = "race_ethnicity") %>%
  mutate(percentage = 100 * num_enrolled_public / num_enrolled_all_school) %>%
  mutate(percentage = paste0(format(round(percentage, 1), nsmall = 1), "%"))


  


```


```{r}
# visual

table_2e <- percentages %>% select(race_ethnicity, num_enrolled_public, percentage) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number", "Percent"),
      align = c("l",rep("r",2)),
      caption = "Number and Percent of Non-citizens enrolled in Public Schools for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "15.4em") %>%  
row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%
 add_footnote(
    c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
    notation = "none",
    threeparttable = TRUE,
    escape = FALSE
  )
table_2e

```


## 2.F: Number and Percent of Population Aged 21-24 Without A High School Diploma for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}


df <- usa_data  %>%  filter(year == 2022) %>% select(race_ethnicity, schltype, school, gradeatt, perwt, educ, age) 


# Create a survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)


# method: Look at 21-24 year olds without a high school diploma out of all 21-24 year olds


# Calculate numerator: weighted count of students aged 21-24 not in school and without a high school diploma
numerator_not_in_school <- survey_design %>%
  filter(age >= 21, age <= 24, school == 1, educ %in% c("0", "1", "2", "3", "4", "5")) %>%
  group_by(race_ethnicity) %>%
  summarise(not_in_school = survey_total(vartype = "se"))

# Calculate denominator: weighted count of total students aged 21-24
denominator_total_students <- survey_design %>%
  filter(age >= 21, age <= 24) %>%
  group_by(race_ethnicity) %>%
  summarise(total_students = survey_total(vartype = "se"))

  
# Calculate dropout rate by joining numerator and denominator
dropout_data <- numerator_not_in_school %>%
  inner_join(denominator_total_students, by = "race_ethnicity") %>%
  mutate(dropout_rate = (not_in_school / total_students) * 100) %>%
  mutate(dropout_rate = paste0(format(round(dropout_rate, 1), nsmall = 1), "%"))


```


```{r}
# visual


table_2f <- dropout_data %>% select(race_ethnicity, not_in_school, dropout_rate) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number", "Percent"),
      align = c("l",rep("r",2)),
      caption = "Number and Percent of Population Aged 21-24 Without A High School Diploma for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "15.4em") %>%  
  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

 add_footnote(
    c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
    notation = "none",
    threeparttable = TRUE,
    escape = FALSE
  )

table_2f

```







## 2.G: Number and Percent of Population Attending Undergraduate or Graduate Colleges or Professional Schools for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}

## select variables
df <- usa_data  %>%  filter(year == 2022) %>% select(race_ethnicity, school, gradeatt, perwt, educ, age)


# Create survey design object
survey_design <-df %>%
  as_survey_design(weights = perwt)

# Number and percentage of Mexican students attending area colleges

# 18-24 year olds


# Group by race_ethnicity, filter for enrolled in college/grad school and ages 18-24
numerator_18_24 <- survey_design %>%  filter(school == 2, gradeatt %in% c(6, 7), age >= 18, age <= 24) %>% 
  group_by(race_ethnicity) %>%
  summarise(college_attendees = survey_total(vartype = "se"))


# denominator calculate total number of 18-24 year olds
denominator_18_24 <- survey_design %>% filter(age >= 18, age <= 24) %>%
  group_by(race_ethnicity) %>%
  summarise(total_students = survey_total(vartype = "se"))


# Calculate percentage of students attending college within each racial and ethnic group
college_attendance_18_24 <- left_join(numerator_18_24, denominator_18_24, by = "race_ethnicity") %>%
  mutate(percentage_college_students = (college_attendees / total_students) * 100) %>% mutate(age = "18_24", percentage_college_students = paste0(format(round(percentage_college_students, 1), nsmall = 1), "%"))




# 25-34 year olds

# Group by race_ethnicity, filter for enrolled in college/grad school and ages 18-24
numerator_25_34 <- survey_design %>%  filter(school == 2, gradeatt %in% c(6, 7), age >= 25, age <= 34) %>% 
  group_by(race_ethnicity) %>%
  summarise(college_attendees = survey_total(vartype = "se"))


# denominator calculate total number of 18-24 year olds
denominator_25_34 <- survey_design %>% filter(age >= 25, age <= 34) %>%
  group_by(race_ethnicity) %>%
  summarise(total_students = survey_total(vartype = "se"))


# Calculate percentage of students attending college within each racial and ethnic group
college_attendance_25_34 <- left_join(numerator_25_34, denominator_25_34, by = "race_ethnicity") %>%
  mutate(percentage_college_students = (college_attendees / total_students) * 100) %>% mutate(age = "25_34", percentage_college_students = paste0(format(round(percentage_college_students, 1), nsmall = 1), "%"))


# combine

df_college <- rbind(college_attendance_18_24, college_attendance_25_34)

```



```{r}
table_2g <- df_college %>% pivot_wider(id_cols =  c(race_ethnicity), names_from = "age", values_from = c(college_attendees, percentage_college_students)) %>% select(race_ethnicity,  college_attendees_18_24,  percentage_college_students_18_24,  college_attendees_25_34,  percentage_college_students_25_34) %>%
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("", "Number", "Percent" , "Number", "Percent"),
      align = c("l",rep("r",4)),
      caption = "Number and Percent of Population Attending Undergraduate or Graduate Colleges or Professional Schools for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em")  %>%
  column_spec(2:5, width = "7.7em")  %>%
  add_header_above( align = c("l", "c", "c"), c("Race/Ethnicity", "Ages 18-24" = 2, "Ages 25-34" = 2),
                    # background = "#cbc9e2", 
                    bold = T, font_size =8) %>% 
  
   row_spec(0, align = "C") %>%  # Center column headers

   add_footnote(
    c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
    notation = "none",
    threeparttable = TRUE,
    escape = FALSE
  )

table_2g

```



\blandscape

## 2.H: Educational Attainment of Population Aged 25 and over for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
df <- usa_data  %>%  filter(year == 2022) %>% select(race_ethnicity, educd, age, perwt)

# collapse educdational categories
df <- df %>%
  mutate(educd = as.character(educd)) %>%
  mutate(education_level = case_when(
    educd %in% c("1") ~ "NA",
    educd %in% c("2") ~ "No Schooling Completed",
    educd %in% c("11", "12", "14", "15", "16", "17", "22", "23", "25", "26", "30", "40", "50", "61") ~ "Less than High School",
    educd %in% c("63", "64") ~ "Regular High School Diploma, GED or Alternative Credential",
    educd %in% c("65", "71") ~ "Some College, No Degree", 
    educd %in% c("81") ~ "Associate's Degree",
    educd %in% c("101") ~ "Bachelor's Degree",
    educd %in% c("114", "115") ~ "Master's Degree or Professional Degree Beyond Bachelor's",
    TRUE ~ "Other"  # Default case if educd is not in the specified categories
  ))


# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)

# Calculate numerator: weighted count of individuals by education level and race_ethnicity
numerator <- survey_design %>%
  filter(age >= 25) %>%
  group_by(race_ethnicity, education_level) %>%
  summarise(weighted_n = survey_total(vartype = "se"))

# Calculate denominator: total population by race_ethnicity
denominator <- survey_design %>%
  filter(age >= 25) %>%
  group_by(race_ethnicity) %>%
  summarise(total_pop = survey_total(vartype = "se"))

# Merge numerator and denominator by race_ethnicity
education_levels <- c("No Schooling Completed", "Less than High School", "Regular High School Diploma, GED or Alternative Credential","Some College, No Degree", "Associate's Degree", "Bachelor's Degree", "Master's Degree or Professional Degree Beyond Bachelor's")

education_rates <- numerator %>%
  left_join(denominator, by = "race_ethnicity") %>%
  mutate(educational_rate = (weighted_n / total_pop) * 100,
  education_level = factor(education_level, levels = education_levels))    %>%  arrange(race_ethnicity, education_level)  %>% mutate(educational_rate = paste0(format(round(educational_rate, 1), nsmall = 1), "%")) %>% filter(!is.na(education_level))
  
```



```{r}
table_2h <- education_rates %>% select(race_ethnicity, education_level, weighted_n, educational_rate)  %>% pivot_wider(id_cols =  c(education_level), names_from = "race_ethnicity", values_from = c(weighted_n, educational_rate))   %>% select(
    education_level, 
     "weighted_n_Mexican", "educational_rate_Mexican",
    'weighted_n_Other Hispanics or Latinos', 'educational_rate_Other Hispanics or Latinos', 
      'weighted_n_White (non-Hispanic or Latino)', 'educational_rate_White (non-Hispanic or Latino)',
      'weighted_n_Black (non-Hispanic or Latino)', 'educational_rate_Black (non-Hispanic or Latino)',
       'weighted_n_Other (non-Hispanic or Latino)',  'educational_rate_Other (non-Hispanic or Latino)'   
  ) %>%


 
   # using kable 
  
  kbl(
  , booktabs = TRUE,
  col.names = c("", "Number", "Percent", "Number", "Percent", "Number", "Percent", "Number", "Percent", "Number", "Percent"),
  align = c("l", rep("r", 10)),
  caption = "Educational Attainment of Population Aged 25 and over for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
  format.args = list(big.mark = ","),
  linesep = ""
) %>%
  kable_classic(
    full_width = FALSE,
    font_size = 8,
    html_font = "Arial",
    position = "center",
    latex_options = "HOLD_position"
  ) %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:11, width = "3.08em") %>%
  add_header_above(
    align = c("l", rep("c", 5)),
    c(
      "Educational Level", 
      "Mexican" = 2, 
      "Other Hispanics or Latinos" = 2, 
      "White (non-Hispanic or Latino)" = 2, 
      "Black (non-Hispanic or Latino)" = 2, 
      "Other (non-Hispanic or Latino)" = 2
    ),
    bold = TRUE, 
    font_size = 8
  ) %>%  row_spec(0, align = "C") %>%  # Center column headers
    add_footnote(
    c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
    notation = "none",
    threeparttable = TRUE,
    escape = FALSE
  )

table_2h

```


\elandscape




\blandscape

## 2.I: Educational Attainment of Population Aged 25 and over for Mexico Born and U.S. Born Mexicans in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
df <- usa_data  %>%  filter(year == 2022) %>% select(mex_foreign_born, educd, age, perwt) %>% filter(!is.na(mex_foreign_born))

# collapse educdational categories
df <- df %>%
  mutate(educd = as.character(educd)) %>%
  mutate(education_level = case_when(
    educd %in% c("1") ~ "NA",
    educd %in% c("2") ~ "No Schooling Completed",
    educd %in% c("11", "12", "14", "15", "16", "17", "22", "23", "25", "26", "30", "40", "50", "61") ~ "Less than High School",
    educd %in% c("63", "64") ~ "Regular High School Diploma, GED or Alternative Credential",
    educd %in% c("65", "71") ~ "Some College, No Degree", 
    educd %in% c("81") ~ "Associate's Degree",
    educd %in% c("101") ~ "Bachelor's Degree",
    educd %in% c("114", "115") ~ "Master's Degree or Professional Degree Beyond Bachelor's",
    TRUE ~ "Other"  # Default case if educd is not in the specified categories
  ))


# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)

# Calculate numerator: weighted count of individuals by education level and mex_foreign_born
numerator <- survey_design %>%
  filter(age >= 25) %>%
  group_by(mex_foreign_born, education_level) %>%
  summarise(weighted_n = survey_total(vartype = "se"))

# Calculate denominator: total population by mex_foreign_born
denominator <- survey_design %>%
  filter(age >= 25) %>%
  group_by(mex_foreign_born) %>%
  summarise(total_pop = survey_total(vartype = "se"))

# Merge numerator and denominator by mex_foreign_born
education_levels <- c("No Schooling Completed", "Less than High School", "Regular High School Diploma, GED or Alternative Credential","Some College, No Degree", "Associate's Degree", "Bachelor's Degree", "Master's Degree or Professional Degree Beyond Bachelor's")

education_rates <- numerator %>%
  left_join(denominator, by = "mex_foreign_born") %>%
  mutate(educational_rate = (weighted_n / total_pop) * 100,
  education_level = factor(education_level, levels = education_levels))    %>%  arrange(mex_foreign_born, education_level)  %>% mutate(educational_rate = paste0(format(round(educational_rate, 1), nsmall = 1), "%")) %>% filter(!is.na(education_level))
  
```



```{r}
table_2i <- education_rates %>% select(mex_foreign_born, education_level, weighted_n, educational_rate)  %>% pivot_wider(id_cols =  c(education_level), names_from = "mex_foreign_born", values_from = c(weighted_n, educational_rate))   %>% select(
    education_level, 
    "weighted_n_Mexico Born",    "educational_rate_Mexico Born",
    "weighted_n_U.S. Born",   "educational_rate_U.S. Born"
  ) %>%


 
   # using kable 
  
  kbl(
  , booktabs = TRUE,
  col.names = c("", "Number", "Percent", "Number", "Percent"),
  align = c("l", rep("r", 4)),
  caption = "Educational Attainment of Population Aged 25 and over for Mexico Born and U.S. Born Mexicans in Chicago, 2018-2022 (ACS 5-year Estimates)",
  format.args = list(big.mark = ","),
  linesep = ""
) %>%
  kable_classic(
    full_width = FALSE,
    font_size = 8,
    html_font = "Arial",
    position = "center",
    latex_options = "HOLD_position"
  ) %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "7.7em") %>%
   add_header_above(
    align = c("l", rep("c", 2)),
    c(
      "Educational Level", 
      "Mexicans Born\nin Mexico" = 2, 
      "Mexicans Born\nin the U.S." = 2
    ),
    bold = TRUE, 
    font_size = 8
  ) %>% row_spec(0, align = "C") %>%  # Center column headers
    add_footnote(
    c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
    notation = "none",
    threeparttable = TRUE,
    escape = FALSE
  )

table_2i

```


\elandscape


# 4. Health 

## 4A: Health Outcomes in Top 4 Community Areas with the Largest Mexican Population Share, (2017-2021 and 2022-2023):




```{r}
# analysis
mexican_top <- df_health %>% arrange(desc(prc_mxc)) %>% top_n(4, prc_mxc) %>% select(commnty, obesity_rate, diabetes_rate, asthma_rate, lbw_rate)   %>%
mutate(
    obesity_rate = paste0(format(round(obesity_rate, 1), nsmall = 1), "%"),
    diabetes_rate = paste0(format(round(diabetes_rate, 1), nsmall = 1), "%"),
    asthma_rate = paste0(format(round(asthma_rate, 1), nsmall = 1), "%"),
    lbw_rate = paste0(format(round(lbw_rate, 1), nsmall = 1), "%")
)


df_white_city <- data.frame(
  commnty = "White Chicago Mean",
 obesity_rate = 24.5,
  diabetes_rate = 6.9, 
  asthma_rate = 9.1,
  lbw_rate = 6.2
)

df_black_city <- data.frame(
  commnty = "Black Chicago Mean",
  obesity_rate = 47.6,
  diabetes_rate = 19.6,
  asthma_rate = 14.9, 
  lbw_rate  =14.6
)

# Create a data frame
df_latino_city <- data.frame(
  commnty = "Hispanic or Latino Chicago Mean",
  obesity_rate = 39.9,
  diabetes_rate = 13.9, 
  asthma_rate = 9.7, 
  lbw_rate = 8.0
)

df_city_avgs <- rbind(df_white_city, df_black_city, df_latino_city)  %>%
mutate(
    obesity_rate = paste0(format(round(obesity_rate, 1), nsmall = 1), "%"),
    diabetes_rate = paste0(format(round(diabetes_rate, 1), nsmall = 1), "%"),
    asthma_rate = paste0(format(round(asthma_rate, 1), nsmall = 1), "%"),
    lbw_rate = paste0(format(round(lbw_rate, 1), nsmall = 1), "%")
)

df_final <- rbind(mexican_top, df_city_avgs) %>%  mutate(across(
    ends_with("Rate"),
    ~ ifelse(. == "  NA%", "-", .)
  ))


```


```{r}

# visuals
table_4a <- df_final %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Four Communities with the Largest Share of Mexican Population", "Adult Obesity Rate(2022-2023)", "Adult Diabetes Rate(2022-2023)", "Adult Asthma Rate(2022-2023)", "Low Birthweight Rate(2017-2021)"),
      align = c("l",rep("c",4)),
      caption = "Health Outcomes for Four Communities with the Largest Share of Mexican Population and by Race/Ethnicity in Chicago",
      format.args = list(big.mark = "%"),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "6.45em") %>%  
row_spec(0, bold = TRUE, font_size = 8) %>%
  row_spec(5:7, bold = TRUE, background = "#f2f0f7") %>%
  
   add_footnote(
  c(
    "\\footnotesize{Data Source:Chicago Health Atlas, Chicago Department of Public Health (20222023) and Illinois Department of Public Health (2017-2021).}",
    "\\footnotesize{Tabulations by Great Cities Institute}",
      "\\footnotesize{Note: Obesity is defined as percent of adults who reported a height and weight that yield a body mass index of 30 or greater.}",
    "\\footnotesize{Note: Diabetes is defined as percent of adults who reported that a doctor, nurse or other health professional has diagnosed them with diabetes.}",
    "\\footnotesize{Note: Asthma is defined as percent of adults who reported that a doctor, nurse or other health professional has diagnosed them with asthma.}",
    "\\footnotesize{Note:Low birthweight is defined as percent of births with a birthweight less than 2500 grams (5.5 pounds).}"
    
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)


table_4a

```


## 4B: No Health Insurance for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis
## select variables, filter for year 2022
df <- usa_data %>%  filter(year == 2022) %>% select(race_ethnicity, hcovany, perwt)

# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)
  
  
# Calculate numerator: weighted count of individuals without health insurance by race_ethnicity
numerator_without_insurance <- survey_design %>%
  filter(hcovany == 1) %>% 
  group_by(race_ethnicity) %>%
  summarise(weighted_n_without_insurance = survey_total(vartype = "se"))

# Calculate denominator: total population by race_ethnicity
denominator <- survey_design %>%
  group_by(race_ethnicity) %>%
  summarise(total_pop = survey_total(vartype = "se")) 
  
# calculate rate

# Join numerator and denominator
df_rates <- numerator_without_insurance %>%
  left_join(denominator, by = "race_ethnicity") %>%
  mutate(
    no_insurance_rate = (weighted_n_without_insurance / total_pop) * 100,
    no_insurance_rate = paste0(format(round(no_insurance_rate, 1), nsmall = 1), "%"), 
      unininsured_sum = sum(weighted_n_without_insurance),
    no_insurance_rate_2 = (weighted_n_without_insurance / unininsured_sum) * 100,
    no_insurance_rate_2 = paste0(format(round(no_insurance_rate_2, 1), nsmall = 1), "%")

  )
  

```


```{r}
# visuals
table_4b <- df_rates  %>% select(race_ethnicity, weighted_n_without_insurance, no_insurance_rate, no_insurance_rate_2)   %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number Without Health Insurance", "No Insurance Rate", "Percent Share of the Uninsured Population"),
      align = c("l",rep("r",3)),
      caption = "No Health Insurance for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "8.6em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

  add_footnote(
    c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
    notation = "none",
    threeparttable = TRUE,
    escape = FALSE
  )


table_4b

```


## 4C: Number and Percent of Mexico Born and U.S. Born Mexicans Without Job Provided Health Insurance in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
## select variables, filter for year 2022
df <- usa_data %>%  filter(year == 2022) %>% select(mex_foreign_born, hinsemp, perwt) %>% filter(!is.na(mex_foreign_born))

# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)
  
  
# Calculate numerator: weighted count of individuals without job provided health insurance by race_ethnicity
numerator_without_insurance <- survey_design %>%
  filter(hinsemp == 1) %>% 
  group_by(mex_foreign_born) %>%
  summarise(weighted_n_without_insurance = survey_total(vartype = "se"))

# Calculate denominator: total population by race_ethnicity
denominator <- survey_design %>%
  group_by(mex_foreign_born) %>%
  summarise(total_pop = survey_total(vartype = "se")) 
  
# calculate rate

# Join numerator and denominator
df_rates <- numerator_without_insurance %>%
  left_join(denominator, by = "mex_foreign_born") %>%
  mutate(
    no_insurance_rate = (weighted_n_without_insurance / total_pop) * 100,
    no_insurance_rate = paste0(format(round(no_insurance_rate, 1), nsmall = 1), "%")


  )
  
```

```{r}
# visuals
table_4c <- df_rates  %>% select(mex_foreign_born, weighted_n_without_insurance, no_insurance_rate)   %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Mexican Nativity", "Number Without Job Provided Health Insurance", "No Job Provided Health Insurance"),
      align = c("l",rep("r",2)),
      caption = "Number and Percent of Mexico Born and U.S. Born Mexicans Without Job Provided Health Insurance in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "8.6em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

    footnote(general = "Source: IPUMS-USA database (2018-2022 American Community Survey 5-Year Estimates). Tabulations by Great Cities Institute",
           general_title = "")    


table_4c

```


## 4E: Mental Health Outcomes for Four Communities with the Largest Share of Mexican Population and by Race/Ethnicity in Chicago, 2022-2023

```{r}
# analysis
mexican_top <- df_health %>% arrange(desc(prc_mxc)) %>% top_n(4, prc_mxc) %>% select(commnty, psych_distress_rate, unmet_mental_health_rate, loneliness_rate)   %>%
 mutate(
    psych_distress_rate = paste0(round(psych_distress_rate, 1), "%"),
    unmet_mental_health_rate = paste0(round(unmet_mental_health_rate, 1), "%"),
    loneliness_rate = paste0(round(loneliness_rate, 1), "%")
)



df_white_city <- data.frame(
  commnty = "White Chicago Mean",
 psych_distress_rate = 7.5,
  unmet_mental_health_rate = 57.2,
  loneliness_rate = 26.5
)

df_black_city <- data.frame(
  commnty = "Black Chicago Mean",
  psych_distress_rate = 9.9,
  unmet_mental_health_rate = 81.4,
  loneliness_rate = 30.2
)

# Create a data frame
df_latino_city <- data.frame(
  commnty = "Hispanic or Latino Chicago Mean",
  psych_distress_rate = 14.9,
  unmet_mental_health_rate = 80.0,
  loneliness_rate = 29.9
)

df_city_avgs <- rbind(df_white_city, df_black_city, df_latino_city)  %>%
  mutate(
    psych_distress_rate = paste0(sprintf("%.1f", psych_distress_rate), "%"),
    unmet_mental_health_rate = paste0(sprintf("%.1f", unmet_mental_health_rate), "%"),
    loneliness_rate = paste0(sprintf("%.1f", loneliness_rate), "%")
  )


df_final <- rbind(mexican_top, df_city_avgs)
```


```{r}
# visuals
table_4e <- df_final %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Four Communities with the Largest Share of Mexican Population", "Adult Serious Psych Distress Rate(2022-2023)", "Adult Unmet Mental Health Need Rate(2022-2023)", "Adult Loneliness Rate(2022-2023)"),
      align = c("l",rep("r",3)),
      caption = "Mental Health Outcomes for Four Communities with the Largest Share of Mexican Population and by Race/Ethnicity in Chicago, 2022-2023",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "9.2em") %>%  
row_spec(0, bold = TRUE, font_size = 8) %>%
  row_spec(5:7, bold = TRUE, background = "#f2f0f7") %>%
  
  add_footnote(
  c(
    "\\footnotesize{Data Source:Chicago Health Atlas, Chicago Department of Public Health (20222023) and Illinois Department of Public Health (2017-2021).}",
    "\\footnotesize{Tabulations by Great Cities Institute}",
      "\\footnotesize{Note: Serious Psychological Distress is the percent of adults who experienced feelings of nervousness, hopelessness, and depression, in the past 30 days.}",
    "\\footnotesize{Note: Unmet mental health need is the percent of adults with psychological distress who are not receiving treatment or medication.}",
    "\\footnotesize{Note: Loneliness rate is defined as the percent of adults who reported feeling left out or felt alone}"
    
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)



table_4e



```


## 4F: Chronic Health Indicators in top 4 Mexican Community Areas,(2017-2021, 2022-2023):

```{r}
# Analysis
mexican_top <- df_health %>% 
  arrange(desc(prc_mxc)) %>% 
  top_n(4, prc_mxc) %>% 
  select(commnty, cancer_diagnosis_rate, hyper_tension_rate, lung_cancer_rate) %>% 
mutate(
    cancer_diagnosis_rate = format(round(cancer_diagnosis_rate, 1), nsmall = 1),
    hyper_tension_rate = paste0(format(round(hyper_tension_rate, 1), nsmall = 1), "%"),
    lung_cancer_rate = format(round(lung_cancer_rate, 1), nsmall = 1)
)

df_white_city <- data.frame(
  commnty = "White City Mean",
  cancer_diagnosis_rate = 676.16, 
  hyper_tension_rate = 29.8, 
  lung_cancer_rate = 83.41
)

df_black_city <- data.frame(
  commnty = "Black City Mean",
  cancer_diagnosis_rate = 529.94, 
  hyper_tension_rate = 43.9, 
  lung_cancer_rate = 77.39
)

df_latino_city <- data.frame(
  commnty = "Hispanic/Latino City Mean",
  cancer_diagnosis_rate = 253.29, 
  hyper_tension_rate = 25.7, 
  lung_cancer_rate = 13.93
)

df_city_avgs <- rbind(df_white_city, df_black_city, df_latino_city) %>% 
 mutate(
    cancer_diagnosis_rate = format(round(cancer_diagnosis_rate, 1), nsmall = 1),
    hyper_tension_rate = paste0(format(round(hyper_tension_rate, 1), nsmall = 1), "%"),
    lung_cancer_rate = format(round(lung_cancer_rate, 1), nsmall = 1)
)


df_final <- rbind(mexican_top, df_city_avgs)

```



```{r}
# visuals
table_4f <- df_final %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Top 4 Community Areas with the Largest Mexican Population Share", "Adult Cancer Diagnosis Rate per 100,000 (2017-2021)", "Adult Hypertension Rate (2022-2023)", "Adult Lung Cancer Diagnosis Rate per 100,000 (2017-2021)"),
      align = c("l",rep("r",3)),
      caption = "Chronic Health Indicators for Four Communities with the Largest Share of Mexican Population and by Race/Ethnicity in Chicago",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "9.2em") %>%  
row_spec(0, bold = TRUE, font_size = 8) %>%
  row_spec(5:7, bold = TRUE, background = "#f2f0f7") %>%   
  
    add_footnote(
  c(
    "\\footnotesize{Data Source:Chicago Health Atlas, Chicago Department of Public Health (20222023) and Illinois Department of Public Health (2017-2021).}",
    "\\footnotesize{Tabulations by Great Cities Institute}",
      "\\footnotesize{Note: Cancer diagnosis is defined as the annual diagnosis rate for all invasive cancers, excluding pre-cancerous conditions.  All ages, risk-adjusted.}",
    "\\footnotesize{Note: Hypertension is the percentage of adults diagnosed with high blood pressure, excluding borderline or pregnancy-related cases.}",
    "\\footnotesize{Note: Lung cancer is defined as lung and bronchus cancer for ages 15 and over, risk-adjusted.}"
    
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)
  


table_4f
```

# 6.Citizenship and Civic Participation


## 6B: Illinois Voting



```{r}
# Create vectors for each column
state <- c("New Jersey", "Michigan", "Connecticut", "Arizona", "Virginia", 
           "Texas", "New York", "Pennsylvania", "Washington", "Massachusetts", 
           "Colorado", "California", "New Mexico", "Florida", "North Carolina", 
           "Illinois", "Georgia")

citizens <- c(996000, 302000, 347000, 1340000, 425000, 
              5599000, 1608000, 497000, 485000, 449000, 
              618000, 8305000, 539000, 3394000, 492000, 
              1016000, 403000)

percent_registered <- c(82, 74.7, 67.8, 66.8, 63.8, 
                        63.2, 61.6, 61.4, 61, 60.4, 
                        60.5, 60.4, 59.9, 58.7, 54.3, 
                        52.4, 47.6)

percent_voting <- c(72.1, 67.9, 56.4, 60.8, 51.3, 
                    53.1, 54.9, 54.3, 53.7, 50.7, 
                    51.1, 60.8, 45.6, 52.7, 48.8, 
                    46.8, 44.2)

rank_registered <- c(1, 2, 3, 4, 5, 
                     6, 7, 8, 9, 10, 
                     11, 12, 13, 14, 15, 
                     16, 17)

rank_voting <- c(1, 2, 5, 3, 11, 
                 9, 6, 7, 8, 13, 
                 12, 3, 16, 10, 14, 
                 15, 17)

# Create the DataFrame
election_df <- data.frame(State = state,
                           # Convert numeric vectors to appropriate data types
                           Citizens = citizens,
                           Percent_Registered = percent_registered,
                           Percent_Voting = percent_voting,
                           Rank_Registered = rank_registered,
                           Rank_Voting = rank_voting)



# kable

  # using kable
table_6b <- election_df %>%  kbl(, booktabs = T,
      col.names = c("State", "# of Citizens", "% Registered" , "% Voting", "Rank Registered", "Rank Voting"),
      align = c("l", rep("r", 5)),
      caption = "2020 Presidential Election /Latino Registration and Voting By States",
     format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "12.5em") %>%
  column_spec(2:5, width = "8.1em") %>%
   row_spec(16, background = "#ADD8E6") %>%  # Highlight 16th row in light blue
  add_footnote(
  c("\\footnotesize{Data source: U.S. Census Bureau, 'Voting and Registration in Election of November 2020,' Table 4a 'Voting and Registration By States,' GCI tabulation of states with more than 300,000 Latino citizens over age 18}"),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)


table_6b
```









## 6C: Chicago Voting

```{r}
# Define the data
category <- c("Percent vote for Brandon Johnson", 
              "Percent registered voter turnout", 
              "Percent of population aged 18 and over that voted")

hispanic_latino <- c("45.6%", "20.5%", "11.3%")
black_non_hispanic_latino <- c("88.0%", "29.0%", "24.5%")
white_non_hispanic_latino <- c("34.0%", "61.1%", "52.7%")
total <- c("52.0%", "38.7%", "28.0%")

# Create the DataFrame
voting_df <- data.frame(
  Category = category,
  Hispanic_or_Latino = hispanic_latino,
  Black_non_Hispanic_or_Latino = black_non_hispanic_latino,
  White_non_Hispanic_or_Latino = white_non_hispanic_latino,
  Total = total,
  stringsAsFactors = FALSE  # Avoid converting strings to factors
)


  # using kable
table_6c <- voting_df %>%  kbl(, booktabs = T,
      col.names = c("", "Hispanic \nor Latino", "Black  (non-Hispanic or Latino)" , "White (non-Hispanic or Latino)", "Total"),
      align = c("l", rep("r", 4)),
      caption = "Estimates of percent vote for Brandon Johnson, voter turnout, and percent of adult population that voted",
     format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "12.5em", bold = TRUE) %>%
  column_spec(2:5, width = "8.1em") %>%
  add_footnote(
  c("\\footnotesize{Data Source: Chicago's 2023 Mayoral Race: A Progressive Victory Amidst Shocking Low Turnout by Black and Latino Voters (Calculations from Chicago Board of Election Commissioners 2023 Municipal Runoff Election Results and 2020 U.S. Census.)}"),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)


table_6c

```



