---
title: "Mexicans in Chicago Report Committee"
author: ""
date: "Great Cities Institute"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    highlight: pygments
    extra_dependencies: ["float"]
    fig_caption: true
mainfont: "Arial"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F, fig.pos="H")


```


```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}

library(ipumsr)
library(tidyverse)
library(purrr)
library(sf)
library(tidycensus)
library(tidyr)
library(readxl)
library(sf)
library(htmltools)
library(leaflet)
library(janitor)
library(data.table)
library(tinytex)
library(survey) # for survey analysis
library(srvyr)  # for tidy survey analysis
library(tmap)
library("kableExtra") 
library(tmaptools)
library("RColorBrewer") 
# library(OpenStreetMap)
# remotes::install_github('yihui/knitr')
# install.packages("knitr")
# library(knitr)
# options(tinytex.verbose = TRUE)
# tinytex::reinstall_tinytex()
# install.packages('tinytex')

# update.packages(ask = FALSE, checkBuilt = TRUE)
# tinytex::tlmgr_update()
# tinytex::latexmk("Housing_Visual.tex")


##### Chicago boarder
Chicago_boundary_shp <- read_sf("C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/Boundaries_Chicago/geo_export_8881adc5-2460-457b-892f-04f4d6028b72.shp") %>% 
  st_transform(crs = 4326)
# mapview(community_shp)


# Get the current bounding box of the data
bbox_data <- st_bbox(Chicago_boundary_shp)

# Calculate the range of x and y values in degrees
xrange <- bbox_data$xmax - bbox_data$xmin
yrange <- bbox_data$ymax - bbox_data$ymin

# Calculate the center of the bounding box
center_x <- (bbox_data$xmin + bbox_data$xmax) / 2
center_y <- (bbox_data$ymin + bbox_data$ymax) / 2
# Modify the bounding box to fit the desired width while keeping the height the same
bbox_new <- c(center_x - 0.24,
              center_y - 0.23,
              center_x + 0.23,
              center_y + 0.23)

# osm_chicago <- read_osm(bbox_new, 
#                         type = "bing")


theme_plots <-
  theme_minimal() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.line = element_line(color="#969696", linewidth = 1),
        plot.title = element_text(face ="bold", size = 10, hjust = 0.5),
      plot.subtitle = element_text(face ="italic", size = 8, color = "#636363"),
      axis.text = element_text(size = 9),
      axis.title = element_text(face ="bold", size = 9),
      plot.caption = element_text(hjust = 0, size = 8, color = "black"),
      legend.text = element_text(size = 8, color = "#636363"),
      legend.title = element_text(face = "bold", size = 8, color = "#636363")) 

```

\renewcommand{\arraystretch}{1.4}
\clearpage

```{r}

usa_data <- read.csv("C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/final_Mexican_IL_2000_22.csv") %>% 
  mutate(race_ethnicity = case_when(hispan ==0 & race == 1 ~ "White (non-Hispanic or Latino)",
                                    hispan ==0 & race == 2 ~ "Black (non-Hispanic or Latino)",
                                    hispan %in% c(2,3,4) ~ "Other Latinos",
                                    hispan ==1 ~ "Mexican", 
                                    hispan ==0 & race %in%c(3:9) ~ "Other (non-Hispanic or Latino)",
                                    TRUE ~ NA_character_)) %>% 
  mutate(race_ethnicity = factor(race_ethnicity, level = c("Mexican", "Other Latinos","White (non-Hispanic or Latino)", "Black (non-Hispanic or Latino)",  "Other (non-Hispanic or Latino)")))

cps_ddi <- read_ipums_micro(read_ipums_ddi("C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/cps_00170.xml")) %>% 
  clean_names() 
```


# Demographic

## 1.A: Mexican population in Chicago:

## 1. Mexican population in Chicago, 2018-2022:

```{r, include=FALSE}
####### 2018-2022 ##################
comm_2022 <- st_read("Data Tables/shp/comm_pop_chi_2022.shp") %>% 
  mutate(lable_prc = paste0(lable, ": ", prc),
         prc2 = paste0( prc, "%")) %>% 
  mutate(ttl_mxc2 = scales::comma(ttl_mxc))

tract_2022 <- st_read("Data Tables/shp/tract_pop_chi_2022.shp") 

pop_dots_2022_origin_50 <- read_sf("Data Tables/shp/pop_dots_2022_origin_50.shp") 


Mexican_18_22 <- pop_dots_2022_origin_50 %>% 
  filter(Origin == 'Mexican')
####################################
```


```{r fig.height=11, fig.width=8}

### Dot Density
# st_write(Mexican, "data/2020/origin_Mexican_2020_50.shp")
Mexican_map_2020 <-

  # tm_shape(community_shp) +
  # tm_polygons(col = "white",
  #            alpha = 0.3)+
  
  tm_shape(Mexican_18_22,  bbox = bbox_new) +
  tm_dots(col = "Origin",
          palette = "#1565c0",
          size = 0.025,
          legend.show = FALSE,
          alpha = 0.3) +
  
  tm_shape(comm_2022) +
  tm_borders(col = "#6c757d",
             alpha = 1,
             lwd = 1)+
  tm_text("lable" , size = 0.6, alpha = 0.9, fontface="bold", fontfamily = "sans")+

  tm_shape(Chicago_boundary_shp) +
  tm_borders(col = "#343a40",
             alpha = 0.7,
             lwd = 2.5) +

  tm_layout( frame = TRUE,
             # main.title.position = c("left", "top"),
             # main.title = "Density of Hispanic or Latino Population by their Origins in Chicago, 2010",
             # main.title.size = 1.2,
            legend.outside = FALSE,
            # legend.width = 1.5,
            legend.title.size = .9,
            legend.text.size = 0.8,
            legend.title.fontface = "bold",
            # legend.position = c("left", "bottom"),
            legend.position = c(0.03, 0.12),
            legend.bg.color = "white",
            legend.bg.alpha = .3) +       # manually position legend) +
  tm_credits("Data source:\nIPUMS NHGIS (2018-2022)\nMap Generated by:\nThe Great Cities Institute",
             fontface = "bold",
             position = c("left","bottom"), size = 0.8)+
  
  tm_credits("Distribution of Mexican Population in Chicago, 2018-2022",
             fontface = "bold",
             position = c("left","top"), size = 1.25)+

  
  tm_add_legend("symbol", 
                labels = "Mexican", 
                col = "#1565c0",
                size = 0.6,
                title = "1 dot = 50 People") +
  
  tm_add_legend("line",
                labels = c("Chicago Border", "Community Area Border"),
                col = c("#343a40", "#6c757d"),
                lwd = c(2.5, 2),
                title = "Border Types",
                size = 0.8) +

  # tm_credits("1 dot = 20 People",
  #            position = c(0.03, 0.32), size = 0.8, fontface = "bold") +
  tm_compass(position = c("right", "bottom"), size = 1) +
  tm_scale_bar(position = c("right", "bottom"), width = 0.15)

Mexican_map_2020

```







```{r fig.height=11, fig.width=8}
### Percentage of the Mexican Population by Community Areas in Chicago, 2018-2022
comm_mexican_map_2022 <-

  tm_shape(comm_2022, bbox = bbox_new) +
  tm_polygons(
    col = "prc",
              style = "jenks",
              border.col = "#ced4da",
              palette="Blues",
              alpha = 0.7,
              n = 5,
              title = "Percentage of the Mexican Population",
              lwd = 0.3) +
  tm_shape(comm_2022) +
  tm_borders(col = "#6c757d",
             alpha = 1,
             lwd = 1)+
  
  # tm_shape(comm_2022 %>% filter(prc>=50)) +
  # tm_borders(col = "#9b2226",
  #            alpha = 1,
  #            lwd = 2)+
  tm_text("lable" , size = 0.6, alpha = 0.7, fontface="bold", fontfamily = "sans")+

  tm_shape(Chicago_boundary_shp) +
  tm_borders(col = "#343a40",
             alpha = 1,
             lwd = 2.5) +

  tm_layout( frame = TRUE,
             # main.title.position = c("left", "top"),
             # main.title = "Density of Hispanic or Latino Population by their Origins in Chicago, 2010",
             # main.title.size = 1.2,
            legend.outside = FALSE,
            legend.width = 1.5,
            legend.title.size = .9,
            legend.text.size = 0.8,
            legend.title.fontface = "bold",
            # legend.position = c("left", "bottom"),
            legend.position = c(0.03, 0.12),
            legend.bg.color = "white",
            legend.bg.alpha = .3) +      # manually position legend) +
  
  tm_credits("Data source:\nIPUMS NHGIS (2018-2022)\nMap Generated by:The Great Cities Institute",
             fontface = "bold",
             position = c("left","bottom"), size = 0.8)+


  tm_credits("Percentage of the Mexican Population by Community Areas\nin Chicago, 2018-2022",
             fontface = "bold",
             position = c("left","top"), size = 1.25)+


  tm_add_legend("line",
                labels = c("Chicago Border", "Community Area Border"),
                col = c("#343a40","#6c757d"),
                lwd = c(2.5, 2),
                title = "Border Types",
                size = 0.8) +

  # tm_credits("1 dot = 20 People",
  #            position = c(0.03, 0.32), size = 0.8, fontface = "bold") +
  tm_compass(position = c("right", "bottom"), size = 1) +
  tm_scale_bar(position = c("right", "bottom"), width = 0.15)

comm_mexican_map_2022


```


```{r fig.height=11, fig.width=8}
### Mexican Population by Community Areas in Chicago, 2018-2022
comm_mexican_pop_map_2022 <-

  tm_shape(comm_2022, bbox = bbox_new) +
  tm_polygons(
    col = "ttl_mxc",
              style = "jenks",
              border.col = "#ced4da",
              palette="Blues",
              alpha = 0.7,
              n = 5,
              title = "Mexican Population",
              lwd = 0.3) +
  tm_shape(comm_2022) +
  tm_borders(col = "#6c757d",
             alpha = 1,
             lwd = 1)+
  
  # tm_shape(comm_2022 %>% filter(prc>=50)) +
  # tm_borders(col = "#9b2226",
  #            alpha = 1,
  #            lwd = 2)+
  tm_text("ttl_mxc2" , size = 0.7, alpha = 0.7, fontface="bold", fontfamily = "sans")+

  tm_shape(Chicago_boundary_shp) +
  tm_borders(col = "#343a40",
             alpha = 1,
             lwd = 2.5) +

  tm_layout( frame = TRUE,
             # main.title.position = c("left", "top"),
             # main.title = "Density of Hispanic or Latino Population by their Origins in Chicago, 2010",
             # main.title.size = 1.2,
            legend.outside = FALSE,
            legend.width = 1.5,
            legend.title.size = .9,
            legend.text.size = 0.8,
            legend.title.fontface = "bold",
            # legend.position = c("left", "bottom"),
            legend.position = c(0.03, 0.12),
            legend.bg.color = "white",
            legend.bg.alpha = .3) +      # manually position legend) +
  
  tm_credits("Data source:\nIPUMS NHGIS (2018-2022)\nMap Generated by:The Great Cities Institute",
             fontface = "bold",
             position = c("left","bottom"), size = 0.8)+


  tm_credits("Mexican Population by Community Areas in Chicago, 2018-2022",
             fontface = "bold",
             position = c("left","top"), size = 1.25)+


  tm_add_legend("line",
                labels = c("Chicago Border", "Community Area Border"),
                col = c("#343a40","#6c757d"),
                lwd = c(2.5, 2),
                title = "Border Types",
                size = 0.8) +

  # tm_credits("1 dot = 20 People",
  #            position = c(0.03, 0.32), size = 0.8, fontface = "bold") +
  tm_compass(position = c("right", "bottom"), size = 1) +
  tm_scale_bar(position = c("right", "bottom"), width = 0.15)

comm_mexican_pop_map_2022

```


## 2. Mexican population in Chicago, 2008-2012:

```{r, include=FALSE}

####### 2010 ######################
comm_2012 <- st_read("Data Tables/shp/comm_pop_chi_2012.shp") %>%
  mutate(lable_prc = paste0(lable, ": ", prc)) %>% 
  mutate(ttl_mxc2 = scales::comma(ttl_mxc))

tract_2012 <- st_read("Data Tables/shp/tract_pop_chi_2012.shp")

pop_dots_2010_origin_50 <- read_sf("Data Tables/shp/pop_dots_2010_origin_50.shp") 

Mexican_2010 <- pop_dots_2010_origin_50 %>% 
  filter(Origin == 'Mexican')
#######################################
```



```{r fig.height=11, fig.width=8}

### Dot Density
Mexican_map_2010 <-
  tm_shape(Mexican_2010,  bbox = bbox_new) +
  tm_dots(col = "Origin",
          palette = "#1565c0",
          size = 0.025,
          legend.show = FALSE,
          alpha = 0.3) +
  
  tm_shape(comm_2012) +
  tm_borders(col = "#6c757d",
             alpha = 1,
             lwd = 1)+
  tm_text("lable" , size = 0.6, alpha = 0.9, fontface="bold", fontfamily = "sans")+

  tm_shape(Chicago_boundary_shp) +
  tm_borders(col = "#343a40",
             alpha = 0.7,
             lwd = 2.5) +

  tm_layout( frame = TRUE,
             # main.title.position = c("left", "top"),
             # main.title = "Density of Hispanic or Latino Population by their Origins in Chicago, 2010",
             # main.title.size = 1.2,
            legend.outside = FALSE,
            # legend.width = 1.5,
            legend.title.size = .9,
            legend.text.size = 0.8,
            legend.title.fontface = "bold",
            # legend.position = c("left", "bottom"),
            legend.position = c(0.03, 0.12),
            legend.bg.color = "white",
            legend.bg.alpha = .3) +       # manually position legend) +
  tm_credits("Data source:\nIPUMS NHGIS (2010)\nMap Generated by:\nThe Great Cities Institute",
             fontface = "bold",
             position = c("left","bottom"), size = 0.8)+
  
  tm_credits("Distribution of Mexican Population in Chicago, 2010",
             fontface = "bold",
             position = c("left","top"), size = 1.25)+

  
  tm_add_legend("symbol", 
                labels = "Mexican", 
                col = "#1565c0",
                size = 0.6,
                title = "1 dot = 50 People") +
  
  tm_add_legend("line",
                labels = c("Chicago Border", "Community Area Border"),
                col = c("#343a40", "#6c757d"),
                lwd = c(2.5, 2),
                title = "Border Types",
                size = 0.8) +

  # tm_credits("1 dot = 20 People",
  #            position = c(0.03, 0.32), size = 0.8, fontface = "bold") +
  tm_compass(position = c("right", "bottom"), size = 1) +
  tm_scale_bar(position = c("right", "bottom"), width = 0.15)

Mexican_map_2010

```







```{r fig.height=11, fig.width=8}
### Percentage of the Mexican Population by Community Areas in Chicago, 2008-2012
comm_mexican_map_2012 <-

  tm_shape(comm_2012, bbox = bbox_new) +
  tm_polygons(
    col = "prc",
              style = "jenks",
              border.col = "#ced4da",
              palette="Blues",
              alpha = 0.7,
              n = 5,
              title = "Percentage of the Mexican Population",
              lwd = 0.3) +
  tm_shape(comm_2012) +
  tm_borders(col = "#6c757d",
             alpha = 1,
             lwd = 1)+
  
  # tm_shape(comm_2022 %>% filter(prc>=50)) +
  # tm_borders(col = "#9b2226",
  #            alpha = 1,
  #            lwd = 2)+
  tm_text("lable" , size = 0.6, alpha = 0.7, fontface="bold", fontfamily = "sans")+

  tm_shape(Chicago_boundary_shp) +
  tm_borders(col = "#343a40",
             alpha = 1,
             lwd = 2.5) +

  tm_layout( frame = TRUE,
             # main.title.position = c("left", "top"),
             # main.title = "Density of Hispanic or Latino Population by their Origins in Chicago, 2010",
             # main.title.size = 1.2,
            legend.outside = FALSE,
            legend.width = 1.5,
            legend.title.size = .9,
            legend.text.size = 0.8,
            legend.title.fontface = "bold",
            # legend.position = c("left", "bottom"),
            legend.position = c(0.03, 0.12),
            legend.bg.color = "white",
            legend.bg.alpha = .3) +      # manually position legend) +
  
  tm_credits("Data source:\nIPUMS NHGIS (2018-2022)\nMap Generated by:The Great Cities Institute",
             fontface = "bold",
             position = c("left","bottom"), size = 0.8)+


  tm_credits("Percentage of the Mexican Population by Community Areas\nin Chicago, 2008-2012",
             fontface = "bold",
             position = c("left","top"), size = 1.25)+


  tm_add_legend("line",
                labels = c("Chicago Border", "Community Area Border"),
                col = c("#343a40","#6c757d"),
                lwd = c(2.5, 2),
                title = "Border Types",
                size = 0.8) +

  # tm_credits("1 dot = 20 People",
  #            position = c(0.03, 0.32), size = 0.8, fontface = "bold") +
  tm_compass(position = c("right", "bottom"), size = 1) +
  tm_scale_bar(position = c("right", "bottom"), width = 0.15)

comm_mexican_map_2012


```





```{r fig.height=11, fig.width=8}
### Mexican Population by Community Areas in Chicago, 2008-2012
comm_mexican_pop_map_2012 <-

  tm_shape(comm_2012, bbox = bbox_new) +
  tm_polygons(
    col = "ttl_mxc",
              style = "jenks",
              border.col = "#ced4da",
              palette="Blues",
              alpha = 0.7,
              n = 5,
              title = "Mexican Population",
              lwd = 0.3) +
  tm_shape(comm_2012) +
  tm_borders(col = "#6c757d",
             alpha = 1,
             lwd = 1) +
  
  tm_text("ttl_mxc2" , size = 0.7, alpha = 0.7, fontface="bold", fontfamily = "sans") +

  tm_shape(Chicago_boundary_shp) +
  tm_borders(col = "#343a40",
             alpha = 1,
             lwd = 2.5) +

  tm_layout( frame = TRUE,
             # main.title.position = c("left", "top"),
             # main.title = "Density of Hispanic or Latino Population by their Origins in Chicago, 2010",
             # main.title.size = 1.2,
            legend.outside = FALSE,
            legend.width = 1.5,
            legend.title.size = .9,
            legend.text.size = 0.8,
            legend.title.fontface = "bold",
            # legend.position = c("left", "bottom"),
            legend.position = c(0.03, 0.12),
            legend.bg.color = "white",
            legend.bg.alpha = .3) +      # manually position legend) +
  
  tm_credits("Data source:\nIPUMS NHGIS (2008-2012)\nMap Generated by:The Great Cities Institute",
             fontface = "bold",
             position = c("left","bottom"), size = 0.8)+


  tm_credits("Mexican Population by Community Areas in Chicago, 2008-2012",
             fontface = "bold",
             position = c("left","top"), size = 1.25)+


  tm_add_legend("line",
                labels = c("Chicago Border", "Community Area Border"),
                col = c("#343a40","#6c757d"),
                lwd = c(2.5, 2),
                title = "Border Types",
                size = 0.8) +

  # tm_credits("1 dot = 20 People",
  #            position = c(0.03, 0.32), size = 0.8, fontface = "bold") +
  tm_compass(position = c("right", "bottom"), size = 1) +
  tm_scale_bar(position = c("right", "bottom"), width = 0.15)

comm_mexican_pop_map_2012

```


## 3. Mexican population in Chicago, 2000:

```{r, include=FALSE}


####### 2000 ######################
comm_2000 <- st_read("Data Tables/shp/comm_pop_chi_2000.shp") %>%
  mutate(lable_prc = paste0(lable, ": ", prc)) %>% 
  mutate(ttl_mxc2 = scales::comma(ttl_mxc))

tract_2000 <- st_read("Data Tables/shp/tract_pop_chi_2000.shp")

pop_dots_2000_origin_50 <- read_sf("Data Tables/shp/pop_dots_2000_origin_50.shp")
Mexican_2000 <- pop_dots_2000_origin_50 %>% 
  filter(Origin == 'Mexican')
###################################

```



```{r fig.height=11, fig.width=8}
### Dot Density
Mexican_map_2000 <-
  tm_shape(Mexican_2000,  bbox = bbox_new) +
  tm_dots(col = "Origin",
          palette = "#1565c0",
          size = 0.025,
          legend.show = FALSE,
          alpha = 0.3) +
  
  tm_shape(comm_2000) +
  tm_borders(col = "#6c757d",
             alpha = 1,
             lwd = 1)+
  tm_text("lable" , size = 0.6, alpha = 0.9, fontface="bold", fontfamily = "sans")+

  tm_shape(Chicago_boundary_shp) +
  tm_borders(col = "#343a40",
             alpha = 0.7,
             lwd = 2.5) +

  tm_layout( frame = TRUE,
             # main.title.position = c("left", "top"),
             # main.title = "Density of Hispanic or Latino Population by their Origins in Chicago, 2010",
             # main.title.size = 1.2,
            legend.outside = FALSE,
            # legend.width = 1.5,
            legend.title.size = .9,
            legend.text.size = 0.8,
            legend.title.fontface = "bold",
            # legend.position = c("left", "bottom"),
            legend.position = c(0.03, 0.12),
            legend.bg.color = "white",
            legend.bg.alpha = .3) +       # manually position legend) +
  tm_credits("Data source:\nIPUMS NHGIS (2000)\nMap Generated by:\nThe Great Cities Institute",
             fontface = "bold",
             position = c("left","bottom"), size = 0.8)+
  
  tm_credits("Distribution of Mexican Population in Chicago, 2000",
             fontface = "bold",
             position = c("left","top"), size = 1.25)+

  
  tm_add_legend("symbol", 
                labels = "Mexican", 
                col = "#1565c0",
                size = 0.6,
                title = "1 dot = 50 People") +
  
  tm_add_legend("line",
                labels = c("Chicago Border", "Community Area Border"),
                col = c("#343a40", "#6c757d"),
                lwd = c(2.5, 2),
                title = "Border Types",
                size = 0.8) +

  # tm_credits("1 dot = 20 People",
  #            position = c(0.03, 0.32), size = 0.8, fontface = "bold") +
  tm_compass(position = c("right", "bottom"), size = 1) +
  tm_scale_bar(position = c("right", "bottom"), width = 0.15)

Mexican_map_2000

```




```{r fig.height=11, fig.width=8}
## Percentage of the Mexican Population by Community Areas in Chicago, 2000
comm_mexican_map_2000 <-

  tm_shape(comm_2000, bbox = bbox_new) +
  tm_polygons(
    col = "prc",
              style = "jenks",
              border.col = "#ced4da",
              palette="Blues",
              alpha = 0.7,
              n = 5,
              title = "Percentage of the Mexican Population",
              lwd = 0.3) +
  tm_shape(comm_2000) +
  tm_borders(col = "#6c757d",
             alpha = 1,
             lwd = 1)+
  
  # tm_shape(comm_2022 %>% filter(prc>=50)) +
  # tm_borders(col = "#9b2226",
  #            alpha = 1,
  #            lwd = 2)+
  tm_text("lable" , size = 0.6, alpha = 0.7, fontface="bold", fontfamily = "sans")+

  tm_shape(Chicago_boundary_shp) +
  tm_borders(col = "#343a40",
             alpha = 1,
             lwd = 2.5) +

  tm_layout( frame = TRUE,
             # main.title.position = c("left", "top"),
             # main.title = "Density of Hispanic or Latino Population by their Origins in Chicago, 2010",
             # main.title.size = 1.2,
            legend.outside = FALSE,
            legend.width = 1.5,
            legend.title.size = .9,
            legend.text.size = 0.8,
            legend.title.fontface = "bold",
            # legend.position = c("left", "bottom"),
            legend.position = c(0.03, 0.12),
            legend.bg.color = "white",
            legend.bg.alpha = .3) +      # manually position legend) +
  
  tm_credits("Data source:\nIPUMS NHGIS (2000)\nMap Generated by:The Great Cities Institute",
             fontface = "bold",
             position = c("left","bottom"), size = 0.8)+


  tm_credits("Percentage of the Mexican Population by Community Areas\nin Chicago, 2000",
             fontface = "bold",
             position = c("left","top"), size = 1.25)+


  tm_add_legend("line",
                labels = c("Chicago Border", "Community Area Border"),
                col = c("#343a40","#6c757d"),
                lwd = c(2.5, 2),
                title = "Border Types",
                size = 0.8) +

  # tm_credits("1 dot = 20 People",
  #            position = c(0.03, 0.32), size = 0.8, fontface = "bold") +
  tm_compass(position = c("right", "bottom"), size = 1) +
  tm_scale_bar(position = c("right", "bottom"), width = 0.15)

comm_mexican_map_2000


```





```{r fig.height=11, fig.width=8}

### Mexican Population by Community Areas in Chicago, 2000
comm_mexican_pop_map_2000 <-

  tm_shape(comm_2000, bbox = bbox_new) +
  tm_polygons(
    col = "ttl_mxc",
              style = "jenks",
              border.col = "#ced4da",
              palette="Blues",
              alpha = 0.7,
              n = 5,
              title = "Mexican Population",
              lwd = 0.3) +
  tm_shape(comm_2000) +
  tm_borders(col = "#6c757d",
             alpha = 1,
             lwd = 1) +
  
  tm_text("ttl_mxc2" , size = 0.7, alpha = 0.7, fontface="bold", fontfamily = "sans") +

  tm_shape(Chicago_boundary_shp) +
  tm_borders(col = "#343a40",
             alpha = 1,
             lwd = 2.5) +

  tm_layout( frame = TRUE,
             # main.title.position = c("left", "top"),
             # main.title = "Density of Hispanic or Latino Population by their Origins in Chicago, 2010",
             # main.title.size = 1.2,
            legend.outside = FALSE,
            legend.width = 1.5,
            legend.title.size = .9,
            legend.text.size = 0.8,
            legend.title.fontface = "bold",
            # legend.position = c("left", "bottom"),
            legend.position = c(0.03, 0.12),
            legend.bg.color = "white",
            legend.bg.alpha = .3) +      # manually position legend) +
  
  tm_credits("Data source:\nIPUMS NHGIS (2000)\nMap Generated by:The Great Cities Institute",
             fontface = "bold",
             position = c("left","bottom"), size = 0.8)+


  tm_credits("Mexican Population by Community Areas in Chicago, 2000",
             fontface = "bold",
             position = c("left","top"), size = 1.25)+


  tm_add_legend("line",
                labels = c("Chicago Border", "Community Area Border"),
                col = c("#343a40","#6c757d"),
                lwd = c(2.5, 2),
                title = "Border Types",
                size = 0.8) +

  # tm_credits("1 dot = 20 People",
  #            position = c(0.03, 0.32), size = 0.8, fontface = "bold") +
  tm_compass(position = c("right", "bottom"), size = 1) +
  tm_scale_bar(position = c("right", "bottom"), width = 0.15)

comm_mexican_pop_map_2000

```


```{r}

homeownership = usa_data %>% 
  filter( city == 1190, pernum == 1) %>%  
  
  
  as_survey_design(weights = hhwt) %>% 
  
  survey_count(year, race_ethnicity, ownershp_f, name="count") %>% 
  
  filter(!is.na(race_ethnicity)) %>% 
  
  group_by(year, race_ethnicity) %>% 
  mutate(total = sum(count),
         per = 100*round(count/total, 4),
         per2 = paste0(per,"%"))
  
 # #### Export 
#  
  
  
```


## 1.B: Population Pyramid and Age

### 1. Population Pyramid and Percentage of Mexican population, 2018-2022:

```{r}
# analysis

# filter for year 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)

# 1B: Population pyramid of the percentage of Mexican population by age group compared to the rest of the population in Chicago and median age in Chicago (Data source: 2018-2022 ACS data)

df <-  data_chi_2018_22 %>%
  mutate(
    age_group = case_when(
      age < 20 ~ "Young (0-19)",
      age >= 20 & age < 40 ~ "Young Adults (20-39)",
      age >= 40 & age < 60 ~ "Middle-aged (40-59)",
      age >= 60 & age < 80 ~ "Older Adults (60-79)",
      age >= 80 ~ "Elderly (80+)"
    )
  )
# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)


# Calculate weighted population counts by age group and ethnicity

# Calculate weighted population by ethnicity and age group
pop_pyramid <- df %>%
  as_survey_design(weights = perwt) %>% 
  survey_count(race_ethnicity, age_group, name="total_weighted_count")  %>%
  group_by(race_ethnicity) %>%
  mutate(total_weighted_percentage = total_weighted_count / sum(total_weighted_count) * 100,
         total_weighted_percentage  = paste0(round(total_weighted_percentage, 1), "%")) 



# get median age

weighted_age  <- df %>%
  group_by(race_ethnicity) %>%
  summarize(
    weighted_age = matrixStats::weightedMedian(age, perwt)
  ) 



# combined

df_final  <- pop_pyramid %>% left_join(weighted_age)

```








```{r}
# visual

table_1b.1 <- df_final  %>% pivot_wider(id_cols = c("race_ethnicity"), names_from = age_group, values_from = c("total_weighted_count", "total_weighted_percentage")) %>% select(race_ethnicity, 
                                                                                                                                                'total_weighted_count_Young (0-19)', 'total_weighted_percentage_Young (0-19)',
                                                                                                                                                 'total_weighted_count_Young Adults (20-39)', 'total_weighted_percentage_Young Adults (20-39)',
                                                                                                                                                 'total_weighted_count_Middle-aged (40-59)', 'total_weighted_percentage_Middle-aged (40-59)',
                                                                                                                                                 'total_weighted_count_Older Adults (60-79)', 'total_weighted_percentage_Older Adults (60-79)',
                                                                                                                                                 'total_weighted_count_Elderly (80+)', 'total_weighted_percentage_Elderly (80+)')   %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("", "Number", "Percent" , "Number", "Percent", "Number", "Percent", "Number", "Percent", "Number", "Percent"),
      align = c("l",rep("c",10)),
      caption = "Population Pyramid of the Mexican Population, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:11, width = "2.58em") %>%  

 add_header_above( align = c("l", rep("c",2)), c("Race/Ethnicity", "Young\n(0-19)" = 2, "Young Adults\n(0-19)" = 2, "Middle Aged\n(40-59)" = 2, "Older Adults\n(60-79)" = 2, "Elderly\n(80+)" = 2),
                    # background = "#cbc9e2", 
                    bold = T, font_size =8) %>% 

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   


table_1b.1
```


### 2. Median Age of Mexican population, 2018-2022:

```{r}
# analysis

# filter for year 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)

# 1B: Population pyramid of the percentage of Mexican population by age group compared to the rest of the population in Chicago and median age in Chicago (Data source: 2018-2022 ACS data)

df <-  data_chi_2018_22 %>%
  mutate(
    age_group = case_when(
      age < 20 ~ "Young (0-19)",
      age >= 20 & age < 40 ~ "Young Adults (20-39)",
      age >= 40 & age < 60 ~ "Middle-aged (40-59)",
      age >= 60 & age < 80 ~ "Older Adults (60-79)",
      age >= 80 ~ "Elderly (80+)"
    )
  )
# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)


# Calculate weighted population counts by age group and ethnicity

# Calculate weighted population by ethnicity and age group
pop_pyramid <- df %>%
  as_survey_design(weights = perwt) %>% 
  survey_count(race_ethnicity, age_group, name="total_weighted_count")  %>%
  group_by(race_ethnicity) %>%
  mutate(total_weighted_percentage = total_weighted_count / sum(total_weighted_count) * 100,
         total_weighted_percentage  = paste0(round(total_weighted_percentage, 1), "%")) 



# get median age

weighted_age  <- df %>%
  group_by(race_ethnicity) %>%
  summarize(
    weighted_age = matrixStats::weightedMedian(age, perwt)
  ) 



# combined

df_final  <- pop_pyramid %>% left_join(weighted_age)
```


```{r}
# visuals
# visuals

table_1b.2 <- weighted_age  %>% select(race_ethnicity, weighted_age) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
       col.names = c("Race/Ethnicity", "Number"),
      align = c("l",rep("c",1)),
      caption = "Median Age of the Mexican Population, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2, width = "25.8em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   

table_1b.2


```



## 1.C: Average Family Size of the Mexican Population, 2018-2022:

```{r}
# analysis

# filter for year 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)


# Filter for the head of household 
data_head_of_household <- data_chi_2018_22 %>%
  filter(pernum == 1)

# 1C Average family size of the Mexican population compared to, other Latinos, Black and White populations in Chicago

# Calculate average family size

# Define the survey design using srvyr with only weights
survey_design <- data_head_of_household %>%
  as_survey_design(weights = hhwt)

# Calculate the weighted average family size by group
avg_family_size <- survey_design %>%
  group_by(race_ethnicity) %>%
  summarize(
        avg_family_size = survey_mean(famsize, vartype = "se", na.rm = TRUE)
  ) %>%  mutate(
    avg_family_size = round(avg_family_size)
  )




```

```{r}
# visuals

table_1c <- avg_family_size %>% select(race_ethnicity, avg_family_size) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
       col.names = c("Race/Ethnicity", "Number"),
      align = c("l",rep("c",1)),
      caption = "Average Family Size of the Mexican Population, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2, width = "25.8em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   

table_1c


```

## 1.D: Income Levels and Poverty

### 1. Income Levels, 2018-2022:

```{r}
# analysis
# filter for 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)

# 1D Income Levels and poverty rates for Mexicans, other Latinos, Black and White Populations in Chicago.

# Filter the data to remove negative and extreme values
filtered_data <- data_chi_2018_22 %>%
  filter(hhincome >= 0 & hhincome < 9999999, pernum == 1)
  
# Define the survey design using srvyr with only weights
survey_design <- filtered_data  %>%
  as_survey_design(weights = hhwt)

# Calculate the weighted average household income by group
weighted_avg_income <- survey_design %>% filter(pernum == 1) %>%
  group_by(race_ethnicity) %>%
  summarize(
    avg_hhincome = survey_mean(hhincome, vartype = "se", na.rm = TRUE)
  )


# weighted median: 

weighted_median_income <- data_chi_2018_22 %>%  filter(pernum == 1) %>%
  group_by(race_ethnicity) %>%
  summarize(
    weighted_median_hhincome = matrixStats::weightedMedian(hhincome,  weights = hhwt)
  ) %>%
  ungroup()


# final df with income levels
income_levels <- weighted_avg_income %>% left_join(weighted_median_income)  %>%
  mutate(
    avg_hhincome = paste0("$", format(avg_hhincome, big.mark = ",", scientific = FALSE)),
    weighted_median_hhincome = paste0("$", format(weighted_median_hhincome, big.mark = ",", scientific = FALSE))
  )


```


```{r}
# visuals
# visual



table_1d.1 <- income_levels %>% select(race_ethnicity, avg_hhincome, weighted_median_hhincome) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Average Household Income", "Median Household Income"),
      align = c("l",rep("c",2)),
      caption = "Average and Weighted Median Income of Mexican Population, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   

table_1d.1
```


### 2. Poverty Rates, 2018-2022:

```{r}
# analysis

# poverty


# Define the survey design using srvyr with only weights
survey_design2 <- data_chi_2018_22  %>%
  as_survey_design(weights = perwt)
  

  
# Calculate poverty rates by group

# Calculate total weighted count by race_category and ethnicity
total_weighted_pop_count <- survey_design2 %>%
  group_by(race_ethnicity) %>%
  summarise(total_weighted_pop_count = survey_total())

# Calculate total weighted count in poverty by race_category and ethnicity
total_weighted_poverty_count <- survey_design2 %>%
  filter(poverty <= 1) %>%
  group_by(race_ethnicity) %>%
  summarise(total_weighted_poverty_count = survey_total())

# Combine total counts and poverty counts
combined <- left_join(total_weighted_pop_count, total_weighted_poverty_count)


poverty_rate <- combined %>% mutate(poverty_rate = (total_weighted_poverty_count/total_weighted_pop_count)*100,
                                    poverty_rate   = paste0(round(poverty_rate, 1), "%")) 

```


```{r}
# visuals


table_1d.2 <- poverty_rate  %>% select(race_ethnicity, total_weighted_poverty_count, poverty_rate) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number", "Percent"),
      align = c("l",rep("c",2)),
      caption = "Mexican Population at or Below 1% Poverty Threshold, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   

table_1d.2
```

## 1.E: Population by Mexican indigenous groups, 2018-2022:

```{r}
# analysis

# filter for 2022

data_chi_2018_22 <- usa_data %>% filter(year == 2022)


# check ipums labels

# Ensure data_chi_2018_22 is loaded

# make tribe a character
data_chi_2018_22$tribe_f <- as.character(data_chi_2018_22$tribe_f)


# create native groups by collapsing categories

data_recoded <- data_chi_2018_22  %>% mutate(
    tribe_label_collapsed = case_when(
      tribe_f %in% c("Not applicable or blank", "American Indian, tribe not specified", "American Indian and Alaska Native, not specified",
                         "American Indian and Alaska Native, tribe not elsewhere classified", 
                         "American Indian, tribe not elsewhere classified", "All other specified American Indian tribe combinations") ~ "Unspecified/Other",
      TRUE ~ tribe_f
    )
  )
 

# Population by Mexican indigenous groups by Public Use Microdata Areas (PUMAs) in Chicago.  

# Filter data for Mexican who identify as AIAN and select relevant columns
mexican_indigenous <- data_recoded %>%
  filter(hispan == 1 & race == 3) %>%
  select(puma, tribe_label_collapsed, perwt)

# Create survey design with weights
survey_design <- mexican_indigenous %>%
  as_survey_design(weights = perwt)

# Count the number of individuals by indigenous tribe within each PUMA
tribe_count <- survey_design %>%
  survey_count(tribe_label_collapsed, name = "total_weighted_count")

# round 
tribe_count  <- tribe_count %>% mutate(
  total_weighted_count_se = round(total_weighted_count_se, 1)
)
```


```{r}
# visuals


table_1e <- tribe_count %>% select(tribe_label_collapsed, total_weighted_count) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number"),
      align = c("l",rep("c",1)),
      caption = "Population by Mexican indigenous groups, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2, width = "25.8em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   

table_1e
```


# Education

## 2.A: Number and percentages of Mexican population in public and private schools. 

```{r}
# analysis
# select variables
data_chi_2018_22 <- usa_data  %>% filter(year == 2022) %>% select(year:gqtyped, perwt, hispan, race, race_ethnicity, school, schltype)



# Number and Percentage of Mexican population in public and private schools compared to other groups

# Step 1: Filter for individuals enrolled in school
enrolled_in_school <- data_chi_2018_22 %>%
  filter(school == 2)


# Define the survey design
survey_design <- enrolled_in_school %>% 
  as_survey_design(weights = perwt)

# Step 3: Calculate the total weighted count for each race/ethnicity group (denominator)
total_counts <- survey_design %>%
  group_by(race_ethnicity) %>%
  summarize(total_weighted_count = survey_total(vartype = "se"))

# Step 4: Calculate the weighted count of race/ethnicity in public and private schools (numerator)
school_counts <- survey_design %>%
  group_by(schltype, race_ethnicity) %>%
  summarize(weighted_count = survey_total(vartype = "se"))
  
# Step 5: Calculate the percentage of race/ethnicity groups in public and private schools
school_percentages <- school_counts %>%
  left_join(total_counts, by = "race_ethnicity") %>%
    mutate(percentage = paste0(round((weighted_count / total_weighted_count) * 100, 2), "%"))

# Step 6: Recode schltype for clarity
df_all <- school_percentages %>%
  mutate(schltype = case_when(
    schltype == 2 ~ "Public",
    schltype == 3 ~ "Private",
    TRUE ~ NA_character_
  ))
```


```{r}
# visual


table_2a <- df_all %>% 
  
pivot_wider(id_cols =  race_ethnicity, names_from = "schltype",  values_from = c(weighted_count, total_weighted_count, percentage)) %>% 
  select(race_ethnicity, weighted_count_Public, percentage_Public, weighted_count_Private, percentage_Private) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("", "Number", "Percent" , "Number", "Percent"),
      align = c("l",rep("c",4)),
      caption = "Number and percentages of Mexican population in public and private schools in Chicago, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "6.45em")   %>%
  add_header_above( align = c("l", rep("c",2)), c("Race/Ethnicity", "Public Schools" = 2, "Private Schools" = 2),
                    # background = "#cbc9e2", 
                    bold = T, font_size =8) %>% 

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   

table_2a 
```




## 2.C: Percentage of Limited English Proficient


```{r}

# analysis

data_chi_2018_22 <- usa_data %>%  filter(year == 2022) %>% select(year:gqtyped, perwt, hispan, race, race_ethnicity, school, speakeng)

# Percentage of Limited English Proficient. 

# Step 1: Define Survey Design
survey_design <- data_chi_2018_22 %>%
  as_survey_design(weights = perwt)

# Step 2: Group by race_ethnicity and calculate weighted counts and totals
lep_rate_by_ethnicity <- survey_design %>%
  group_by(race_ethnicity) %>%
  summarise(
    lep_count = survey_total(speakeng %in% c(1, 6)),
    total_count = survey_total(),
    lep_percentage = 100 * survey_mean(speakeng %in% c(1, 6), vartype = "se")
  ) %>%
  mutate(lep_percentage = paste0(round(lep_percentage, 1), "%"))


```


```{r}
# visual



table_2c <- lep_rate_by_ethnicity %>% select(race_ethnicity, lep_count, lep_percentage) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number", "Percent"),
      align = c("l",rep("c",2)),
      caption = "Number and percentages of Mexican population with limited English proficiency, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   

table_2c

```

## 2.D: Number and Percentages of Mexican non-citizens enrolled in public schools

```{r}
# analysis


# Step 1: Filter for individuals enrolled in school
enrolled_in_school <- usa_data  %>%  filter(year == 2022) %>%
  filter(school == 2)


# Step 2: Filter for non-citizen individuals enrolled in public schools
enrolled_in_public_schools <- enrolled_in_school %>%
  filter(schltype == 2 & citizen == 3)
  
# Step 3: Create survey design
  
survey_design_school <- enrolled_in_school %>%
  as_survey_design(weights = perwt)
  
survey_design_public_schools <- enrolled_in_public_schools %>%
  as_survey_design(weights = perwt)


# Step 4: Calculate denominator (total count enrolled in school) for each subgroup
denominator_counts <- survey_design_school %>%
  group_by(race_ethnicity) %>%
  summarise(num_enrolled_all_school = survey_total()) 
  
  
  
# Step 5: Calculate numerator (count of public school enrollment) for each subgroup
numerator_counts <- survey_design_public_schools %>%
  group_by(race_ethnicity) %>%
  summarise(num_enrolled_public = survey_total())
  
  
# Step 6: Calculate percentages and estimate SEs

percentages <- numerator_counts %>%
  left_join(denominator_counts, by = "race_ethnicity") %>%
  mutate(percentage = 100 * num_enrolled_public / num_enrolled_all_school) %>%
  mutate(percentage = paste0(round(percentage, 1), "%"))

  


```


```{r}
# visual

table_2d <- percentages %>% select(race_ethnicity, num_enrolled_public, percentage) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number", "Percent"),
      align = c("l",rep("c",2)),
      caption = "Number and percentages of Mexican non-citizens enrolled in public schools, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  
row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%
  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   

table_2d

```


## 2.E: Number and Percentages of Mexican 21-24 years old without a high school diploma


```{r}


df <- usa_data  %>%  filter(year == 2022) %>% select(race_ethnicity, schltype, school, gradeatt, perwt, educ, age) 


# Create a survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)


# method: Look at 21-24 year olds without a high school diploma out of all 21-24 year olds


# Calculate numerator: weighted count of students aged 21-24 not in school and without a high school diploma
numerator_not_in_school <- survey_design %>%
  filter(age >= 21, age <= 24, school == 1, educ %in% c("0", "1", "2", "3", "4", "5")) %>%
  group_by(race_ethnicity) %>%
  summarise(not_in_school = survey_total(vartype = "se"))

# Calculate denominator: weighted count of total students aged 21-24
denominator_total_students <- survey_design %>%
  filter(age >= 21, age <= 24) %>%
  group_by(race_ethnicity) %>%
  summarise(total_students = survey_total(vartype = "se"))

  
# Calculate dropout rate by joining numerator and denominator
dropout_data <- numerator_not_in_school %>%
  inner_join(denominator_total_students, by = "race_ethnicity") %>%
  mutate(dropout_rate = (not_in_school / total_students) * 100) %>%
  mutate(dropout_rate = paste0(round(dropout_rate, 1), "%"))

```


```{r}
# visual


table_2e <- dropout_data %>% select(race_ethnicity, not_in_school, dropout_rate) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number", "Percent"),
      align = c("l",rep("c",2)),
      caption = "Number and percentages of Mexican people aged 21-24 years old without a high school diploma, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  
  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   

table_2e

```




## 2.F: Number and Percentage of Mexican students attending undergraduate/graduate colleges or professional schools



```{r}

## select variables
df <- usa_data  %>%  filter(year == 2022) %>% select(race_ethnicity, school, gradeatt, perwt, educ, age)


# Create survey design object
survey_design <-df %>%
  as_survey_design(weights = perwt)

# Number and percentage of Mexican students attending area colleges

# 18-24 year olds


# Group by race_ethnicity, filter for enrolled in college/grad school and ages 18-24
numerator_18_24 <- survey_design %>%  filter(school == 2, gradeatt %in% c(6, 7), age >= 18, age <= 24) %>% 
  group_by(race_ethnicity) %>%
  summarise(college_attendees = survey_total(vartype = "se"))


# denominator calculate total number of 18-24 year olds
denominator_18_24 <- survey_design %>% filter(age >= 18, age <= 24) %>%
  group_by(race_ethnicity) %>%
  summarise(total_students = survey_total(vartype = "se"))


# Calculate percentage of students attending college within each racial and ethnic group
college_attendance_18_24 <- left_join(numerator_18_24, denominator_18_24, by = "race_ethnicity") %>%
  mutate(percentage_college_students = (college_attendees / total_students) * 100) %>% mutate(age = "18_24", percentage_college_students = paste0(round(percentage_college_students, 1), "%"))



# 25-34 year olds

# Group by race_ethnicity, filter for enrolled in college/grad school and ages 18-24
numerator_25_34 <- survey_design %>%  filter(school == 2, gradeatt %in% c(6, 7), age >= 25, age <= 34) %>% 
  group_by(race_ethnicity) %>%
  summarise(college_attendees = survey_total(vartype = "se"))


# denominator calculate total number of 18-24 year olds
denominator_25_34 <- survey_design %>% filter(age >= 25, age <= 34) %>%
  group_by(race_ethnicity) %>%
  summarise(total_students = survey_total(vartype = "se"))


# Calculate percentage of students attending college within each racial and ethnic group
college_attendance_25_34 <- left_join(numerator_25_34, denominator_25_34, by = "race_ethnicity") %>%
  mutate(percentage_college_students = (college_attendees / total_students) * 100) %>% mutate(age = "25_34", percentage_college_students = paste0(round(percentage_college_students, 1), "%"))


# combine

df_college <- rbind(college_attendance_18_24, college_attendance_25_34)

```



```{r}
table_2f <- df_college %>% pivot_wider(id_cols =  c(race_ethnicity), names_from = "age", values_from = c(college_attendees, percentage_college_students)) %>% select(race_ethnicity,  college_attendees_18_24,  percentage_college_students_18_24,  college_attendees_25_34,  percentage_college_students_25_34) %>%
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("", "Number", "Percent" , "Number", "Percent"),
      align = c("l",rep("c",4)),
      caption = "Number and Percentage of Mexican students attending undergraduate/graduate colleges or professional schools, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em")  %>%
  column_spec(2:5, width = "6.45em")  %>%
  add_header_above( align = c("l", rep("c",2)), c("Race/Ethnicity", "Ages 18-24" = 2, "Ages 25-34" = 2),
                    # background = "#cbc9e2", 
                    bold = T, font_size =8) %>% 

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   


table_2f

```



## 2.G: Educational Attainment of Mexicans 25 and Older



```{r}
df <- usa_data  %>%  filter(year == 2022) %>% select(race_ethnicity, educ, age, perwt)




# collapse educational categories
df <- df %>%   mutate(educ = as.character(educ)) %>% mutate(education_level = case_when(
    educ %in% c("0", "1", "2") ~ "Grade School or Less",
    educ %in% c("3", "4", "5") ~ "Some High School",
    educ %in% c("6") ~ "High School",
    educ %in% c("7") ~ "Some College",
     educ %in% c("8", "9", "10", "11") ~ "2 years of college or higher",
    TRUE ~ "Other"
  ))


# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)

# Calculate numerator: weighted count of individuals by education level and race_ethnicity
numerator <- survey_design %>%
  filter(age >= 25) %>%
  group_by(race_ethnicity, education_level) %>%
  summarise(weighted_n = survey_total(vartype = "se"))

# Calculate denominator: total population by race_ethnicity
denominator <- survey_design %>%
  filter(age >= 25) %>%
  group_by(race_ethnicity) %>%
  summarise(total_pop = survey_total(vartype = "se"))

# Merge numerator and denominator by race_ethnicity
education_levels <- c("Grade School or Less", "Some High School", "High School","Some College", "2 years of college or higher")

education_rates <- numerator %>%
  left_join(denominator, by = "race_ethnicity") %>%
  mutate(educational_rate = (weighted_n / total_pop) * 100,
  education_level = factor(education_level, levels = education_levels))    %>%  arrange(race_ethnicity, education_level)  %>% mutate(educational_rate = paste0(round(educational_rate, 1), "%"))
  
```


```{r}
table_2g <- education_rates %>% select(race_ethnicity, education_level, weighted_n, educational_rate)  %>% pivot_wider(id_cols =  c(race_ethnicity), names_from = "education_level", values_from = c(weighted_n, educational_rate))   %>% select(race_ethnicity, 'weighted_n_Grade School or Less', 'educational_rate_Grade School or Less',  'weighted_n_Some High School', 'educational_rate_Some High School', 'weighted_n_High School', 'educational_rate_High School', 'weighted_n_Some College', 'educational_rate_Some College', 'weighted_n_2 years of college or higher', 'educational_rate_2 years of college or higher') %>% 


 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("",  "Number", "Percent", "Number", "Percent" , "Number", "Percent" , "Number", "Percent" , "Number", "Percent" ),
      align = c("l",rep("c",11)),
      caption = "Educational Attainment of Mexicans 25 and Older, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em")  %>%
  column_spec(2:11, width = "2.58em")  %>%
  add_header_above( align = c("l", rep("c",2)), c("Race/Ethnicity",  "Grade School\nor Less" = 2, "Some High School" = 2, "High School" = 2, "Some College" = 2, "2 years or more\nof college" = 2),
                    # background = "#cbc9e2", 
                    bold = T, font_size =8) %>% 

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "") 


table_2g

```





# Housing & Community Development

## 3.A: Rates of home ownership and rental housing among Mexicans with historical trends and comparisons to other groups.

```{r}
# homeownership = read.csv("Data Tables/3.A.homeownership.csv" )

table_a3 <-  homeownership %>% 
  
  filter(ownershp_f == "Owned or being bought (loan)") %>% 

  pivot_wider(id_cols =  c(race_ethnicity, ownershp_f), names_from = "year", values_from = c(count, per, per2)) %>% 

  mutate(diff2000_2022 = count_2022 - count_2000,
         diff_per = round(per_2022 - per_2000, 4), 
         diff_per2 = paste0(diff_per,"%")
         ) %>% 
  select(race_ethnicity, count_2000, per2_2000, count_2012, per2_2012, count_2022, per2_2022, diff2000_2022, diff_per2) %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("", "Number", "Percent" , "Number", "Percent", "Number", "Percent", "Number", "Percent"),
      align = c("l",rep("c",8)),
      caption = "Homeownership by Race/Ethnicity in Chicago, 2000, 2008-2012, and 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:9, width = "3.8em") %>%
  add_header_above( align = c("l", rep("c",4)), c("Race/Ethnicity", "2000" = 2, "2008-2012" = 2, "2018-2022" = 2, "Change from\n2000 to 2018-2022" = 2),
                    # background = "#cbc9e2", 
                    bold = T, font_size =8) %>% 
  # row_spec(0, bold = T, 
  #          # background = "#f2f0f7",
  #          font_size = 8) %>% 
  
  footnote(general = "Source: IPUMS-USA database (2000, 2008-2012, 2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   

table_a3

```


## 3.B: Number and percentage of Mexicans who, live in crowded housing, or are rent-burdened.

```{r}


## Be careful rooms == 0 is NA but it is overcrowded
overcrowded = usa_data %>%
  filter( city == 1190, pernum == 1, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?


  as_survey_design(weights = hhwt) %>%
  survey_count(year, race_ethnicity, rooms, numprec, name="count") %>%
  mutate(p_per_room = numprec/rooms,
         overcrowded = if_else(p_per_room >1, "Overcrowded", "Not Overcowded")) %>%

  group_by(year, race_ethnicity) %>%
  mutate(total = sum(count, na.rm = T)) %>%
  filter(rooms != 0 ) %>%  ## rooms == 0 is NA but it is overcrowded
  ungroup() %>%
  group_by(year, race_ethnicity, overcrowded, total) %>%

  summarise(count_sum = sum(count)) %>%
  mutate(per = 100*round(count_sum/total, 4),
         per2 = paste0(per,"%")) %>%
  ungroup()

```


### 1. Crowded Housing

```{r}

table_3.b1 <- overcrowded %>%

  filter(overcrowded == "Overcrowded" ) %>%

  select(race_ethnicity, count_sum , per2) %>%

   # using kable

  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number", "Percent"  ),
      align = c("l",rep("c",2)),
      caption = "Number and Percentage of Households Living in Crowded Housing
      by Race/Ethnicity in Chicago, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = T,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%
  # add_header_above( c("Race/Ethnicity", "2000" = 2, "2008-2012" = 2, "2018-2022" = 2, "Change from\n2000 to 2018-2022" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>%

  add_footnote(c("Housing is defined as overcrowded if more than one person in the room.", "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute"),
 notation = "none")

table_3.b1

```


### 2. Rent burdened

```{r}
rent_burdened = usa_data %>%
  select(pernum, city, year, rentgrs, hhincome, ownershp_f, hhwt, race_ethnicity ) %>% 
  filter( city == 1190, pernum == 1, year == 2022) %>%  

  group_by(race_ethnicity, ownershp_f) %>%
  mutate(total = sum(hhwt, na.rm = T)) %>%
  ungroup() %>%
  filter(hhincome != 9999999 ) %>%
  filter(rentgrs!= 0) %>%

  mutate(monthly_income = hhincome/12, 
        rent_income = rentgrs/monthly_income,
         burdened_30 = if_else(rent_income >0.3 , "More than 30%","Less than 30%"),
         burdened_50 = if_else(rent_income >0.5, "More than 50%", "Less than 50%")) %>% 
  as_survey_design(weights = hhwt) %>%
  survey_count(year, race_ethnicity, total,ownershp_f, burdened_30, burdened_50 , name="count") %>%
  mutate(per = 100*round(count/total, 4),
         per2 = paste0(per,"%")) 
 

  ####### prepare data for the table

table_3.b2 = rent_burdened %>%
  filter( burdened_30 != "Less than 30%") %>%
  group_by(race_ethnicity ) %>%

  summarise(more_30_count = sum(count[burdened_30== "More than 30%"]),
            more_50_count = sum(count[burdened_50== "More than 50%"]),
            more_30_per = 100 *round(more_30_count/total, 4),
            more_30_per2 = paste0(more_30_per, "%"),
            more_50_per = 100 *round(more_50_count/total, 4),
            more_50_per2 = paste0(more_50_per, "%")) %>%
  distinct()



table_3.b2_2 <- table_3.b2 %>%
  select(race_ethnicity, more_30_count, more_30_per2, more_50_count, more_50_per2) %>%

  # using kable

  kbl(booktabs = T,
      col.names = c("", "Number", "Percent" , "Number", "Percent" ),
      align = c("l",rep("c",4)),
      caption = "Rent Burdened households by Race/Ethnicity in Chicago, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "6.45em") %>%
  add_header_above( align = c("l", rep("c",2)), c("Race/Ethnicity", "More than 30%" = 2, "More than 50%" = 2),
                    # background = "#cbc9e2",
                    bold = T, font_size =8) %>%
  # row_spec(0, bold = T,
  #          # background = "#f2f0f7",
  #          font_size = 8) %>%

  # footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute.Rent Burdened Households Defined as Spending 30% or More of Income on Rent",
  #          general_title = "") %>% 
  add_footnote(c("Rent Burdened Households Defined as Spending 30% or More of Income on Rent", "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute"),
 notation = "none") 


table_3.b2_2
```


### 3. Own Cost Burdened

```{r}

owncost_burdened = usa_data %>%
  select(pernum, city, year, owncost, hhincome, ownershp_f, hhwt, race_ethnicity ) %>% 
  filter( city == 1190, pernum == 1, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?

  group_by(race_ethnicity, ownershp_f) %>%
  mutate(total = sum(hhwt, na.rm = T)) %>%
  ungroup() %>%
  filter(hhincome != 9999999 ) %>%
  filter(owncost!= 99999) %>%

  mutate(monthly_income = hhincome/12,
         owncost_income = owncost/monthly_income,
         burdened_30 = if_else(owncost_income >0.3 , "More than 30%","Less than 30%"),
         burdened_50 = if_else(owncost_income >0.5, "More than 50%", "Less than 50%")) %>%

  as_survey_design(weights = hhwt) %>%
  survey_count(year, race_ethnicity, total,ownershp_f, burdened_30, burdened_50 , name="count") %>%
  mutate(per = 100*round(count/total, 4),
         per2 = paste0(per,"%"))



####### prepare data for the table

table_3.b3 = owncost_burdened %>%
  filter( burdened_30 != "Less than 30%") %>%
  group_by(race_ethnicity ) %>%

  summarise(more_30_count = sum(count[burdened_30== "More than 30%"]),
            more_50_count = sum(count[burdened_50== "More than 50%"]),
            more_30_per = 100 *round(more_30_count/total, 4),
            more_30_per2 = paste0(more_30_per, "%"),
            more_50_per = 100 *round(more_50_count/total, 4),
            more_50_per2 = paste0(more_50_per, "%")) %>%
  distinct()



table_3.b3_2 <- table_3.b3 %>%
  select(race_ethnicity, more_30_count, more_30_per2, more_50_count, more_50_per2) %>%

  # using kable

  kbl(booktabs = T,
      col.names = c("", "Number", "Percent" , "Number", "Percent" ),
      align = c("l",rep("c",4)),
      caption = "Cost-Burdened Homeowners by Race and Ethnicity in Chicago (2018-2022)",
      format.args = list(big.mark = ","),
      linesep = "", table.attr = "style='width:30%;'") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "6.45em") %>% 
  add_header_above(align = c("l", rep("c",2)), c("Race/Ethnicity", "More than 30%" = 2, "More than 50%" = 2),
                    # background = "#cbc9e2",
                    bold = T, font_size =8) %>%
  # row_spec(0, bold = T,
  #          # background = "#f2f0f7",
  #          font_size = 8) %>%

  
  add_footnote(c("Own Cost Burdened Households Defined as ?", "Source: IPUMS-USA database (2018-2022).Tabulations by Great Cities Institute"),
 notation = "none")

table_3.b3_2
```


## 3.D: House Value

```{r}

community_shp <- read_sf("C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/Boundaries - Community Areas (current)/geo_export_2081dd0e-84f5-45c8-b501-f07b67ad9491.shp") %>% 
  st_transform( crs = 4326) %>% 
  select(community,geometry)


chicago_shp <- read_sf("C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/shp files/selected_centroid_2020.shp") %>% 
  select(GEOID, geometry) %>% 
  st_transform( crs = 4326)


census_centroids_2022 <- st_centroid(chicago_shp)

# Spatial join: attach community area names to census tracts based on centroid location
census_community_2022 <- st_join(census_centroids_2022, 
                                 community_shp, 
                                 join = st_within) %>% 
  mutate(lable = str_to_title(community)) %>% 
  mutate(lable = gsub(" ", "\n", lable))

# write_sf(census_community_2022, "C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/shp files/census_community_2022.shp")

# search_variables <- load_variables(2022,
#                          dataset = "acs5",
#                          cache = TRUE)
# 
# view(search_variables)

# options(tigris_use_cache = TRUE)

house_value_data <-  get_acs(
  cache_table = TRUE,
  geography = "tract",
  variables = c("Not Hispanic or Latino"="B03001_002",
                "Hispanic or Latino"="B03001_003",
                "Mexican" = "B03001_004", 
                "total_pop"= "B03001_001"),
                
  summary_var =  "B25077_001",              
  state = 17,
  year = 2022) %>% 
  rename("Mvalue" = "summary_est") %>% 

  filter(GEOID %in% c(chicago_shp$GEOID)) %>% 
  left_join(census_community_2022 %>% st_drop_geometry(), by = "GEOID") %>% 
  pivot_wider(id_cols = c(GEOID , NAME, Mvalue, community ), names_from = variable, values_from = estimate) %>% 

  group_by(community) %>% 
  summarise(total_his = sum(`Hispanic or Latino`, na.rm = T),
         total_mexican = sum(Mexican, na.rm = T),
         total_pop = sum(total_pop, na.rm = T),
         prc_mexican = total_mexican/total_pop, 
         Mvalue = median(Mvalue, na.rm = T)) 

comunity_house_value_shp <- left_join(community_shp, house_value_data,  by = join_by(community == community)) %>% 
  mutate(lable = str_to_title(community)) %>% 
  mutate(lable = gsub(" ", "\n", lable))


```



```{r fig.height=11, fig.width=8}
value_map_2020 <-

  tm_shape(comunity_house_value_shp, bbox = bbox_new) +
  tm_polygons(col = "Mvalue",
              style = "quantile",
              border.col = "white",
              palette="Blues",
              alpha = 0.7,
              n = 5,
              title = "Median House Values",
              lwd = 1) +

  tm_shape(comunity_house_value_shp %>% filter(prc_mexican>=0.5)) +
  tm_borders(col = "#9b2226",
             alpha = 1,
             lwd = 2)+
  tm_text("lable" , size = 0.7, alpha = 0.7, fontface="bold", fontfamily = "sans") +

  tm_shape(Chicago_boundary_shp) +
  tm_borders(col = "#343a40",
             alpha = 1,
             lwd = 2.5) +

  tm_layout( frame = TRUE,
             # main.title.position = c("left", "top"),
             # main.title = "Density of Hispanic or Latino Population by their Origins in Chicago, 2010",
             # main.title.size = 1.2,
            legend.outside = FALSE,
            legend.width = 1.5,
            legend.title.size = .9,
            legend.text.size = 0.8,
            legend.title.fontface = "bold",
            # legend.position = c("left", "bottom"),
            legend.position = c(0.03, 0.12),
            legend.bg.color = "white",
            legend.bg.alpha = .3) +       # manually position legend) +
  
  tm_credits("Data source:\n5-year ACS (2018-2022)\nMap Generated by:The Great Cities Institute",
             fontface = "bold",
             position = c("left","bottom"), size = 0.8)+
  
  
  tm_credits("Median House Value in Chicago, 2018-2022",
             fontface = "bold",
             position = c("left","top"), size = 1.25)+

  
  tm_add_legend("line",
                labels = c("Chicago Border","" ,"Community Areas with More Than\n50% Share of Mexican Population", ""),
                col = c("#343a40","white" ,"#9b2226", "white"),
                lwd = c(2.5, 2),
                title = "Border Types",
                size = 0.8) +

  # tm_credits("1 dot = 20 People",
  #            position = c(0.03, 0.32), size = 0.8, fontface = "bold") +
  tm_compass(position = c("right", "bottom"), size = 1) +
  tm_scale_bar(position = c("right", "bottom"), width = 0.15)

value_map_2020

```

### Graph

```{r fig.width=6, fig.height=5 ,fig.align='center'}
####### Barchart: "op Ten Counties in Illinois with the Highest Population in Labor Force
Mhouse_value_plot <- comunity_house_value_shp %>% 
  filter(prc_mexican>=0.5) %>% 
  arrange(desc(Mvalue)) %>% 
  mutate(community2 = str_to_title(community)) %>% 
  mutate(community2 = factor(community2),
         community2 = fct_reorder(community2, Mvalue)) %>% 
  ggplot(aes(y = community2, x = Mvalue) ) +
  geom_bar( stat = "identity", 
            fill = "#6baed6",
            color = "#9b2226",
           width=0.7,
           linewidth = 1) +
  labs(y="", x = "",
       title="Median House Value in Chicago Community Areas with More Than\n50% Share of Mexican Population(2018-2022)",
       caption = "Source: American Community Survey 5-year Estimate (2018-2022) \nTabulations by Great Cities Institute",
       fill="") +
  geom_text(aes(label = paste0(scales::comma(Mvalue))),
            position=position_dodge(width=0.9), hjust = -0.3, vjust = 0.8,
            size = 3) +
  scale_y_discrete(labels = function(x) str_wrap(x, width = 24))+ # Modify labels of ggplot2 barplot
  scale_x_continuous(labels = function(x) paste0(x, "%"), limits = c(0,550000), breaks = seq(0, 550000, by = 550000))+
  theme_plots +
  theme(
        axis.text.y = element_text(size = 8),
        panel.background = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size = 7, color = "#636363"),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.line.x = element_blank())
  
  
  Mhouse_value_plot
```

# Health 

## 4.A: Number and Percentages of Mexicans without Health Insurance, 2018-2022:


```{r}
# analysis
## select variables, filter for year 2022
df <- usa_data %>%  filter(year == 2022) %>% select(race_ethnicity, hcovany, perwt)

# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)
  
  
# Calculate numerator: weighted count of individuals without health insurance by race_ethnicity
numerator_without_insurance <- survey_design %>%
  filter(hcovany == 1) %>% 
  group_by(race_ethnicity) %>%
  summarise(weighted_n_without_insurance = survey_total(vartype = "se"))

# Calculate denominator: total population by race_ethnicity
denominator <- survey_design %>%
  group_by(race_ethnicity) %>%
  summarise(total_pop = survey_total(vartype = "se")) 
  
# calculate rate

# Join numerator and denominator
df_rates <- numerator_without_insurance %>%
  left_join(denominator, by = "race_ethnicity") %>%
  mutate(no_insurance_rate = (weighted_n_without_insurance / total_pop) * 100, 
         no_insurance_rate = paste0(round(no_insurance_rate, 1), "%"))
  


```


```{r}
# visuals
# visuals
table_1a <- df_rates  %>% select(race_ethnicity, weighted_n_without_insurance, no_insurance_rate)   %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number without health insurance", "No Insurance Rate"),
      align = c("l",rep("c",2)),
      caption = "No Health Insurance, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "12.9em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "")   


table_1a 

```




# Employment and Business

## Mexican Employment
### 1. Unemployment Rate

```{r}

unemployment = usa_data %>% 
  filter( city == 1190, year == 2022) %>%  
  
  as_survey_design(weights = perwt) %>% 
  survey_count(race_ethnicity, employment=empstat_f, laborforce= labforce_f,  name="count") %>% 
  filter(laborforce == "Yes, in the labor force") %>% 
  group_by(race_ethnicity) %>% 
  mutate(in_laborforce = sum(count)) %>% 
  filter(employment == "Unemployed") %>% 
  mutate(unemployment_rate = 100*round(count/in_laborforce, 4),
         prc2 = paste0(unemployment_rate, "%"))
  

table_5.a1 <- unemployment %>%
  select(race_ethnicity,  in_laborforce, count, prc2) %>%
  
  # using kable
  kbl(booktabs = T,
      col.names = c("Race/Ethnicity",  "In the Labor Force" ,"Unemployed", "Percent"),
      align = c("l",rep("c",3)),
      caption = "Unemployment Rate by Race/Ethnicity in Chicago, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "8.6em") %>% 
  # add_header_above(align = c("l", rep("c",3)), c("Race/Ethnicity", "More than 30%" = 2, "More than 50%" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>%

  
  add_footnote(c( "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute"),
 notation = "none")

table_5.a1

```

\clearpage

### 2. Numbers and percentages of Mexicans employed, by industry


```{r}


industry = usa_data %>%
  filter( city == 1190, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?

  

  mutate(industry = case_when(ind %in% c(0170:0490)  ~ "Agriculture, Forestry, Fishing, and Hunting, and Mining",
                              ind == 0770  ~ "Construction",
                              ind %in% c(1070:3990)  ~ "Manufacturing",
                              ind %in% c(4070:4590)  ~ "Wholesale Trade",
                              ind %in% c(4670:5790)  ~ "Retail Trade",
                              ind %in% c(6070:6390, 0570:0690) ~ "Transportation and Warehousing, and Utilities",
                              ind %in% c(6470:6780)  ~ "Information",
                              ind %in% c(6870:7190)  ~ "Finance and Insurance, and Real Estate, and Rental and Leasing",
                              ind %in% c(7270:7790)  ~ "Professional, Scientific, and Management, and Administrative, and Waste Management Services",
                              ind %in% c(7860:8470)  ~ "Educational Services, and Health Care and Social Assistance",
                              ind %in% c(8561:8690)  ~ "Arts, Entertainment, and Recreation, and Accommodation and Food Services",
                              ind %in% c(8770:9290)  ~ "Other Services, Except Public Administration",
                              ind %in% c(9370:9590)  ~ "Public Administration",
                              ind %in% c(9670:9870)  ~ "Military",
                              ind == 9920  ~ "Unemployed, last worked 5 years ago or earlier or never worked",

                                    TRUE ~ NA_character_)) %>%

  as_survey_design(weights = perwt) %>%
  survey_count(race_ethnicity, industry, employment=empstat_f, name="count") %>%
  filter(employment == "Employed") %>%
  group_by(race_ethnicity) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>%
  mutate(prc = 100* round(count/total_employed, 4),
         prc2 = paste0(prc,"%")) %>%
  ungroup()




table_5a.2 <-  industry %>%
  # filter(race_ethnicity %in% c("Other Latinos")) %>% 
  select(industry, count,  prc2) %>%

   # using kable
  kable(booktabs = T,
        col.names = c("Industry", "Number", "% Share of Employed"   ),
      align = c("l",rep("c",2)),
      caption = "Employment Distribution by Industry and Race/Ethnicity in Chicago, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                # latex_options = "HOLD_position",
                ) %>%
  column_spec(1, width = "40em") %>%
  column_spec(2:3, width = "7em") %>%
  # add_header_above( c("Race/Ethnicity", "2000" = 2, "2008-2012" = 2, "2018-2022" = 2, "Change from\n2000 to 2018-2022" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>%

  pack_rows("Mexican", 1,14, hline_after = T  ) %>% 
  pack_rows("Other Latinos", 15, 28, hline_before = T , hline_after = T) %>% 
  pack_rows("White (non-Hispanic or Latino)", 29, 42,  hline_before = T , hline_after = T) %>% 
  pack_rows("Black (non-Hispanic or Latino)", 43, 56,  hline_before = T, hline_after = T ) %>% 
  pack_rows("Other (non-Hispanic or Latino)", 57, 70, hline_before = T, hline_after = T) %>% 
  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "") 
  

table_5a.2

```

\clearpage

### 3. Numbers and percentages of Mexicans employed, by Occupation


```{r}

occupation = usa_data %>%
  filter( city == 1190, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?

  mutate(occupation = case_when(occ %in% c(0010:3550)  ~ "Management, Business, and Financial Occupations",
                                occ %in% c(3601:4655)  ~ "Service Occupations",
                                occ %in% c(4700:5940)  ~ "Sales and Office Occupations",
                                occ %in% c(6005:7640)  ~ "Natural Resources, Construction, and Maintenance Occupations",
                                occ %in% c(7700:9760)  ~ "Production, Transportation, and Material Moving Occupations",
                                    TRUE ~ NA_character_)) %>%

  as_survey_design(weights = perwt) %>%
  survey_count(race_ethnicity, occupation, employment=empstat_f, name="count") %>%
  filter(employment == "Employed") %>%
  group_by(race_ethnicity) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>%
  mutate(prc = 100* round(count/total_employed, 4),
         prc2 = paste0(prc,"%")) %>%
  filter(!is.na(occupation)) %>% 
  ungroup()




table_5a.3 <-  occupation %>%
  # filter(race_ethnicity %in% c("Other Latinos")) %>% 
  select(occupation, count,  prc2) %>%

   # using kable
  kable(booktabs = T,
        col.names = c("Occupations", "Number", "% Share of Employed"  ),
      align = c("l",rep("c",2)),
      caption = "Employment Distribution by Occupations Category and Race/Ethnicity in Chicago, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                # latex_options = "HOLD_position",
                ) %>%
  column_spec(1, width = "40em") %>%
  column_spec(2:3, width = "7em") %>%
  # add_header_above( c("Race/Ethnicity", "2000" = 2, "2008-2012" = 2, "2018-2022" = 2, "Change from\n2000 to 2018-2022" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>%

  pack_rows("Mexican", 1,5, hline_after = T  ) %>% 
  pack_rows("Other Latinos", 6, 10, hline_before = T , hline_after = T) %>% 
  pack_rows("White (non-Hispanic or Latino)", 11, 15,  hline_before = T , hline_after = T) %>% 
  pack_rows("Black (non-Hispanic or Latino)", 16, 20,  hline_before = T, hline_after = T ) %>% 
  pack_rows("Other (non-Hispanic or Latino)", 21, 25, hline_before = T, hline_after = T) %>% 
  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "") 
  

table_5a.3

```


## 5.B: Numbers and percentages of unionized Mexican workers

```{r}
  
options(knitr.kable.NA = '-') 




Union <- cps_ddi %>% 
  
  filter( year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?
  mutate(race_f = as_factor(race),
         hispan_f = as_factor(hispan), 
         laborforce_f = as_factor(labforce),
         employment_f = as_factor(empstat),
         union_f = as_factor(union)) %>%

  mutate(race_ethnicity = case_when(hispan ==0 & race == 100 ~ "White (non-Hispanic or Latino)",
                                    hispan ==0 & race == 200 ~ "Black (non-Hispanic or Latino)",
                                    hispan %in% c(200:612) ~ "Other Latinos",
                                    hispan %in% c(100:109) ~ "Mexican", 
                                    hispan ==0 & !race %in%c(100,200) ~ "Other (non-Hispanic or Latino)",
                                    TRUE ~ NA_character_)) %>% 
  mutate(race_ethnicity = factor(race_ethnicity, level = c("Mexican", "Other Latinos","White (non-Hispanic or Latino)", "Black (non-Hispanic or Latino)",  "Other (non-Hispanic or Latino)"))) %>% 
  
  mutate(employment_f = case_when(empstat %in% c(1, 10, 12) ~ "Employed",
                              TRUE ~ "other")) %>% 
  
  
  filter(labforce != 0 |  empstat !=0) %>% 
  filter(employment_f == "Employed") %>%
  as_survey_design(weights = asecwt) %>% 
  
  survey_count(race_ethnicity, laborforce_f, employment_f,  union_f, name="count") %>% 
  group_by(race_ethnicity) %>% 
  mutate(total_employed = sum(count)) %>% 
  filter(union_f %in% c("Member of labor union", "Covered by union but not a member")) %>% 
  # group_by(race_ethnicity, total_employed) %>% 
  mutate(prc = 100*round(count/total_employed, 4),
            prc2 = paste0(prc, "%")) %>% 
  select(race_ethnicity,union_f, total_employed,  count, prc2) %>% 
  pivot_wider(names_from = union_f, values_from = c(count, prc2))
  


table5b <- Union %>% 
  select(race_ethnicity,  total_employed,  `count_Member of labor union`, `prc2_Member of labor union`, `count_Covered by union but not a member`, `prc2_Covered by union but not a member`  ) %>% 


  kbl(booktabs = T,
      col.names = c("",  "" ,"Number", "Percent", "Number", "Percent"),
      align = c("l",rep("c",5)),
      caption = " Numbers and percentages of unionized Mexican workers by Race/Ethnicity in Illinois, 2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "5.16em") %>%
  add_header_above(align = c("l", rep("c",3)), c("Race/Ethnicity",  "Total Employed","Member of labor union" =2 ,"Covered by union but not a member" = 2),
                    # background = "#cbc9e2",
                    bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>%

  
  add_footnote(c( "Source: IPUMS-CPS database (2022). Tabulations by Great Cities Institute"),
 notation = "none")

table5b

```

# 6.Citizenship and Civic Participation

## 6.A:Number and percentages of Mexicans by status

```{r}
# Define df at the top
df <- usa_data %>% 
  select(race_ethnicity, citizen, perwt, year) %>%
  mutate(
    citizenship_status = case_when(
      citizen == 1 ~ "Citizen",
      citizen == 2 ~ "Citizen",
      citizen == 3 ~ "Not a Citizen",
      TRUE ~ "Unknown"  # This includes N/A and any other undefined categories
    )
  )


# 2022
  survey_design <- df %>%
    as_survey_design(weights = perwt)
  
  numerator_2022 <- survey_design %>%
    filter(year == 2022 & citizen == 3) %>%
    group_by(race_ethnicity, citizenship_status) %>%
    summarise(weighted_n = survey_total(vartype = "se"))
  
  denominator_2022 <- survey_design %>%
     filter(year == 2022)%>%
    group_by(race_ethnicity) %>%
    summarise(total_pop = survey_total(vartype = "se"))
  
  df_rates_2022 <- numerator_2022 %>%
    left_join(denominator_2022, by = "race_ethnicity") %>%
    mutate(rate = round(weighted_n / total_pop, 4)*100) %>% 
    mutate(year = 2022)
  
  
# 2012
  
  numerator_2012 <- survey_design %>%
    filter(year == 2012 & citizen == 3) %>%
    group_by(race_ethnicity, citizenship_status) %>%
    summarise(weighted_n = survey_total(vartype = "se"))
  
  denominator_2012 <- survey_design %>%
     filter(year == 2012)%>%
    group_by(race_ethnicity) %>%
    summarise(total_pop = survey_total(vartype = "se"))
  
  df_rates_2012 <- numerator_2012 %>%
    left_join(denominator_2012, by = "race_ethnicity") %>%
    mutate(rate = round(weighted_n / total_pop, 4)*100) %>% 
    mutate(year = 2012)
  
  
  
# 2000
   numerator_2000 <- survey_design %>%
    filter(year == 2000 & citizen == 3) %>%
    group_by(race_ethnicity, citizenship_status) %>%
    summarise(weighted_n = survey_total(vartype = "se"))
  
  denominator_2000 <- survey_design %>%
     filter(year == 2000)%>%
    group_by(race_ethnicity) %>%
    summarise(total_pop = survey_total(vartype = "se"))
  
  df_rates_2000 <- numerator_2000 %>%
    left_join(denominator_2000, by = "race_ethnicity") %>%
    mutate(rate = round(weighted_n / total_pop, 4)*100) %>% 
    mutate(year = 2000)
   
  

# Combine all years
citizenship <- bind_rows(df_rates_2022, df_rates_2012, df_rates_2000)
```



```{r}
 table_6.a1  <- citizenship %>%
  ungroup() %>%
  filter(citizenship_status != "Unknown") %>% 
  select(year, race_ethnicity,  citizenship_status, rate) %>%
  pivot_wider(id_cols = race_ethnicity,   names_from =c(citizenship_status, year) , values_from = rate) %>% 
  mutate(change_non_citizen = round(`Not a Citizen_2022` - `Not a Citizen_2000`, 2)) %>% 
    
    select(race_ethnicity, `Not a Citizen_2000` , `Not a Citizen_2012`,  `Not a Citizen_2022`,change_non_citizen) %>%
  mutate(across(-race_ethnicity, ~ paste0(., "%"))) %>% 
  
  kbl(booktabs = T,
      col.names = c("", "Non citizen Rate", "Non citizen Rate", "Non citizen Rate", "Non citizen Rate"),
      align = c("l",rep("c",4)),
      caption = "Citizenship by Race/Ethnicity in Chicago, 2000-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "6.45em") %>%
  add_header_above(align = c("l", rep("c",4)), c("Race/Ethnicity", "2000" = 1, "2008-2012" = 1, "2018-2022" = 1, "Change from\n2000 to 2018-2022" = 1), 
                   bold = T, font_size =8) %>%
  # row_spec(0, bold = T,
  #          # background = "#f2f0f7",
  #          font_size = 8) %>%

  
  add_footnote(c( "Source: IPUMS-USA database (2000-2022). Tabulations by Great Cities Institute"),
 notation = "none")              

table_6.a1                                                                                                                                                                                                                                                                                                                                                                                                      
```

