---
title: "Mexican Report"
author: "Employment and Business"
date: "Great Cities Institute"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    highlight: pygments
    extra_dependencies: ["float"]
    fig_caption: true
mainfont: "Arial"
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F, fig.pos="H")
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

library(ipumsr)
library(tidyverse)
library(purrr)
library(sf)
library(tidycensus)
library(tidyr)
library(readxl)
library(sf)
library(htmltools)
library(leaflet)
library(janitor)
library(data.table)
library(survey) # for survey analysis
library(srvyr)  # for tidy survey analysis
library(palmerpenguins)
library("kableExtra") 
library(stringr)
library(knitr)
library(palmerpenguins)
options(tinytex.verbose = TRUE)


usa_data <- read.csv("C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/final_Mexican_IL_2000_22.csv") %>% 
  mutate(race_ethnicity = case_when(hispan ==0 & race == 1 ~ "White (non-Hispanic or Latino)",
                                    hispan ==0 & race == 2 ~ "Black (non-Hispanic or Latino)",
                                    hispan %in% c(2,3,4) ~ "Other Latinos",
                                    hispan ==1 ~ "Mexican", 
                                    hispan ==0 & race %in%c(3:9) ~ "Other (non-Hispanic or Latino)",
                                    TRUE ~ NA_character_)) %>% 
  mutate(race_ethnicity = factor(race_ethnicity, level = c("Mexican", "Other Latinos","White (non-Hispanic or Latino)", "Black (non-Hispanic or Latino)",  "Other (non-Hispanic or Latino)"))) %>% 
  mutate(hispan_breakdown = case_when(
  hispand == 0 ~ "Not Hispanic",
  hispand == 100 ~ "Mexican", 
  hispand == 200 ~ "Puerto Rican",
  hispand == 300 ~ "Cuban",
  hispand == 411 ~ "Costa Rican",
  hispand == 412 ~ "Guatemalan",
  hispand == 413 ~ "Honduran",
  hispand == 414 ~ "Nicaraguan",
  hispand == 415 ~ "Panamanian",
  hispand == 416 ~ "Salvadoran",
  hispand == 417 ~ "Central American, not specified",
  hispand == 420 ~ "Argentinean",
  hispand == 421 ~ "Bolivian",
  hispand == 422 ~ "Chilean",
  hispand == 423 ~ "Colombian",
  hispand == 424 ~ "Ecuadorian",
  hispand == 425 ~ "Paraguayan",
  hispand == 426 ~ "Peruvian",
  hispand == 427 ~ "Uruguayan",
  hispand == 428 ~ "Venezuelan",
  hispand == 431 ~ "South American, not specified",
  hispand == 450 ~ "Spaniard",
  hispand == 460 ~ "Dominican",
  hispand == 498 ~ "Other, not specified",
  TRUE ~ "Other"
),
mex_foreign_born = case_when(
    race_ethnicity == "Mexican" & bpl %in% as.character(1:56) ~ "U.S. Born",
    race_ethnicity == "Mexican" & bpl == "200" ~ "Mexico Born",
    TRUE ~ NA_character_  # Optional: default case for other values or groups
  )) %>% ## filter for hispanic groups here ##
  mutate(mex_foreign_born = factor(mex_foreign_born, level = c("Mexico Born","U.S. Born"))) %>% 
  
  mutate(hispan_breakdown = factor(hispan_breakdown, level = c("Mexican", "Puerto Rican","Ecuadorian", "Cuban", "Guatemalan", "Venezuelan", "Other, not specified", "Colombian", "Salvadoran", "Peruvian", "Honduran", "Spaniard", "Dominican", "Argentinean", "Chilean", "Nicaraguan", "Panamanian", "Costa Rican", "Bolivian", "Uruguayan", "Paraguayan", "South American, not specified", "Central American, not specified", "Not Hispanic"))
  )
  
# install.packages("knitr")




```


\renewcommand{\arraystretch}{1.4}

\clearpage

# Employment and Business


## 5A.1: Unemployment Rate

```{r}

unemployment = usa_data %>% 
  filter( city == 1190, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?
  
  as_survey_design(weights = perwt) %>% 
  survey_count(race_ethnicity, employment=empstat_f, laborforce= labforce_f,  name="count") %>% 
  filter(laborforce == "Yes, in the labor force") %>% 
  group_by(race_ethnicity) %>% 
  mutate(in_laborforce = sum(count)) %>% 
  select(-count_se) %>% 
  pivot_wider(names_from = "employment",
              values_from = "count") %>% 

  mutate(unemployment_rate = 100*round(Unemployed/in_laborforce, 4),
         prc2 = paste0(unemployment_rate, "%"))
  

table_5.a1 <- unemployment %>%
  select(race_ethnicity,  in_laborforce, Employed, Unemployed, prc2) %>%
  
  # using kable
  kbl(booktabs = T,
      col.names = c("Race/Ethnicity",  "In the Labor Force" , "Employed" ,"Unemployed", "Unemployment Rate"),
      align = c("l",rep("c",4)),
      caption = "Unemployment Rate by Race/Ethnicity in Chicago, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "6.45em") %>% 
  # add_header_above(align = c("l", rep("c",3)), c("Race/Ethnicity", "More than 30%" = 2, "More than 50%" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>%

  
  add_footnote(c( "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute"),
 notation = "none")

table_5.a1

```



## 5A.1-2: The Labor Force Participation Rate 

```{r}
LFP_rate = usa_data %>%
  as_survey_design(weights = perwt) %>%
  survey_count(year, race_ethnicity, empstatd,empstatd_f, laborforce= labforce_f,  name="count") %>% 
  filter(empstatd != 0) %>% 
  group_by(year, race_ethnicity) %>% 
  mutate(denominator = sum(count, na.rm = TRUE)) %>% 
  filter(empstatd!= 30) %>% 
  group_by(year, race_ethnicity, denominator) %>% 
  summarise(count = sum(count, na.rm = TRUE)) %>% 
  mutate(prc = 100*round(count/denominator, 4),
         prc2 = paste0(prc, "%"))



table_5.a1_2 <- LFP_rate %>% 
  select(year, race_ethnicity, prc2) %>% 
  pivot_wider(names_from = year,
              values_from = prc2) %>% 
  # using kable
  kbl(booktabs = T,
      col.names = c("Race/Ethnicity",  "2000" , "2008-2012" ,"2018-2022" ),
      align = c("l",rep("c",3)),
      caption = "Labor Force Participation Rate by Race/Ethnicity in Chicago, 2000 to 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "8.6em") %>% 
  # add_header_above(align = c("l", rep("c",3)), c("Race/Ethnicity", "More than 30%" = 2, "More than 50%" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>%

  
  add_footnote(c( "Source: IPUMS-USA database (2000, 2008-2012, 2018-2022). Tabulations by Great Cities Institute"),
 notation = "none")

table_5.a1_2
```

## 5A.2: Numbers and percentages of Mexicans employed, by industry

\blandscape

```{r}

industry = usa_data %>%
  filter( city == 1190, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?

  mutate(industry = case_when(ind %in% c(0170:0490)  ~ "Agriculture, Forestry, Fishing, and Hunting, and Mining",
                              ind == 0770  ~ "Construction",
                              ind %in% c(1070:3990)  ~ "Manufacturing",
                              ind %in% c(4070:4590)  ~ "Wholesale Trade",
                              ind %in% c(4670:5790)  ~ "Retail Trade",
                              ind %in% c(6070:6390, 0570:0690) ~ "Transportation and Warehousing, and Utilities",
                              ind %in% c(6470:6780)  ~ "Information",
                              ind %in% c(6870:7190)  ~ "Finance and Insurance, and Real Estate, and Rental and Leasing",
                              ind %in% c(7270:7790)  ~ "Professional, Scientific, and Management, and Administrative, and Waste Management Services",
                              ind %in% c(7860:8470)  ~ "Educational Services, and Health Care and Social Assistance",
                              ind %in% c(8561:8690)  ~ "Arts, Entertainment, and Recreation, and Accommodation and Food Services",
                              ind %in% c(8770:9290)  ~ "Other Services, Except Public Administration",
                              ind %in% c(9370:9590)  ~ "Public Administration",
                              ind %in% c(9670:9870)  ~ "Military",
                              ind == 9920  ~ "Unemployed, last worked 5 years ago or earlier or never worked",

                                    TRUE ~ NA_character_)) %>%
  
  as_survey_design(weights = perwt) %>%
  survey_count(race_ethnicity, industry, employment=empstat_f, name="count") %>%
  filter(employment == "Employed") %>%
  
  group_by(industry ) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>%
  mutate(prc = 100* round(count/total_employed, 4),
         prc2 = paste0(prc,"%")) %>%
  ungroup()




table_5a.2 <-  industry %>%
  # filter(race_ethnicity %in% c("Other Latinos")) %>% 
  select(race_ethnicity, industry, count,  prc2) %>%
  pivot_wider(names_from = race_ethnicity,
              values_from = c(count, prc2)) %>% 
  select(industry, count_Mexican, prc2_Mexican, `count_Other Latinos`, `prc2_Other Latinos`, `count_White (non-Hispanic or Latino)`, `prc2_White (non-Hispanic or Latino)`, `count_Black (non-Hispanic or Latino)`, `prc2_Black (non-Hispanic or Latino)`, `count_Other (non-Hispanic or Latino)`, `prc2_Other (non-Hispanic or Latino)`) %>% 
  

   # using kable
  kable(booktabs = T,
        col.names = c("", "Number", "% Share of Employed", 
                      "Number", "% Share of Employed",  
                      "Number", "% Share of Employed", 
                      "Number", "% Share of Employed",
                      "Number", "% Share of Employed"),
      align = c("l",rep("c",8)),
      caption = "Employment Distribution by Industry and Race/Ethnicity in Chicago, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position",
                ) %>%
  column_spec(1, width = "12em") %>%
  column_spec(2:11, width = "4.8em") %>%
  add_header_above( c("Industry", "Mexican" = 2, "Other Latinos" = 2,"White\n(non-Hispanic or Latino)" = 2, "Black\n(non-Hispanic or Latino)" = 2,  "Other\n(non-Hispanic or Latino)" = 2),
                    # background = "#cbc9e2",
                    bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>%
  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "") 
  # kableExtra::landscape()
  

table_5a.2


```

\elandscape

\clearpage

```{r}

industry2 = usa_data %>%
  filter( city == 1190, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?

  as_survey_design(weights = perwt) %>%
  survey_count(race_ethnicity, ind, employment=empstat_f, name="count") %>%
  filter(employment == "Employed") %>%
  
  group_by(ind) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>%
  mutate(prc = 100* round(count/total_employed, 4),
         prc2 = paste0(prc,"%")) %>% 
  ungroup()
  # mutate(mexican_count = ifelse(race_ethnicity == "Mexican", count, NA)) %>%
  # arrange(desc(mexican_count))


top_ind_mxc <- industry2 %>%
  filter(race_ethnicity == "Mexican") %>%
  arrange(desc(count)) %>%
  slice(1:10) %>%
  pull(ind)


top_ind_latinos <- industry2 %>%
  filter(race_ethnicity == "Other Latinos") %>%
  arrange(desc(count)) %>%
  slice(1:10) %>%
  pull(ind)


industry3 <- industry2 %>% 
  filter(ind %in% top_ind_mxc & race_ethnicity == "Mexican" | ind %in% top_ind_latinos & race_ethnicity == "Other Latinos" ) %>% 
  
  mutate(sub_industry = case_when(ind == 8680 ~ "Restaurants and other food services",
                    ind == 770 ~ "The cleaning of buildings and dwellings is incidental during construction and immediately after construction",
                    ind == 7860 ~ "Elementary and secondary schools",
                    ind == 8191 ~ "General medical and surgical hospitals, and specialty (except psychiatric and substance abuse) hospitals",
                    ind == 7690 ~ "Services to buildings and dwellings (except cleaning during construction and immediately after construction)",
                    ind == 4971 ~ "Supermarkets and other grocery (except convenience) stores",
                    ind == 8770 ~ "Automotive repair and maintenance",
                    ind == 7870 ~ "Colleges, universities, and professional schools, including junior colleges",
                    ind == 7770 ~ "Landscaping services",
                    ind == 3990 ~ "Not specified manufacturing industries",
                    ind == 7380 ~ "Computer systems design and related services",
                    ind == 8470 ~ "Child day care services",
                    ind == 9470 ~ "Justice, public order, and safety activities"),
         
         group_ind = case_when(ind == 8680 ~ "Accommodation and Food Services",
                    ind == 770 ~ "Construction",
                    ind == 7860 ~ "Educational Services",
                    ind == 8191 ~ "Health Care and Social Assistance",
                    ind == 7690 ~ "Administrative and support and waste management services",
                    ind == 4971 ~ "Retail Trade",
                    ind == 8770 ~ "Other Services, Except Public Administration",
                    ind == 7870 ~ "Educational Services",
                    ind == 7770 ~ "Administrative and support and waste management services",
                    ind == 3990 ~ "Manufacturing",
                    ind == 7380 ~ "Professional, Scientific, and Technical Services",
                    ind == 8470 ~ "Health Care and Social Assistance",
                    ind == 9470 ~ "Public Administration"
                    )) %>% 
  arrange(factor(race_ethnicity, levels = levels(industry2$race_ethnicity)), desc(count)) %>%
  ungroup()



table_5a.2_2 <-  industry3 %>%

  select(group_ind, sub_industry, count,  prc2) %>%

   # using kable
  kable(booktabs = T,
        col.names = c("Group", "Sub Industry", "Number", "% Share of Employed"   ),
      align = c("l", "l", rep("c",2)),
      caption = "The Top 10 sub-industries where Mexicans and Other Latinos have the highest employment in Chicago, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                # latex_options = "HOLD_position",
                ) %>%
  column_spec(1:2, width = "20em") %>%
  column_spec(3:4, width = "7em") %>%
  # add_header_above( c("Race/Ethnicity", "2000" = 2, "2008-2012" = 2, "2018-2022" = 2, "Change from\n2000 to 2018-2022" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8)  %>% 
  pack_rows("Mexican", 1,10, hline_after = T  ) %>% 
  pack_rows("Other Latinos", 11, 20, hline_before = T , hline_after = T) %>% 
    footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "") 

table_5a.2_2 



```


\clearpage

\blandscape
## 5A.3: Numbers and percentages of Mexicans employed, by Occupation


```{r}

occupation = usa_data %>%
  filter( city == 1190, year == 2022) %>%  
  
  mutate(occupation = case_when(occ %in% c(0010:3550)  ~ "Management, Business, and Financial Occupations",
                                occ %in% c(3601:4655)  ~ "Service Occupations",
                                occ %in% c(4700:5940)  ~ "Sales and Office Occupations",
                                occ %in% c(6005:7640)  ~ "Natural Resources, Construction, and Maintenance Occupations",
                                occ %in% c(7700:9760)  ~ "Production, Transportation, and Material Moving Occupations",
                                    TRUE ~ NA_character_)) %>%

  as_survey_design(weights = perwt) %>%
  survey_count(race_ethnicity, occupation, employment=empstat_f, name="count") %>%
  filter(employment == "Employed") %>%
  group_by(occupation) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>%
  mutate(prc = 100* round(count/total_employed, 4),
         prc2 = paste0(prc,"%")) %>%
  filter(!is.na(occupation)) %>% 
  ungroup()




table_5a.3 <-  occupation %>%
  # filter(race_ethnicity %in% c("Other Latinos")) %>% 
  select(race_ethnicity, occupation, count,  prc2) %>%
  pivot_wider(names_from = race_ethnicity,
              values_from = c(count, prc2)) %>% 
  select(occupation, count_Mexican, prc2_Mexican, `count_Other Latinos`, `prc2_Other Latinos`, `count_White (non-Hispanic or Latino)`, `prc2_White (non-Hispanic or Latino)`, `count_Black (non-Hispanic or Latino)`, `prc2_Black (non-Hispanic or Latino)`, `count_Other (non-Hispanic or Latino)`, `prc2_Other (non-Hispanic or Latino)`) %>% 

   # using kable
  kable(booktabs = T,
        col.names = c("", "Number", "% Share of Employed", 
                      "Number", "% Share of Employed",  
                      "Number", "% Share of Employed", 
                      "Number", "% Share of Employed",
                      "Number", "% Share of Employed"),
      align = c("l",rep("c",8)),
      caption = "Employment Distribution by Occupation Categories and Race/Ethnicity in Chicago, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position",
                ) %>%
  column_spec(1, width = "12em") %>%
  column_spec(2:11, width = "4.8em") %>%
  add_header_above( c("Occupation", "Mexican" = 2, "Other Latinos" = 2,"White\n(non-Hispanic or Latino)" = 2, "Black\n(non-Hispanic or Latino)" = 2,  "Other\n(non-Hispanic or Latino)" = 2),
                    # background = "#cbc9e2",
                    bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>%
  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "") 
  

table_5a.3

```

\elandscape
\clearpage

```{r}

occupation2 <- usa_data %>%
  filter( city == 1190, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?

  as_survey_design(weights = perwt) %>%
  survey_count(race_ethnicity, occ, employment=empstat_f, name="count") %>%
  filter(employment == "Employed") %>%
  group_by(occ) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>%
  mutate(prc = 100* round(count/total_employed, 4),
         prc2 = paste0(prc,"%")) %>%
  filter(!is.na(occ)) %>% 
  ungroup()


top_occ_mxc <- occupation2 %>%
  filter(race_ethnicity == "Mexican") %>%
  arrange(desc(count)) %>%
  slice(1:10) %>%
  pull(occ)

top_occ_latinos <- occupation2 %>%
  filter(race_ethnicity == "Other Latinos") %>%
  arrange(desc(count)) %>%
  slice(1:10) %>%
  pull(occ)



occupation3 <- occupation2 %>% 
  filter(occ %in% top_occ_mxc & race_ethnicity == "Mexican"|
         occ %in% top_occ_latinos & race_ethnicity == "Other Latinos") %>% 
  
  mutate(sub_Occupation = case_when(occ == 4220 ~ "Janitors and building cleaners",
                                    occ == 9620 ~ "Laborers and freight, stock, and material movers, hand",
                                    occ == 4020 ~ "Cooks",
                                    occ == 4720 ~ "Cashiers",
                                    occ == 9130 ~ "Driver/sales workers and truck drivers",
                                    occ == 5240 ~ "Customer service representatives",
                                    occ == 6260 ~ "Construction laborers",
                                    occ == 4760 ~ "Retail salespersons",
                                    occ == 4110 ~ "Waiters and waitresses",
                                    occ == 8990 ~ "Miscellaneous production workers, including equipment operators and tenders",
                                    occ == 440 ~ "Other managers",
                                    occ == 4700 ~ "First-line supervisors of retail sales workers"), 
  Occupation_group = case_when(occ == 4220 ~ "Service Occupations",
                               occ == 9620 ~ "Transportation and Material Moving Occupations",
                               occ == 4020 ~ "Service Occupations",
                               occ == 4720 ~ "Sales and Related Occupations",
                               occ == 9130 ~ "Transportation and Material Moving Occupations",
                               occ == 5240 ~ "Office and Administrative Support Occupations",
                                occ == 6260 ~ "Construction and Extraction Occupations",
                               occ == 4760 ~ "Sales and Related Occupations",
                               occ == 4110 ~ "Service Occupations",
                               occ == 8990 ~ "Production Occupations",
                               occ == 440 ~ "Management, Business, and Financial Occupations",
                               occ == 4700 ~ "Sales and Related Occupations"
                               
                                    )) %>%
  arrange(factor(race_ethnicity, levels = levels(occupation2$race_ethnicity)), desc(count)) %>%
  ungroup()

##########


table_5a.3_2 <-  occupation3 %>%

  select(Occupation_group, sub_Occupation, count,  prc2) %>%

   # using kable
  kable(booktabs = T,
        col.names = c("Occupation Group","Occupation", "Number", "% Share of Employed"   ),
      align = c("l","l" ,rep("c",2)),
      caption = "The Top 10 Occupations with the Highest Employment of Mexicans and Other Latinos in Chicago (2018-2022)",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                # latex_options = "HOLD_position",
                ) %>%
  column_spec(1:2, width = "20em") %>%
  column_spec(3:4, width = "7em") %>%
  # add_header_above( c("Race/Ethnicity", "2000" = 2, "2008-2012" = 2, "2018-2022" = 2, "Change from\n2000 to 2018-2022" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>% 
  pack_rows("Mexican", 1,10, hline_after = T  ) %>% 
  pack_rows("Other Latinos", 11, 20, hline_before = T , hline_after = T) 

  

table_5a.3_2 


```

\clearpage

## 5C: Numbers and percentages of unionized Mexican workers

```{r}
  
options(knitr.kable.NA = '-') 

cps_ddi <- read_ipums_micro(read_ipums_ddi("C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/cps_00171.xml")) %>% 
  clean_names() 


Union <- cps_ddi %>% 
  filter( year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?
  mutate(race_f = as_factor(race),
         hispan_f = as_factor(hispan), 
         laborforce_f = as_factor(labforce),
         employment_f = as_factor(empstat),
         union_f = as_factor(union)) %>%

  mutate(race_ethnicity = case_when(hispan ==0 & race == 100 ~ "White (non-Hispanic or Latino)",
                                    hispan ==0 & race == 200 ~ "Black (non-Hispanic or Latino)",
                                    hispan %in% c(200:612) ~ "Other Latinos",
                                    hispan %in% c(100:109) ~ "Mexican", 
                                    hispan ==0 & !race %in%c(100,200) ~ "Other (non-Hispanic or Latino)",
                                    TRUE ~ NA_character_)) %>% 
  mutate(race_ethnicity = factor(race_ethnicity, level = c("Mexican", "Other Latinos","White (non-Hispanic or Latino)", "Black (non-Hispanic or Latino)",  "Other (non-Hispanic or Latino)"))) %>% 
  
  mutate(employment_f = case_when(empstat %in% c(1, 10, 12) ~ "Employed",
                              TRUE ~ "other")) %>% 
  
  
  filter(labforce != 0 |  empstat !=0) %>% 
  filter(employment_f == "Employed") %>%
  as_survey_design(weights = earnwt) %>% 
  
  survey_count(race_ethnicity, laborforce_f, employment_f,  union_f, name="count") %>% 
  group_by(race_ethnicity) %>% 
  mutate(total_employed = sum(count)) %>% 
  filter(union_f %in% c("Member of labor union", "Covered by union but not a member")) %>% 
  # group_by(race_ethnicity, total_employed) %>% 
  mutate(prc = 100*round(count/total_employed, 4),
            prc2 = paste0(prc, "%")) %>% 
  select(race_ethnicity,union_f, total_employed,  count, prc2) %>% 
  pivot_wider(names_from = union_f, values_from = c(count, prc2))
  


table5b <- Union %>% 
  select(race_ethnicity,  total_employed,  `count_Member of labor union`, `prc2_Member of labor union`, `count_Covered by union but not a member`, `prc2_Covered by union but not a member`  ) %>% 


  kbl(booktabs = T,
      col.names = c("",  "" ,"Number", "Percent", "Number", "Percent"),
      align = c("l",rep("c",5)),
      caption = " Numbers and percentages of unionized Mexican workers by Race/Ethnicity in Illinois, 2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "5.16em") %>%
  add_header_above(align = c("l", rep("c",3)), c("Race/Ethnicity",  "Total Employed","Member of labor union" =2 ,"Covered by union but not a member" = 2),
                    # background = "#cbc9e2",
                    bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>%

  
  add_footnote(c( "Source: IPUMS-CPS database (2022). Tabulations by Great Cities Institute"),
 notation = "none")

table5b

```


\clearpage

## 5D: Transportation

```{r}
tranwork <- usa_data %>% 

  filter( city == 1190, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?
  
  as_survey_design(weights = perwt) %>% 
  survey_count(race_ethnicity, employment=empstat_f, laborforce= labforce_f, tranwork_f, name="count") %>% 
  filter(employment == "Employed") %>%
  mutate(race_ethnicity = if_else(race_ethnicity%in% c("White (non-Hispanic or Latino)", "Black (non-Hispanic or Latino)", "Other (non-Hispanic or Latino)"), "Non-Hispanic or Latino",race_ethnicity )) %>% 
  
  
  group_by(race_ethnicity) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>% 
  group_by(race_ethnicity, tranwork_f, total_employed) %>% 
  summarise(count = sum(count, na.rm =TRUE)) %>% 
  mutate(prc = 100 * round(count/total_employed, 4),
         prc2 = paste0(prc, "%")) 
  # mutate(tranwork_f = if_else(tranwork_f== "Light rail, streetcar, or trolley (Carro p\u00fablico in PR)", "Light rail, streetcar, or trolley", tranwork_f ))
  

# options(knitr.kable.NA = '-')   
  
table_5.d <- tranwork %>%
  pivot_wider(id_cols =tranwork_f, names_from = race_ethnicity,
              values_from = c(count, prc2 )) %>%
  # mutate(tranwork_f = factor(tranwork_f, levels = c("Auto, truck, or van", "Motorcycle", "Bus", "Light rail, streetcar, or trolley", "Subway or elevated", "Long-distance train or commuter train", "Taxicab", "Ferryboat",	"Bicycle", "Walked only", "Other", "Worked at home", "N/A"))) %>% 
  select(tranwork_f, count_Mexican, prc2_Mexican, `count_Other Latinos` ,`prc2_Other Latinos`, `count_Non-Hispanic or Latino`, `prc2_Non-Hispanic or Latino` ) %>%
  
  # using kable
  kbl(booktabs = T,
      col.names = c("", "Number", "% Share of Employed",
                    "Number", "% Share of Employed",
                    "Number", "% Share of Employed"),
      align = c("l",rep("c",6)), 
      caption = "Primary Means of Transportation to Work Race/Ethnicity in Chicago, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                # latex_options = "HOLD_position",
                ) %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:7, width = "5.16em") %>%
  add_header_above( c("Means of Transportation", "Mexican" = 2, "Other Latinos" = 2, "Non-Hispanic or Latino" = 2),
                    # background = "#cbc9e2",
                    bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8) %>%

  footnote(general = "Source: IPUMS-USA database (2018-2022). Tabulations by Great Cities Institute",
           general_title = "") 

table_5.d
```


