---
title: "Mexican Report"
author: "Employment and Business"
date: "Great Cities Institute"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    highlight: pygments
    extra_dependencies: ["float"]
    fig_caption: true
mainfont: "Arial"
header-includes:
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{caption} 
  - \usepackage{booktabs} 
  - \usepackage{longtable}
  - \usepackage{threeparttablex}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F, fig.pos="H")
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

library(ipumsr)
library(tidyverse)
library(purrr)
library(sf)
library(tidycensus)
library(tidyr)
library(readxl)
library(sf)
library(htmltools)
library(leaflet)
library(janitor)
library(data.table)
library(survey) # for survey analysis
library(srvyr)  # for tidy survey analysis
library(palmerpenguins)
library("kableExtra") 
library(stringr)
library(knitr)
library(palmerpenguins)
options(tinytex.verbose = TRUE)


usa_data <- read.csv("C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/final_Mexican_IL_2000_22.csv") %>% 
  mutate(race_ethnicity = case_when(hispan ==0 & race == 1 ~ "White (non-Hispanic or Latino)",
                                    hispan ==0 & race == 2 ~ "Black (non-Hispanic or Latino)",
                                    hispan %in% c(2,3,4) ~ "Other Hispanics or Latinos",
                                    hispan ==1 ~ "Mexican", 
                                    hispan ==0 & race %in%c(3:9) ~ "Other (non-Hispanic or Latino)",
                                    TRUE ~ NA_character_)) %>% 
  mutate(race_ethnicity = factor(race_ethnicity, level = c("Mexican", "Other Hispanics or Latinos","White (non-Hispanic or Latino)", "Black (non-Hispanic or Latino)",  "Other (non-Hispanic or Latino)"))) %>% 
  mutate(hispan_breakdown = case_when(
  hispand == 0 ~ "Not Hispanic",
  hispand == 100 ~ "Mexican", 
  hispand == 200 ~ "Puerto Rican",
  hispand == 300 ~ "Cuban",
  hispand == 411 ~ "Costa Rican",
  hispand == 412 ~ "Guatemalan",
  hispand == 413 ~ "Honduran",
  hispand == 414 ~ "Nicaraguan",
  hispand == 415 ~ "Panamanian",
  hispand == 416 ~ "Salvadoran",
  hispand == 417 ~ "Central American, not specified",
  hispand == 420 ~ "Argentinean",
  hispand == 421 ~ "Bolivian",
  hispand == 422 ~ "Chilean",
  hispand == 423 ~ "Colombian",
  hispand == 424 ~ "Ecuadorian",
  hispand == 425 ~ "Paraguayan",
  hispand == 426 ~ "Peruvian",
  hispand == 427 ~ "Uruguayan",
  hispand == 428 ~ "Venezuelan",
  hispand == 431 ~ "South American, not specified",
  hispand == 450 ~ "Spaniard",
  hispand == 460 ~ "Dominican",
  hispand == 498 ~ "Other, not specified",
  TRUE ~ "Other"
),
mex_foreign_born = case_when(
    race_ethnicity == "Mexican" & bpl %in% as.character(1:56) ~ "U.S. Born",
    race_ethnicity == "Mexican" & bpl == "200" ~ "Mexico Born",
    TRUE ~ NA_character_  # Optional: default case for other values or groups
  )) %>% ## filter for hispanic groups here ##
  mutate(mex_foreign_born = factor(mex_foreign_born, level = c("Mexico Born","U.S. Born"))) %>% 
  
  mutate(hispan_breakdown = factor(hispan_breakdown, level = c("Mexican", "Puerto Rican","Ecuadorian", "Cuban", "Guatemalan", "Venezuelan", "Other, not specified", "Colombian", "Salvadoran", "Peruvian", "Honduran", "Spaniard", "Dominican", "Argentinean", "Chilean", "Nicaraguan", "Panamanian", "Costa Rican", "Bolivian", "Uruguayan", "Paraguayan", "South American, not specified", "Central American, not specified", "Not Hispanic"))
  )
  
# install.packages("knitr")


theme_plots <-
  theme_minimal() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.line = element_line(color="#969696", linewidth = 1),
        plot.title = element_text(face ="bold", size = 10, hjust = 0.5),
      plot.subtitle = element_text(face ="italic", size = 8, color = "#636363"),
      axis.text = element_text(size = 9),
      axis.title = element_text(face ="bold", size = 9),
      plot.caption = element_text(hjust = 0, size = 8, color = "black"),
      legend.text = element_text(size = 8, color = "#636363"),
      legend.title = element_text(face = "bold", size = 8, color = "#636363")) 

color_race_ethnicity <- c("Mexican" = "#CE1126", 
                          "Other Hispanics or Latinos" = "#F9AA51",
                               "White (non-Hispanic or Latino)" = "#9CA168",
                               "Black (non-Hispanic or Latino)" = "#8CBEC0",
                               "Other (non-Hispanic or Latino)" = "#0B8489")

options(knitr.kable.NA = '-')
```


\renewcommand{\arraystretch}{1.4}

\clearpage

# Employment and Business


## 5A.1: Unemployment Rate

```{r}

unemployment = usa_data %>% 
  filter( city == 1190, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?
  
  as_survey_design(weights = perwt) %>% 
  survey_count(race_ethnicity, employment=empstat_f, laborforce= labforce_f,  name="count") %>% 
  filter(laborforce == "Yes, in the labor force") %>% 
  group_by(race_ethnicity) %>% 
  mutate(in_laborforce = sum(count)) %>% 
  select(-count_se) %>% 
  pivot_wider(names_from = "employment",
              values_from = "count") %>% 

  mutate(unemployment_rate = 100*round(Unemployed/in_laborforce, 4),
         prc2 = paste0(format(round(unemployment_rate, 1), nsmall = 1), "%"))
  

table_5.a1 <- unemployment %>%
  select(race_ethnicity,  in_laborforce, Employed, Unemployed, prc2) %>%
  
  # using kable
  kbl(booktabs = T,
      col.names = c("Race/Ethnicity",  "In the Labor Force" , "Employed" ,"Unemployed", "Unemployment Rate"),
      align = c("l",rep("r",4)),
      caption = "Labor Force Status for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "6.45em") %>% 

  row_spec(0, bold = T, align = "c",
           font_size = 8) %>%

  add_footnote(c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
notation = "none",
threeparttable = T,
escape = FALSE)

table_5.a1

```



## 5A.1-2: The Labor Force Participation Rate

```{r}
LFP_rate = usa_data %>%
  as_survey_design(weights = perwt) %>%
  survey_count(year, race_ethnicity, empstatd,empstatd_f, laborforce= labforce_f,  name="count") %>%
  filter(empstatd != 0) %>%
  group_by(year, race_ethnicity) %>%
  mutate(denominator = sum(count, na.rm = TRUE)) %>%
  filter(empstatd!= 30) %>%
  group_by(year, race_ethnicity, denominator) %>%
  summarise(count = sum(count, na.rm = TRUE)) %>%
  mutate(prc = 100*round(count/denominator, 4),
         prc2 = paste0(format(round(prc, 1), nsmall = 1), "%"))



table_5.a1_2 <- LFP_rate %>%
  select(year, race_ethnicity, prc2) %>%
  pivot_wider(names_from = year,
              values_from = prc2) %>%
  # using kable
  kbl(booktabs = T,
      col.names = c("Race/Ethnicity",  "2000" , "2008-2012" ,"2018-2022" ),
      align = c("l",rep("r",3)),
      caption = "Labor Force Participation Rate for Mexicans and Other Racial/Ethnic Groups in Chicago, 2000 (Decennial Census), 2008-2012, and 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "8.6em") %>%
  # add_header_above(align = c("l", rep("c",3)), c("Race/Ethnicity", "More than 30%" = 2, "More than 50%" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T,
           # background = "#f2f0f7",
           font_size = 8,
           align = "c") %>%


  add_footnote(c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
notation = "none",
threeparttable = T,
escape = FALSE)

table_5.a1_2
```



```{r fig.width=6, fig.height=4 ,fig.align='center'}



LFP_rate %>%
  mutate(year = case_when(year == 2000 ~ "2000",
                          year == 2012 ~ "2008-2012",
                          year == 2022 ~ "2018-2022"),
         year = factor(year, levels = c("2000", "2008-2012", "2018-2022"))) %>%

  ggplot(aes(x =  year,
             y = prc,
             fill = race_ethnicity)) +
  geom_bar(position=position_dodge(width=0.7), stat = "identity",
           color = "white",
           width=0.7) +
  labs(y="", x = "",
       title="Labor Force Participation Rate for Mexicans and Other Racial/Ethnic Groups in Chicago,\n2000 (Decennial Census), 2008-2012, and 2018-2022 (ACS 5-year Estimates)",
       caption = "Data Source: IPUMS USA, University of Minnesota, www.ipums.org\n(2000 Decennial Census, 2018-2012, and 2018-2022 American Community Survey 5-year Estimates).\nTabulations by Great Cities Institute.",
       fill="") +
  scale_fill_manual(values = color_race_ethnicity) +
  geom_text(aes(label = prc2),
            position=position_dodge(width=0.7),
            vjust = -0.5,
            hjust = 0.5,
            size = 2.5) +

  scale_x_discrete(labels = function(x) str_wrap(x, width = 20)) + # Modify labels of ggplot2 barplot
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0,100), breaks = seq(0, 100, by = 100)) +

  theme_plots +
  theme(
    # axis.text.x = element_text(size = 8),
        panel.background = element_blank(),
        # axis.ticks.y = element_blank(),
        # axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size = 7, color = "#636363"),
        legend.key.size = unit(0.5, "lines"))+
  guides(fill = guide_legend(nrow = 2, byrow = TRUE))


```


## 5A.2: Numbers and percentages of Mexicans employed, by industry

\blandscape

```{r}

industry = usa_data %>%
  filter(year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?

  mutate(industry = case_when(ind %in% c(0170:0490)  ~ "Agriculture, Forestry, Fishing, and Hunting, and Mining",
                              ind == 0770  ~ "Construction",
                              ind %in% c(1070:3990)  ~ "Manufacturing",
                              ind %in% c(4070:4590)  ~ "Wholesale Trade",
                              ind %in% c(4670:5790)  ~ "Retail Trade",
                              ind %in% c(6070:6390, 0570:0690) ~ "Transportation and Warehousing, and Utilities",
                              ind %in% c(6470:6780)  ~ "Information",
                              ind %in% c(6870:7190)  ~ "Finance and Insurance, and Real Estate, and Rental and Leasing",
                              ind %in% c(7270:7790)  ~ "Professional, Scientific, and Management, and Administrative, and Waste Management Services",
                              ind %in% c(7860:8470)  ~ "Educational Services, and Health Care and Social Assistance",
                              ind %in% c(8561:8690)  ~ "Arts, Entertainment, and Recreation, and Accommodation and Food Services",
                              ind %in% c(8770:9290)  ~ "Other Services, Except Public Administration",
                              ind %in% c(9370:9590)  ~ "Public Administration",
                              ind %in% c(9670:9870)  ~ "Military",
                              ind == 9920  ~ "Unemployed, last worked 5 years ago or earlier or never worked",

                                    TRUE ~ NA_character_)) %>%

  as_survey_design(weights = perwt) %>%
  survey_count(race_ethnicity, industry, employment=empstat_f, name="count") %>%
  filter(employment == "Employed") %>%

  group_by(industry ) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>%
  mutate(prc = 100* round(count/total_employed, 4),
         prc2 = paste0(format(round(prc, 1), nsmall = 1), "%")) %>%
  ungroup()




table_5a.2 <-  industry %>%
  # filter(race_ethnicity %in% c("Other Latinos")) %>%
  select(race_ethnicity, industry, count,  prc2) %>%
  pivot_wider(names_from = race_ethnicity,
              values_from = c(count, prc2)) %>%
  select(industry, count_Mexican, prc2_Mexican, `count_Other Hispanics or Latinos`, `prc2_Other Hispanics or Latinos`, `count_White (non-Hispanic or Latino)`, `prc2_White (non-Hispanic or Latino)`, `count_Black (non-Hispanic or Latino)`, `prc2_Black (non-Hispanic or Latino)`, `count_Other (non-Hispanic or Latino)`, `prc2_Other (non-Hispanic or Latino)`) %>%


   # using kable
  kable(booktabs = T,
        col.names = c("", "Number", "% Share of Industry Employment",
                      "Number", "% Share of Industry Employment",
                      "Number", "% Share of Industry Employment",
                      "Number", "% Share of Industry Employment",
                      "Number", "% Share of Industry Employment"),
      align = c("l",rep("r",10)),
      caption = "Employment by Industry and Percent Share of Industry Employment for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "") %>%

  kable_classic(full_width = F,
                font_size = 7.5,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position",
                ) %>%
  column_spec(1, width = "13em") %>%
  column_spec(2:11, width = "5.2em") %>%
  add_header_above( c("Industry", "Mexican" = 2, "Other Hispanics\nor Latinos" = 2,"White\n(non-Hispanic or Latino)" = 2, "Black\n(non-Hispanic or Latino)" = 2,  "Other\n(non-Hispanic or Latino)" = 2),
                    # background = "#cbc9e2",
                    bold = T, font_size =7.5) %>%
  row_spec(0, align = "c",
           # bold = T,
           # background = "#f2f0f7",
           
           font_size = 7
           ) %>%
  footnote(general = "\\\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
           general_title = "",
           threeparttable = T,
           escape = FALSE)
#   add_footnote(c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
# notation = "none",
# threeparttable = T,
# escape = FALSE
# )


table_5a.2


```

\elandscape

\clearpage



```{r}

industry2 = usa_data %>%
  filter( city == 1190, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?

  as_survey_design(weights = perwt) %>%
  survey_count(race_ethnicity, ind, employment=empstat_f, name="count") %>%
  filter(employment == "Employed") %>%

  group_by(ind) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>%
  mutate(prc = 100* round(count/total_employed, 4),
         prc2 = paste0(format(round(prc, 1), nsmall = 1), "%")) %>%
  ungroup()
  # mutate(mexican_count = ifelse(race_ethnicity == "Mexican", count, NA)) %>%
  # arrange(desc(mexican_count))


top_ind_mxc <- industry2 %>%
  filter(race_ethnicity == "Mexican") %>%
  arrange(desc(count)) %>%
  slice(1:10) %>%
  pull(ind)


top_ind_latinos <- industry2 %>%
  filter(race_ethnicity == "Other Hispanics or Latinos") %>%
  arrange(desc(count)) %>%
  slice(1:10) %>%
  pull(ind)


industry3 <- industry2 %>%
  filter(ind %in% top_ind_mxc & race_ethnicity == "Mexican" | ind %in% top_ind_latinos & race_ethnicity == "Other Hispanics or Latinos" ) %>%

  mutate(sub_industry = case_when(ind == 8680 ~ "Restaurants and other food services",
                    ind == 770 ~ "The cleaning of buildings and dwellings is incidental during construction and immediately after construction",
                    ind == 7860 ~ "Elementary and secondary schools",
                    ind == 8191 ~ "General medical and surgical hospitals, and specialty (except psychiatric and substance abuse) hospitals",
                    ind == 7690 ~ "Services to buildings and dwellings (except cleaning during construction and immediately after construction)",
                    ind == 4971 ~ "Supermarkets and other grocery (except convenience) stores",
                    ind == 8770 ~ "Automotive repair and maintenance",
                    ind == 7870 ~ "Colleges, universities, and professional schools, including junior colleges",
                    ind == 7770 ~ "Landscaping services",
                    ind == 3990 ~ "Not specified manufacturing industries",
                    ind == 7380 ~ "Computer systems design and related services",
                    ind == 8470 ~ "Child day care services",
                    ind == 9470 ~ "Justice, public order, and safety activities"),

         group_ind = case_when(ind == 8680 ~ "Accommodation and Food Services",
                    ind == 770 ~ "Construction",
                    ind == 7860 ~ "Educational Services",
                    ind == 8191 ~ "Health Care and Social Assistance",
                    ind == 7690 ~ "Administrative and support and waste management services",
                    ind == 4971 ~ "Retail Trade",
                    ind == 8770 ~ "Other Services, Except Public Administration",
                    ind == 7870 ~ "Educational Services",
                    ind == 7770 ~ "Administrative and support and waste management services",
                    ind == 3990 ~ "Manufacturing",
                    ind == 7380 ~ "Professional, Scientific, and Technical Services",
                    ind == 8470 ~ "Health Care and Social Assistance",
                    ind == 9470 ~ "Public Administration"
                    )) %>%
  arrange(factor(race_ethnicity, levels = levels(industry2$race_ethnicity)), desc(count)) %>%
  ungroup()



table_5a.2_2 <-  industry3 %>%

  select(group_ind, sub_industry, count,  prc2) %>%

   # using kable
  kable(booktabs = T,
        col.names = c("Group", "Sub-industries", "Number", "% Share of Industry Employment"   ),
      align = c("l", "l", rep("r",2)),
      caption = "Employment by Industry and Percent Share of Industry Employment for 10 Sub-industries Where Hispanics or Latinos Have the Highest Number of Employees for Mexicans and Other Hispanics or Latinos in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                # latex_options = "HOLD_position",
                ) %>%
  column_spec(1:2, width = "20em") %>%
  column_spec(3:4, width = "5.5em") %>%
  # add_header_above( c("Race/Ethnicity", "2000" = 2, "2008-2012" = 2, "2018-2022" = 2, "Change from\n2000 to 2018-2022" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T, align = "c",
           # background = "#f2f0f7",
           font_size = 8)  %>%
  pack_rows("Mexican", 1,10, hline_after = T, indent=FALSE  ) %>%
  pack_rows("Other Hispanics or Latinos", 11, 20, hline_before = T , hline_after = T, indent=FALSE) %>%
  footnote(general = "\\\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
           general_title = "",
           threeparttable = T,
           escape = FALSE)



table_5a.2_2

```



\blandscape


## 5A.3: Numbers and percentages of Mexicans employed, by Occupation


```{r}

occupation = usa_data %>%
  filter( city == 1190, year == 2022) %>%

  mutate(occupation = case_when(occ %in% c(0010:3550)  ~ "Management, Business, and Financial Occupations",
                                occ %in% c(3601:4655)  ~ "Service Occupations",
                                occ %in% c(4700:5940)  ~ "Sales and Office Occupations",
                                occ %in% c(6005:7640)  ~ "Natural Resources, Construction, and Maintenance Occupations",
                                occ %in% c(7700:9760)  ~ "Production, Transportation, and Material Moving Occupations",
                                    TRUE ~ NA_character_)) %>%

  as_survey_design(weights = perwt) %>%
  survey_count(race_ethnicity, occupation, employment=empstat_f, name="count") %>%
  filter(employment == "Employed") %>%
  group_by(occupation) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>%
  mutate(prc = 100* round(count/total_employed, 4),
         prc2 = paste0(format(round(prc, 1), nsmall = 1), "%")) %>%
  filter(!is.na(occupation)) %>%
  ungroup()




table_5a.3 <-  occupation %>%
  # filter(race_ethnicity %in% c("Other Latinos")) %>%
  select(race_ethnicity, occupation, count,  prc2) %>%
  pivot_wider(names_from = race_ethnicity,
              values_from = c(count, prc2)) %>%
  select(occupation, count_Mexican, prc2_Mexican, `count_Other Hispanics or Latinos`, `prc2_Other Hispanics or Latinos`, `count_White (non-Hispanic or Latino)`, `prc2_White (non-Hispanic or Latino)`, `count_Black (non-Hispanic or Latino)`, `prc2_Black (non-Hispanic or Latino)`, `count_Other (non-Hispanic or Latino)`, `prc2_Other (non-Hispanic or Latino)`) %>%

   # using kable
  kable(booktabs = T,
        col.names = c("", "Number", "% Share of Occupation Employment",
                      "Number", "% Share of Occupation Employment",
                      "Number", "% Share of Occupation Employment",
                      "Number", "% Share of Occupation Employment",
                      "Number", "% Share of Occupation Employment"),
      align = c("l",rep("r",10)),
      caption = "Employment by Occupation and Percent Share of Occupation Employment for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "") %>%

  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position",
                ) %>%
  column_spec(1, width = "13em") %>%
  column_spec(2:11, width = "5.2em") %>%
  add_header_above( c("Occupation", "Mexican" = 2, "Other Hispanics\nor Latinos" = 2,"White\n(non-Hispanic or Latino)" = 2, "Black\n(non-Hispanic or Latino)" = 2,  "Other\n(non-Hispanic or Latino)" = 2),
                    # background = "#cbc9e2",
                    bold = T, font_size =8) %>%
  row_spec(0, 
           # bold = T, 
           align = "c",
           # background = "#f2f0f7",
           font_size = 7
           ) %>%
  footnote(general = "\\\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
           general_title = "",
           threeparttable = T,
           escape = FALSE)


table_5a.3

```

\elandscape

\clearpage

```{r fig.width=6, fig.height= 4.5 ,fig.align='center'}



occupation %>%
  mutate(race_ethnicity = fct_rev(race_ethnicity)) %>%
  group_by(occupation) %>%
  mutate(prc_mexican = ifelse(race_ethnicity == "Mexican", prc, NA)) %>%
  ungroup() %>%
  arrange(prc_mexican) %>%
  mutate(occupation = factor(occupation, levels = unique(occupation[order(prc_mexican)]))) %>%

  ggplot(aes(x = prc,
             y = occupation,
             fill = race_ethnicity)) +
  geom_bar(position = "stack", stat = "identity",
           color = "white",
           width=2/3) +
  labs(y="", x = "",
       title="Percent Share of Occupation Employment for Mexicans and Other\nRacial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
       caption = "Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American\n Community Survey 5-year Estimates).Tabulations by Great Cities Institute.",
       fill="") +
  scale_fill_manual(values = color_race_ethnicity) +

  geom_text(aes(label = prc2),
            position = position_stack(vjust = 0.5),
            # vjust = -0.5,
            # hjust = 0.5,
            size = 2.5  ,
            fontface = "bold",
            color = "white")+
  scale_y_discrete(labels = function(x) str_wrap(x, width = 20)) + # Modify labels of ggplot2 barplot
  scale_x_continuous(labels = function(x) paste0(x, "%"), limits = c(0,101), breaks = seq(0, 101, by = 100)) +

  theme_plots +
  theme(
    axis.text.y = element_text(size = 8),
        panel.background = element_blank(),
        # axis.ticks.y = element_blank(),
        # axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.line.x = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size = 7, color = "#636363"),
        legend.key.size = unit(0.5, "lines"))+
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) # Arrange legend items in two rows


```
\clearpage


```{r}

occupation2 <- usa_data %>%
  filter( city == 1190, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?

  as_survey_design(weights = perwt) %>%
  survey_count(race_ethnicity, occ, employment=empstat_f, name="count") %>%
  filter(employment == "Employed") %>%
  group_by(occ) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>%
  mutate(prc = 100* round(count/total_employed, 4),
         prc2 = paste0(format(round(prc, 1), nsmall = 1), "%")) %>%
  filter(!is.na(occ)) %>%
  ungroup()


top_occ_mxc <- occupation2 %>%
  filter(race_ethnicity == "Mexican") %>%
  arrange(desc(count)) %>%
  slice(1:10) %>%
  pull(occ)

top_occ_latinos <- occupation2 %>%
  filter(race_ethnicity == "Other Hispanics or Latinos") %>%
  arrange(desc(count)) %>%
  slice(1:10) %>%
  pull(occ)



occupation3 <- occupation2 %>%
  filter(occ %in% top_occ_mxc & race_ethnicity == "Mexican"|
         occ %in% top_occ_latinos & race_ethnicity == "Other Hispanics or Latinos") %>%

  mutate(sub_Occupation = case_when(occ == 4220 ~ "Janitors and building cleaners",
                                    occ == 9620 ~ "Laborers and freight, stock, and material movers, hand",
                                    occ == 4020 ~ "Cooks",
                                    occ == 4720 ~ "Cashiers",
                                    occ == 9130 ~ "Driver/sales workers and truck drivers",
                                    occ == 5240 ~ "Customer service representatives",
                                    occ == 6260 ~ "Construction laborers",
                                    occ == 4760 ~ "Retail salespersons",
                                    occ == 4110 ~ "Waiters and waitresses",
                                    occ == 8990 ~ "Miscellaneous production workers, including equipment operators and tenders",
                                    occ == 440 ~ "Other managers",
                                    occ == 4700 ~ "First-line supervisors of retail sales workers"),
  Occupation_group = case_when(occ == 4220 ~ "Service Occupations",
                               occ == 9620 ~ "Transportation and Material Moving Occupations",
                               occ == 4020 ~ "Service Occupations",
                               occ == 4720 ~ "Sales and Related Occupations",
                               occ == 9130 ~ "Transportation and Material Moving Occupations",
                               occ == 5240 ~ "Office and Administrative Support Occupations",
                                occ == 6260 ~ "Construction and Extraction Occupations",
                               occ == 4760 ~ "Sales and Related Occupations",
                               occ == 4110 ~ "Service Occupations",
                               occ == 8990 ~ "Production Occupations",
                               occ == 440 ~ "Management, Business, and Financial Occupations",
                               occ == 4700 ~ "Sales and Related Occupations"

                                    )) %>%
  arrange(factor(race_ethnicity, levels = levels(occupation2$race_ethnicity)), desc(count)) %>%
  ungroup()

##########


table_5a.3_2 <-  occupation3 %>%

  select(Occupation_group, sub_Occupation, count,  prc2) %>%

   # using kable
  kable(booktabs = T,
        col.names = c("Occupation Group","Sub-occupation ", "Number", "% Share of Occupation Employment"   ),
      align = c("l","l" ,rep("r",2)),
      caption = "Employment by Sub-occupation and Percent Share of Occupation Employment for 10 Sub-occupations Where Hispanics or Latinos Have the Highest Number of Employees for Mexicans and Other Hispanics or Latinos in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                # latex_options = "HOLD_position",
                ) %>%
  column_spec(1:2, width = "20em") %>%
  column_spec(3, width = "5em") %>%
  column_spec(4, width = "6em") %>% 
  # add_header_above( c("Race/Ethnicity", "2000" = 2, "2008-2012" = 2, "2018-2022" = 2, "Change from\n2000 to 2018-2022" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T, align = "c",
           # background = "#f2f0f7",
           font_size = 7.5) %>%
  pack_rows("Mexican", 1,10, hline_after = T  ) %>%
  pack_rows("Other Hispanics or Latinos", 11, 20, hline_before = T , hline_after = T) %>% 
  footnote(general = "\\\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
           general_title = "",
           threeparttable = T,
           escape = FALSE)



table_5a.3_2


```


\clearpage

## 5C: Transportation

```{r}
tranwork <- usa_data %>%
  
  filter( city == 1190, year == 2022) %>%  #### dont forget to filter PUMA   ## Should I filter pernum or not?

  as_survey_design(weights = perwt) %>%
  survey_count(race_ethnicity, employment=empstat_f, laborforce= labforce_f, tranwork_f, name="count") %>%
  filter(employment == "Employed") %>%
  mutate(race_ethnicity = if_else(race_ethnicity%in% c("White (non-Hispanic or Latino)", "Black (non-Hispanic or Latino)", "Other (non-Hispanic or Latino)"), "Non-Hispanic or Latino",race_ethnicity )) %>%


  group_by(race_ethnicity) %>%
  mutate(total_employed = sum(count, na.rm = T)) %>%
  group_by(race_ethnicity, tranwork_f, total_employed) %>%
  summarise(count = sum(count, na.rm =TRUE)) %>%
  mutate(prc = 100 * round(count/total_employed, 4),
         prc2 = paste0(format(round(prc, 1), nsmall = 1), "%")) 
  # mutate(tranwork_f = if_else(tranwork_f== "Light rail, streetcar, or trolley (Carro p\u00fablico in PR)", "Light rail, streetcar, or trolley", tranwork_f ))


# options(knitr.kable.NA = '-')

table_5.d <- tranwork %>%
  pivot_wider(id_cols =tranwork_f, names_from = race_ethnicity,
              values_from = c(count, prc2 )) %>%
  filter( tranwork_f != "N/A") %>% 
  # mutate(tranwork_f = factor(tranwork_f, levels = c("Auto, truck, or van", "Motorcycle", "Bus", "Light rail, streetcar, or trolley", "Subway or elevated", "Long-distance train or commuter train", "Taxicab", "Ferryboat",	"Bicycle", "Walked only", "Other", "Worked at home", "N/A"))) %>%
  select(tranwork_f, count_Mexican, prc2_Mexican, `count_Other Hispanics or Latinos` ,`prc2_Other Hispanics or Latinos`, `count_Non-Hispanic or Latino`, `prc2_Non-Hispanic or Latino` ) %>%

  # using kable
  kbl(booktabs = T,
      col.names = c("", "Number", "Percent",
                    "Number", "Percent",
                    "Number", "Percent"),
      align = c("l",rep("r",6)),
      caption = "Primary Means of Transportation to Work for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "") %>%
  kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position",
                ) %>%
  column_spec(1, width = "12em") %>%
  column_spec(2:7, width = "5.5em") %>%
  add_header_above( c("Means of\nTransportation", "Mexican" = 2, "Other Hispanics\nor Latinos" = 2, "Non-Hispanic\nor Latino" = 2),
                    # background = "#cbc9e2",
                    bold = T, font_size =8) %>%
  row_spec(0, 
           # bold = T, 
           align = "c"
           # background = "#f2f0f7",
           # font_size = 8
           ) %>%

  footnote(general = "\\\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}",
           general_title = "",
           threeparttable = T,
           escape = FALSE)

table_5.d
```




\clearpage

## 5D: Numbers and percentages of unionized Mexican workers

```{r}


cps_ddi <- read_ipums_micro(read_ipums_ddi("C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/cps_00172.xml")) %>%
  clean_names()


Union <- cps_ddi %>%
  filter( year == 2022) %>%  
  mutate(race_f = as_factor(race),
         hispan_f = as_factor(hispan),
         laborforce_f = as_factor(labforce),
         employment_f = as_factor(empstat),
         union_f = as_factor(union)) %>%

  mutate(race_ethnicity = case_when(hispan ==0 & race == 100 ~ "White (non-Hispanic or Latino)",
                                    hispan ==0 & race == 200 ~ "Black (non-Hispanic or Latino)",
                                    hispan %in% c(200:612) ~ "Other Hispanics or Latinos",
                                    hispan %in% c(100:109) ~ "Mexican",
                                    hispan ==0 & !race %in%c(100,200) ~ "Other (non-Hispanic or Latino)",
                                    TRUE ~ NA_character_)) %>%
  mutate(race_ethnicity = factor(race_ethnicity, level = c("Mexican", "Other Hispanics or Latinos","White (non-Hispanic or Latino)", "Black (non-Hispanic or Latino)",  "Other (non-Hispanic or Latino)"))) %>%

  mutate(employment_f = case_when(empstat %in% c(1, 10, 12) ~ "Employed",
                              TRUE ~ "other")) %>%


  filter(labforce != 0 |  empstat !=0) %>%
  filter(employment_f == "Employed") %>%
  as_survey_design(weights = earnwt) %>%

  survey_count(race_ethnicity, laborforce_f, employment_f,  union_f, name="count") %>%
  group_by(race_ethnicity) %>%
  mutate(total_employed = sum(count)) %>%
  filter(union_f %in% c("Member of labor union", "Covered by union but not a member")) %>%
  # group_by(race_ethnicity, total_employed) %>%
  mutate(prc = 100*round(count/total_employed, 4),
            prc2 = paste0(format(round(prc, 1), nsmall = 1), "%")) %>%
  select(race_ethnicity,union_f, total_employed,  count, prc2) %>%
  pivot_wider(names_from = union_f, values_from = c(count, prc2))



table5b <- Union %>%
  select(race_ethnicity,  total_employed,  `count_Member of labor union`, `prc2_Member of labor union`, `count_Covered by union but not a member`, `prc2_Covered by union but not a member`  ) %>%
  mutate(`count_Covered by union but not a member` = round(`count_Covered by union but not a member`, 0)) %>% 


  kbl(booktabs = T,
      col.names = c("",  "" ,"Number", "Percent", "Number", "Percent"),
      align = c("l",rep("r",5)),
      caption = "Union Membership and Coverage for Mexicans and Other Racial/Ethnic Groups in the U.S., 2022",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "5.16em") %>%
  add_header_above(align = c("l", rep("c",3)), c("Race/Ethnicity",  "Total\nEmployed","Member of\nLabor Union" =2 ,"Covered by\nUnion but Not\na Member" = 2),
                    # background = "#cbc9e2",
                    bold = T, font_size =8) %>%
  row_spec(0, bold = T, align = "c",
           # background = "#f2f0f7",
           font_size = 8) %>%
  add_footnote(c("\\footnotesize{Data Source: Data Source: IPUMS CPS, University of Minnesota, www.ipums.org. (2022 Current Population Survey). Tabulations by Great Cities Institute.}"),
notation = "none",
threeparttable = T,
escape = FALSE)

   # footnote(general = "\\\\footnotesize{Data Source: IPUMS CPS, University of Minnesota, www.ipums.org. (2022 Current Population Survey). Tabulations by Great Cities Institute.}",
   #         general_title = "",
   #         threeparttable = T,
   #         escape = FALSE)

 table5b

``` 


## 5E: Latino GDP, 2018

```{r}
latino_gdp <- data.frame(
  Region = c("Illinois Latino GDP (5th highest state)",
             "Chicago Metro Area (IL, IN, WI) Latino GDP (5th highest metro area)"),
  GDP_Billions_Dollars = c(100.1, 97.5)
)



latino_gdp_tb <-  latino_gdp %>%
  kbl(booktabs = T,
      col.names = c("",  "Latino GDP (billions of dollars)"),
      align = c("l",rep("r",1)),
      caption = "Illinois and Chicago Metro Area Latino GDP, 2018",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "20em") %>%
  column_spec(2, width = "10em") %>%
  # add_header_above(align = c("l", rep("c",3)), c("Race/Ethnicity",  "Total\nEmployed","Member of\nLabor Union" =2 ,"Covered by\nUnion but Not\na Member" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T, align = "c",
           # background = "#f2f0f7",
           font_size = 8) %>%
  footnote(general = "\\\\footnotesize{Data Source: 2023 U.S. Latino GDP Report, www.LatinoGDP.us (CLU-CERF, Bank of America State and Metro Latino GDP Reports).}",
           general_title = "",
           threeparttable = T,
           escape = FALSE)
  

latino_gdp_tb
```



```{r}
country_gdp <- data.frame(
  Country = c("United States", "China", "Japan", "Germany", "U.S. Latinos",
              "India", "United Kingdom", "France", "Italy", "Canada", "Russia"),
  GDP_Billions_Dollars = c(23315.1, 17759.3, 5005.5, 4262.8, 3159.7,
                           3150.3, 3123.2, 2957.4, 2115.8, 2001.5, 1836.6)
)

country_gdp_tb <- country_gdp %>% 
  kbl(booktabs = T,
      col.names = c("Country",  "GDP (billions of dollars)"),
      align = c("l",rep("r",1)),
      caption = "2021 Latino GDP with 10 Largest Countries",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "20em") %>%
  column_spec(2, width = "10em") %>%
  # add_header_above(align = c("l", rep("c",3)), c("Race/Ethnicity",  "Total\nEmployed","Member of\nLabor Union" =2 ,"Covered by\nUnion but Not\na Member" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T, align = "c",
           # background = "#f2f0f7",
           font_size = 8) %>%
  row_spec(5, bold = TRUE, background = "#f2f0f7" ) %>% 
  footnote(general = "\\\\footnotesize{Data Source: 2023 U.S. Latino GDP Report, www.LatinoGDP.us (International Monetary Fund and Organization for Economic Cooperation and Development).}",
           general_title = "",
           threeparttable = T,
           escape = FALSE)
  
country_gdp_tb
```

## 5F: Business Ownership

```{r}
business_ownership <- data.frame(
  Ownership = c("Hispanic", 
                "Not Hispanic", 
                "Owned equally by both groups", 
                "American Indian and Alaska Native (non-Hispanic)",
                "Asian (non-Hispanic)", 
                "Black or African American (non-Hispanic)", 
                "Native Hawaiian and Other Pacific Islander (non-Hispanic)", 
                "White (non-Hispanic)"),
  Without_Employees = c(124000, 708000, 1200, 3000, 80000, 146000, 700, 488000),
  With_Employees = c(15499, 173587, 1216, 441, 20290, 5541, 71, 147353),
  Total_Businesses = c(139499, 881587, 2416, 3441, 100290, 151541, 771, 635353),
  pop = c(
  2187650,  # Hispanic or Latino
  7379395,  # Not Hispanic or Latino
  NA,       # Owned equally by both groups: Not applicable
  7183,     # American Indian and Alaska Native (non-Hispanic)
  663560,   # Asian (non-Hispanic)
  1522030,  # Black or African American (non-Hispanic)
  2120,     # Native Hawaiian and Other Pacific Islander (non-Hispanic)
  4898082   # White (non-Hispanic)
),
Prc_Ownership_With_Employees = c(
    8.2,   # Hispanic
    NA,    # Not Hispanic: Percentage is not provided in the dataset
    NA,    # Owned equally by both groups: Not applicable
    0.2,   # American Indian and Alaska Native (non-Hispanic)
    10.7,  # Asian (non-Hispanic)
    2.9,   # Black or African American (non-Hispanic)
    0.0,   # Native Hawaiian and Other Pacific Islander (non-Hispanic)
    77.9   # White (non-Hispanic)
  )) %>% 
  mutate(total_pop = (pop[Ownership=="Hispanic"] +pop[Ownership=="Not Hispanic"])) %>% 
  group_by(Ownership) %>% 
  mutate(prc_pop = 100 *(pop/total_pop), 
         prc_pop2 = paste0(format(round(prc_pop, 1), nsmall = 1), "%")) %>% 
  mutate(Prc_Ownership_With_Employees2 = paste0(Prc_Ownership_With_Employees, "%"))
  

  


business_ownership_tb <- business_ownership %>% 
  select(Ownership, Without_Employees,With_Employees, Total_Businesses ) %>% 
  kbl(booktabs = T,
      col.names = c("Ownership",  "Without Employees", "With Employees", "Total Businesses"),
      align = c("l",rep("r",3)),
      caption = "Share of Business Ownership (2021) and Population by Race/Ethnicity (2018-2022 ACS 5-year Estimates) in Chicago−Naperville−Elgin, IL−IN−WI Metro Area",
      format.args = list(big.mark = ","),
      linesep = "") %>%
 kable_classic(full_width = F,
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>%
  column_spec(1, width = "19em") %>%
  column_spec(2, width = "7em") %>%
  # add_header_above(align = c("l", rep("c",3)), c("Race/Ethnicity",  "Total\nEmployed","Member of\nLabor Union" =2 ,"Covered by\nUnion but Not\na Member" = 2),
  #                   # background = "#cbc9e2",
  #                   bold = T, font_size =8) %>%
  row_spec(0, bold = T, align = "c",
           # background = "#f2f0f7",
           font_size = 8) %>%
  row_spec(3, hline_after = T, extra_latex_after = "%") %>% 
  add_footnote(c("\\footnotesize{Note: Counts include only businesses classifiable by owner demographic group}", "\\footnotesize{Data Sources: Annual Business Survey, 2021 (Census); Nonemployer Statistics by Demographics, 2021 (Census)
}"),
notation = "none",
threeparttable = T,
escape = FALSE)
  
business_ownership_tb

```


```{r fig.width=6, fig.height=4 ,fig.align='center'}

race_group = c(Prc_Ownership_With_Employees = "#4D2B18",
               prc_pop = "#D3A668")

business_ownership %>%
  filter(!Ownership%in%c("Owned equally by both groups", "Not Hispanic")) %>% 
  select( Ownership, prc_pop, Prc_Ownership_With_Employees ) %>% 
  pivot_longer(cols = c(prc_pop, Prc_Ownership_With_Employees), names_to = "group", values_to = "prc") %>% 

  ggplot(aes(x =  Ownership,
             y = prc,
             fill = group)) +
  geom_bar(position=position_dodge(width=0.7), stat = "identity",
           color = "white",
           width=0.7) +
  labs(y="", x = "",
       title="Figure 6: Share of Business Ownership (2021) and Population\nby Race/Ethnicity (2018-2022 ACS 5-year Estimates) in\nChicago−Naperville−Elgin, IL−IN−WI Metro Area",
       fill="") +
  scale_fill_manual(values = race_group,
                    labels = c("% of Business Ownership", "% of Population")) +
  geom_text(aes(label = paste0(format(round(prc, 1), nsmall = 1), "%")),
            position=position_dodge(width=0.7),
            vjust = -0.5,
            hjust = 0.5,
            size = 2.5) +

  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) + # Modify labels of ggplot2 barplot
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0,100), breaks = seq(0, 100, by = 100)) +

  theme_plots +
  theme(
    # axis.text.x = element_text(size = 8),
        panel.background = element_blank(),
        # axis.ticks.y = element_blank(),
        # axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        legend.position = "bottom",
        legend.text = element_text(size = 7, color = "#636363"),
        legend.key.size = unit(0.5, "lines"))
  # guides(fill = guide_legend(nrow = 2, byrow = TRUE))


```
