---
title: "Mexican Report"
author: "Health"
date: "Great Cities Institute"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    highlight: pygments
    extra_dependencies: ["float"]
    fig_caption: true
mainfont: "Arial"
header-includes: 
  - \usepackage{caption} 
  - \usepackage{booktabs} 
  - \usepackage{longtable}
  - \usepackage{threeparttablex} 
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F, fig.pos="H")


```


```{r}

library(ipumsr)
library(tidyverse)
library(purrr)
library(sf)
library(tidycensus)
library(tidyr)
library(readxl)
library(sf)
library(htmltools)
library(leaflet)
library(janitor)
library(data.table)
library(tinytex)
library(survey) # for survey analysis
library(srvyr)  # for tidy survey analysis
library(tmap)
library("kableExtra") 
library(tmaptools)
library("RColorBrewer") 
# library(OpenStreetMap)
# remotes::install_github('yihui/knitr')
# install.packages("knitr")
# library(knitr)
# options(tinytex.verbose = TRUE)
# tinytex::reinstall_tinytex()
# install.packages('tinytex')

# update.packages(ask = FALSE, checkBuilt = TRUE)
# tinytex::tlmgr_update()
# tinytex::latexmk("demographic_visual.tex")
options(knitr.kable.NA = '-')
##### Chicago boarder
Chicago_boundary_shp <- read_sf("C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/Boundaries_Chicago/geo_export_8881adc5-2460-457b-892f-04f4d6028b72.shp") %>% 
  st_transform(crs = 4326)
# mapview(community_shp)


### usa data

# Read Data

usa_data <- read.csv("C:/Users/elhamp2/Box/Great Cities Institute/Research/Mexican Report/final_Mexican_IL_2000_22.csv") %>% 
  mutate(race_ethnicity = case_when(hispan ==0 & race == 1 ~ "White (non-Hispanic or Latino)",
                                    hispan ==0 & race == 2 ~ "Black (non-Hispanic or Latino)",
                                    hispan %in% c(2,3,4) ~ "Other Hispanics or Latinos",
                                    hispan ==1 ~ "Mexican", 
                                    hispan ==0 & race %in%c(3:9) ~ "Other (non-Hispanic or Latino)",
                                    TRUE ~ NA_character_),
         
  race_ethnicity = factor(race_ethnicity, level = c("Mexican", "Other Hispanics or Latinos","White (non-Hispanic or Latino)", "Black (non-Hispanic or Latino)",  "Other (non-Hispanic or Latino)")),
  
  hispan_breakdown = case_when(
  hispand == 0 ~ "Not Hispanic",
  hispand == 100 ~ "Mexican", 
  hispand == 200 ~ "Puerto Rican",
  hispand == 300 ~ "Cuban",
  hispand == 411 ~ "Costa Rican",
  hispand == 412 ~ "Guatemalan",
  hispand == 413 ~ "Honduran",
  hispand == 414 ~ "Nicaraguan",
  hispand == 415 ~ "Panamanian",
  hispand == 416 ~ "Salvadoran",
  hispand == 417 ~ "Central American, not specified",
  hispand == 420 ~ "Argentinean",
  hispand == 421 ~ "Bolivian",
  hispand == 422 ~ "Chilean",
  hispand == 423 ~ "Colombian",
  hispand == 424 ~ "Ecuadorian",
  hispand == 425 ~ "Paraguayan",
  hispand == 426 ~ "Peruvian",
  hispand == 427 ~ "Uruguayan",
  hispand == 428 ~ "Venezuelan",
  hispand == 431 ~ "South American, not specified",
  hispand == 450 ~ "Spaniard",
  hispand == 460 ~ "Dominican",
  hispand == 498 ~ "Other, not specified",
  TRUE ~ "Other"
),
mex_foreign_born = case_when(
    race_ethnicity == "Mexican" & bpl %in% as.character(1:56) ~ "U.S. Born",
    race_ethnicity == "Mexican" & bpl == "200" ~ "Mexico Born",
    TRUE ~ NA_character_  # Optional: default case for other values or groups
  )) %>% ## filter for hispanic groups here ##
 mutate(hispan_breakdown = factor(hispan_breakdown, level = c("Mexican", "Puerto Rican","Ecuadorian", "Cuban", "Guatemalan", "Colombian"))
  )

# Construct the full file path
chicago_ca_pop <- read_sf("Data Tables/shp/comm_pop_chi_2022.shp") %>% as.data.frame()

chicago_health <- read_csv("Health Data/Chicago Health Atlas Data Download - Community areas.csv") %>% clean_names() %>% rename(
  obesity_rate = hcsobp_2022_2023,
  diabetes_rate = hcsdiap_2022_2023,
  asthma_rate = hcsathp_2022_2023,
  lbw_rate = vrbwp_2017_2021,
  unmet_mental_health_rate = hcsumhap_2022_2023,
  cancer_diagnosis_rate = ccr_2017_2021,
  hyper_tension_rate = hcshytp_2022_2023,
  psych_distress_rate =   hcsspdp_2022_2023,
  lung_cancer_rate =ccg_2017_2021,
  loneliness_rate = chavqos_2022_2023
)


# merge chicago ca with health atlas
df_health <- chicago_ca_pop %>% select(commnty, prc_mxc, prc_nh_w, prc_nh_b) %>% mutate(commnty = str_to_title(tolower(commnty))) %>% left_join(chicago_health, by = c("commnty" = "name")) %>% select(-layer)  %>%
 mutate(across(matches("rate$"), ~ format(round(as.numeric(.), 1), nsmall = 1))) %>% mutate(across(ends_with("rate"), ~ as.numeric(as.character(.))))



```




\renewcommand{\arraystretch}{1.4}
\clearpage





# 4. Health 

## 4A: Health Outcomes in Top 4 Community Areas with the Largest Mexican Population Share, (2017-2021 and 2022-2023):



```{r}
# analysis
mexican_top <- df_health %>% arrange(desc(prc_mxc)) %>% top_n(4, prc_mxc) %>% select(commnty, obesity_rate, diabetes_rate, asthma_rate, lbw_rate)   %>%
mutate(
    obesity_rate = paste0(format(round(obesity_rate, 1), nsmall = 1), "%"),
    diabetes_rate = paste0(format(round(diabetes_rate, 1), nsmall = 1), "%"),
    asthma_rate = paste0(format(round(asthma_rate, 1), nsmall = 1), "%"),
    lbw_rate = paste0(format(round(lbw_rate, 1), nsmall = 1), "%")
)


df_white_city <- data.frame(
  commnty = "White Chicago Mean",
 obesity_rate = 24.5,
  diabetes_rate = 6.9, 
  asthma_rate = 9.1,
  lbw_rate = 6.2
)

df_black_city <- data.frame(
  commnty = "Black Chicago Mean",
  obesity_rate = 47.6,
  diabetes_rate = 19.6,
  asthma_rate = 14.9, 
  lbw_rate  =14.6
)

# Create a data frame
df_latino_city <- data.frame(
  commnty = "Hispanic or Latino Chicago Mean",
  obesity_rate = 39.9,
  diabetes_rate = 13.9, 
  asthma_rate = 9.7, 
  lbw_rate = 8.0
)

df_city_avgs <- rbind(df_white_city, df_black_city, df_latino_city)  %>%
mutate(
    obesity_rate = paste0(format(round(obesity_rate, 1), nsmall = 1), "%"),
    diabetes_rate = paste0(format(round(diabetes_rate, 1), nsmall = 1), "%"),
    asthma_rate = paste0(format(round(asthma_rate, 1), nsmall = 1), "%"),
    lbw_rate = paste0(format(round(lbw_rate, 1), nsmall = 1), "%")
)

df_final <- rbind(mexican_top, df_city_avgs) %>%  mutate(across(
    ends_with("Rate"),
    ~ ifelse(. == "  NA%", "-", .)
  ))


```


```{r}

# visuals
table_4a <- df_final %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Four Communities with the Largest Share of Mexican Population", "Adult Obesity Rate(2022-2023)", "Adult Diabetes Rate(2022-2023)", "Adult Asthma Rate(2022-2023)", "Low Birthweight Rate(2017-2021)"),
      align = c("l",rep("r",4)),
      caption = "Health Outcomes for Four Communities with the Largest Share of Mexican Population and by Race/Ethnicity in Chicago",
      format.args = list(big.mark = "%"),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "6.45em") %>%  
row_spec(0, bold = TRUE, font_size = 8, align = "c") %>%
  row_spec(5:7, bold = TRUE, background = "#f2f0f7") %>%
  row_spec(c(4), hline_after = T) %>%  
  
   add_footnote(
  c(
      "\\footnotesize{Note: Obesity is defined as percent of adults who reported a height and weight that yield a body mass index of 30 or greater.}",
    "\\footnotesize{Diabetes is defined as percent of adults who reported that a doctor, nurse or other health professional has diagnosed them with diabetes.}",
    "\\footnotesize{Asthma is defined as percent of adults who reported that a doctor, nurse or other health professional has diagnosed them with asthma.}",
    "\\footnotesize{Low birthweight is defined as percent of births with a birthweight less than 2500 grams (5.5 pounds).}",
    "\\footnotesize{Data Source:Chicago Health Atlas, Chicago Department of Public Health (2022–2023) and Illinois Department of Public Health (2017-2021). Tabulations by Great Cities Institute}"),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)


table_4a

```


## 4B: No Health Insurance for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
# analysis
## select variables, filter for year 2022
df <- usa_data %>%  filter(year == 2022) %>% select(race_ethnicity, hcovany, perwt)

# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)
  
  
# Calculate numerator: weighted count of individuals without health insurance by race_ethnicity
numerator_without_insurance <- survey_design %>%
  filter(hcovany == 1) %>% 
  group_by(race_ethnicity) %>%
  summarise(weighted_n_without_insurance = survey_total(vartype = "se"))

# Calculate denominator: total population by race_ethnicity
denominator <- survey_design %>%
  group_by(race_ethnicity) %>%
  summarise(total_pop = survey_total(vartype = "se")) 
  
# calculate rate

# Join numerator and denominator
df_rates <- numerator_without_insurance %>%
  left_join(denominator, by = "race_ethnicity") %>%
  mutate(
    no_insurance_rate = (weighted_n_without_insurance / total_pop) * 100,
    no_insurance_rate = paste0(format(round(no_insurance_rate, 1), nsmall = 1), "%"), 
      unininsured_sum = sum(weighted_n_without_insurance),
    no_insurance_rate_2 = (weighted_n_without_insurance / unininsured_sum) * 100,
    no_insurance_rate_2 = paste0(format(round(no_insurance_rate_2, 1), nsmall = 1), "%")

  )
  

```


```{r}
# visuals
table_4b <- df_rates  %>% select(race_ethnicity, weighted_n_without_insurance, no_insurance_rate, no_insurance_rate_2)   %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number Without Health Insurance", "No Insurance Rate", "Percent Share of the Uninsured Population"),
      align = c("l",rep("r",3)),
      caption = "No Health Insurance for Mexicans and Other Racial/Ethnic Groups in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "8.6em") %>%  

  row_spec(0, bold = T,align = "c",
             #background = "#f2f0f7",
            font_size = 8) %>%

  add_footnote(
    c("\\footnotesize{Data Source: IPUMS USA, University of Minnesota, www.ipums.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute.}"),
    notation = "none",
    threeparttable = TRUE,
    escape = FALSE
  )


table_4b

```


## 4C: Number and Percent of Mexico Born and U.S. Born Mexicans Without Job Provided Health Insurance in Chicago, 2018-2022 (ACS 5-year Estimates)

```{r}
## select variables, filter for year 2022
df <- usa_data %>%  filter(year == 2022) %>% select(mex_foreign_born, hinsemp, perwt) %>% filter(!is.na(mex_foreign_born))

# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)
  
  
# Calculate numerator: weighted count of individuals without job provided health insurance by race_ethnicity
numerator_without_insurance <- survey_design %>%
  filter(hinsemp == 1) %>% 
  group_by(mex_foreign_born) %>%
  summarise(weighted_n_without_insurance = survey_total(vartype = "se"))

# Calculate denominator: total population by race_ethnicity
denominator <- survey_design %>%
  group_by(mex_foreign_born) %>%
  summarise(total_pop = survey_total(vartype = "se")) 
  
# calculate rate

# Join numerator and denominator
df_rates <- numerator_without_insurance %>%
  left_join(denominator, by = "mex_foreign_born") %>%
  mutate(
    no_insurance_rate = (weighted_n_without_insurance / total_pop) * 100,
    no_insurance_rate = paste0(format(round(no_insurance_rate, 1), nsmall = 1), "%")


  )
  
```

```{r}
# visuals
table_4c <- df_rates  %>% select(mex_foreign_born, weighted_n_without_insurance, no_insurance_rate)   %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Mexican Nativity", "Number Without Job Provided Health Insurance", "No Job Provided Health Insurance"),
      align = c("l",rep("r",2)),
      caption = "Number and Percent of Mexico Born and U.S. Born Mexicans Without Job Provided Health Insurance in Chicago, 2018-2022 (ACS 5-year Estimates)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "7.9em") %>%  

  row_spec(0, bold = T, align = "c",
             #background = "#f2f0f7",
            font_size = 8) %>%
  add_footnote(
  c("\\footnotesize{Data Source: IPUMS NHGIS, University of Minnesota, www.nhgis.org (2018-2022 American Community Survey 5-year Estimates). Tabulations by Great Cities Institute. }"),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)


table_4c

```


## 4E: Mental Health Outcomes for Four Communities with the Largest Share of Mexican Population and by Race/Ethnicity in Chicago, 2022-2023

```{r}
# analysis
mexican_top <- df_health %>% arrange(desc(prc_mxc)) %>% top_n(4, prc_mxc) %>% select(commnty, psych_distress_rate, unmet_mental_health_rate, loneliness_rate)   %>%
 mutate(
    psych_distress_rate = paste0(round(psych_distress_rate, 1), "%"),
    unmet_mental_health_rate = paste0(round(unmet_mental_health_rate, 1), "%"),
    loneliness_rate = paste0(round(loneliness_rate, 1), "%")
)



df_white_city <- data.frame(
  commnty = "White Chicago Mean",
 psych_distress_rate = 7.5,
  unmet_mental_health_rate = 57.2,
  loneliness_rate = 26.5
)

df_black_city <- data.frame(
  commnty = "Black Chicago Mean",
  psych_distress_rate = 9.9,
  unmet_mental_health_rate = 81.4,
  loneliness_rate = 30.2
)

# Create a data frame
df_latino_city <- data.frame(
  commnty = "Hispanic or Latino Chicago Mean",
  psych_distress_rate = 14.9,
  unmet_mental_health_rate = 80.0,
  loneliness_rate = 29.9
)

df_city_avgs <- rbind(df_white_city, df_black_city, df_latino_city)  %>%
  mutate(
    psych_distress_rate = paste0(sprintf("%.1f", psych_distress_rate), "%"),
    unmet_mental_health_rate = paste0(sprintf("%.1f", unmet_mental_health_rate), "%"),
    loneliness_rate = paste0(sprintf("%.1f", loneliness_rate), "%")
  )


df_final <- rbind(mexican_top, df_city_avgs)
```


```{r}
# visuals
table_4e <- df_final %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Four Communities with the Largest Share of Mexican Population", "Adult Serious Psych Distress Rate(2022-2023)", "Adult Unmet Mental Health Need Rate(2022-2023)", "Adult Loneliness Rate(2022-2023)"),
      align = c("l",rep("r",3)),
      caption = "Mental Health Outcomes for Four Communities with the Largest Share of Mexican Population and by Race/Ethnicity in Chicago, 2022-2023",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "9.2em") %>%  
row_spec(0, bold = TRUE, font_size = 8, align = "c") %>%
  row_spec(5:7, bold = TRUE, background = "#f2f0f7") %>%
   row_spec(c(4), hline_after = T) %>%  
  
  add_footnote(
  c( 
      "\\footnotesize{Note: Serious Psychological Distress is the percent of adults who experienced feelings of nervousness, hopelessness, and depression, in the past 30 days.}",
    "\\footnotesize{Unmet mental health need is the percent of adults with psychological distress who are not receiving treatment or medication.}",
    "\\footnotesize{Loneliness rate is defined as the percent of adults who reported feeling left out or felt alone.}",
    "\\footnotesize{Data Source: Chicago Health Atlas, Chicago Department of Public Health (2022–2023) and Illinois Department of Public Health (2017-2021).Tabulations by Great Cities Institute.}"
    
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)



table_4e



```


## 4F: Chronic Health Indicators in top 4 Mexican Community Areas,(2017-2021, 2022-2023):

```{r}
# Analysis
mexican_top <- df_health %>% 
  arrange(desc(prc_mxc)) %>% 
  top_n(4, prc_mxc) %>% 
  select(commnty, cancer_diagnosis_rate, hyper_tension_rate, lung_cancer_rate) %>% 
mutate(
    cancer_diagnosis_rate = format(round(cancer_diagnosis_rate, 1), nsmall = 1),
    hyper_tension_rate = paste0(format(round(hyper_tension_rate, 1), nsmall = 1), "%"),
    lung_cancer_rate = format(round(lung_cancer_rate, 1), nsmall = 1)
)

df_white_city <- data.frame(
  commnty = "White Chicago Mean",
  cancer_diagnosis_rate = 676.16, 
  hyper_tension_rate = 29.8, 
  lung_cancer_rate = 83.41
)

df_black_city <- data.frame(
  commnty = "Black Chicago Mean",
  cancer_diagnosis_rate = 529.94, 
  hyper_tension_rate = 43.9, 
  lung_cancer_rate = 77.39
)

df_latino_city <- data.frame(
  commnty = "Hispanic or Latino Chicago Mean",
  cancer_diagnosis_rate = 253.29, 
  hyper_tension_rate = 25.7, 
  lung_cancer_rate = 13.93
)

df_city_avgs <- rbind(df_white_city, df_black_city, df_latino_city) %>% 
 mutate(
    cancer_diagnosis_rate = format(round(cancer_diagnosis_rate, 1), nsmall = 1),
    hyper_tension_rate = paste0(format(round(hyper_tension_rate, 1), nsmall = 1), "%"),
    lung_cancer_rate = format(round(lung_cancer_rate, 1), nsmall = 1)
)


df_final <- rbind(mexican_top, df_city_avgs)

```



```{r}
# visuals
table_4f <- df_final %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Top 4 Community Areas with the Largest Mexican Population Share", "Adult Cancer Diagnosis Rate per 100,000 (2017-2021)", "Adult Hypertension Rate (2022-2023)", "Adult Lung Cancer Diagnosis Rate per 100,000 (2017-2021)"),
      align = c("l",rep("r",3)),
      caption = "Chronic Health Indicators for Four Communities with the Largest Share of Mexican Population and by Race/Ethnicity in Chicago",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "9.2em") %>%  
row_spec(0, bold = TRUE, font_size = 8, align = "c") %>%
  row_spec(c(4), hline_after = T) %>%  
  row_spec(5:7, bold = TRUE, background = "#f2f0f7") %>%   
  
    add_footnote(
  c("\\footnotesize{Note: Cancer diagnosis is defined as the annual diagnosis rate for all invasive cancers, excluding pre-cancerous conditions.  All ages, risk-adjusted.}",
    "\\footnotesize{Hypertension is the percentage of adults diagnosed with high blood pressure, excluding borderline or pregnancy-related cases.}",
    "\\footnotesize{Lung cancer is defined as lung and bronchus cancer for ages 15 and over, risk-adjusted.}", "\\footnotesize{Data Source: Chicago Health Atlas, Chicago Department of Public Health (2022–2023) and Illinois Department of Public Health (2017-2021).Tabulations by Great Cities Institute.}"
    
  ),
  notation = "none",
  threeparttable = TRUE,
  escape = FALSE
)
  

table_4f
```
