---
title: "Mexican Report"
author: "Health"
date: "Great Cities Institute"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
    highlight: pygments
    extra_dependencies: ["float"]
    fig_caption: true
mainfont: "Arial"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F, fig.pos="H")


```


```{r}

library(ipumsr)
library(tidyverse)
library(purrr)
library(sf)
library(tidycensus)
library(tidyr)
library(readxl)
library(sf)
library(htmltools)
library(leaflet)
library(janitor)
library(data.table)
library(tinytex)
library(survey) # for survey analysis
library(srvyr)  # for tidy survey analysis
library(tmap)
library("kableExtra") 
library(tmaptools)
library("RColorBrewer") 
# library(OpenStreetMap)
# remotes::install_github('yihui/knitr')
# install.packages("knitr")
# library(knitr)
# options(tinytex.verbose = TRUE)
# tinytex::reinstall_tinytex()
# install.packages('tinytex')

# update.packages(ask = FALSE, checkBuilt = TRUE)
# tinytex::tlmgr_update()
# tinytex::latexmk("demographic_visual.tex")
options(knitr.kable.NA = '-')
##### Chicago boarder
Chicago_boundary_shp <- read_sf("C:/Users/dsegovi2/Box/Great Cities Institute/Research/Mexican Report/Boundaries_Chicago/geo_export_8881adc5-2460-457b-892f-04f4d6028b72.shp") %>% 
  st_transform(crs = 4326)
# mapview(community_shp)


### usa data

# Read Data

usa_data <- read.csv("C:/Users/dsegovi2/Box/Great Cities Institute/Research/Mexican Report/final_Mexican_IL_2000_22.csv") %>% 
  mutate(race_ethnicity = case_when(hispan ==0 & race == 1 ~ "White (non-Hispanic or Latino)",
                                    hispan ==0 & race == 2 ~ "Black (non-Hispanic or Latino)",
                                    hispan %in% c(2,3,4) ~ "Other Latinos",
                                    hispan ==1 ~ "Mexican", 
                                    hispan ==0 & race %in%c(3:9) ~ "Other (non-Hispanic or Latino)",
                                    TRUE ~ NA_character_),
         
  race_ethnicity = factor(race_ethnicity, level = c("Mexican", "Other Latinos","White (non-Hispanic or Latino)", "Black (non-Hispanic or Latino)",  "Other (non-Hispanic or Latino)")),
  
  hispan_breakdown = case_when(
  hispand == 0 ~ "Not Hispanic",
  hispand == 100 ~ "Mexican", 
  hispand == 200 ~ "Puerto Rican",
  hispand == 300 ~ "Cuban",
  hispand == 411 ~ "Costa Rican",
  hispand == 412 ~ "Guatemalan",
  hispand == 413 ~ "Honduran",
  hispand == 414 ~ "Nicaraguan",
  hispand == 415 ~ "Panamanian",
  hispand == 416 ~ "Salvadoran",
  hispand == 417 ~ "Central American, not specified",
  hispand == 420 ~ "Argentinean",
  hispand == 421 ~ "Bolivian",
  hispand == 422 ~ "Chilean",
  hispand == 423 ~ "Colombian",
  hispand == 424 ~ "Ecuadorian",
  hispand == 425 ~ "Paraguayan",
  hispand == 426 ~ "Peruvian",
  hispand == 427 ~ "Uruguayan",
  hispand == 428 ~ "Venezuelan",
  hispand == 431 ~ "South American, not specified",
  hispand == 450 ~ "Spaniard",
  hispand == 460 ~ "Dominican",
  hispand == 498 ~ "Other, not specified",
  TRUE ~ "Other"
),
mex_foreign_born = case_when(
    race_ethnicity == "Mexican" & bpl %in% as.character(1:56) ~ "U.S. Born",
    race_ethnicity == "Mexican" & bpl == "200" ~ "Mexico Born",
    TRUE ~ NA_character_  # Optional: default case for other values or groups
  )) %>% ## filter for hispanic groups here ##
 mutate(hispan_breakdown = factor(hispan_breakdown, level = c("Mexican", "Puerto Rican","Ecuadorian", "Cuban", "Guatemalan", "Colombian"))
  )

# Construct the full file path
chicago_ca_pop <- read_sf("Data Tables/shp/comm_pop_chi_2022.shp") %>% as.data.frame()

chicago_health <- read_csv("Health Data/Chicago Health Atlas Data Download - Community areas.csv") %>% clean_names() %>% rename(
  obesity_rate = hcsobp_2022_2023,
  diabetes_rate = hcsdiap_2022_2023,
  asthma_rate = hcsathp_2022_2023,
  lbw_rate = vrbwp_2017_2021,
  unmet_mental_health_rate = hcsumhap_2022_2023,
  cancer_diagnosis_rate = ccr_2017_2021,
  hyper_tension_rate = hcshytp_2022_2023,
  psych_distress_rate =   hcsspdp_2022_2023,
  lung_cancer_rate =ccg_2017_2021,
  loneliness_rate = chavqos_2022_2023
)


# merge chicago ca with health atlas
df_health <- chicago_ca_pop %>% select(commnty, prc_mxc, prc_nh_w, prc_nh_b) %>% mutate(commnty = str_to_title(tolower(commnty))) %>% left_join(chicago_health, by = c("commnty" = "name")) %>% select(-layer)  %>%
 mutate(across(matches("rate$"), ~ format(round(as.numeric(.), 1), nsmall = 1))) %>% mutate(across(ends_with("rate"), ~ as.numeric(as.character(.))))



```





\renewcommand{\arraystretch}{1.4}
\clearpage



# 4. Health 

## 4A: Health Outcomes in Top 4 Community Areas with the Largest Mexican Population Share, (2017-2021 and 2022-2023):



```{r}
# analysis
mexican_top <- df_health %>% arrange(desc(prc_mxc)) %>% top_n(4, prc_mxc) %>% select(commnty, obesity_rate, diabetes_rate, asthma_rate, lbw_rate)   %>%
mutate(
    obesity_rate = paste0(format(round(obesity_rate, 1), nsmall = 1), "%"),
    diabetes_rate = paste0(format(round(diabetes_rate, 1), nsmall = 1), "%"),
    asthma_rate = paste0(format(round(asthma_rate, 1), nsmall = 1), "%"),
    lbw_rate = paste0(format(round(lbw_rate, 1), nsmall = 1), "%")
)


df_white_city <- data.frame(
  commnty = "White City Mean",
 obesity_rate = 24.5,
  diabetes_rate = 6.9, 
  asthma_rate = 9.1,
  lbw_rate = 6.2
)

df_black_city <- data.frame(
  commnty = "Black City Mean",
  obesity_rate = 47.6,
  diabetes_rate = 19.6,
  asthma_rate = 14.9, 
  lbw_rate  =14.6
)

# Create a data frame
df_latino_city <- data.frame(
  commnty = "Hispanic/Latino City Mean",
  obesity_rate = 39.9,
  diabetes_rate = 13.9, 
  asthma_rate = 9.7, 
  lbw_rate = 8.0
)

df_city_avgs <- rbind(df_white_city, df_black_city, df_latino_city)  %>%
mutate(
    obesity_rate = paste0(format(round(obesity_rate, 1), nsmall = 1), "%"),
    diabetes_rate = paste0(format(round(diabetes_rate, 1), nsmall = 1), "%"),
    asthma_rate = paste0(format(round(asthma_rate, 1), nsmall = 1), "%"),
    lbw_rate = paste0(format(round(lbw_rate, 1), nsmall = 1), "%")
)

df_final <- rbind(mexican_top, df_city_avgs) %>%  mutate(across(
    ends_with("Rate"),
    ~ ifelse(. == "NA%", "-", .)
  ))

```


```{r}

# visuals
table_4a <- df_final %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Top 4 Community Areas with the Largest Mexican Population Share", "Adult Obesity Rate(22-23)", "Adult Diabetes Rate(22-23)", "Adult Asthma Rate(22-23)", "Low Birthweight Rate(17-21)"),
      align = c("l",rep("c",4)),
      caption = "Health Outcomes, (2017-2021 and 2022-2023)",
      format.args = list(big.mark = "%"),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:5, width = "6.45em") %>%  
row_spec(0, bold = TRUE, font_size = 8) %>%
  row_spec(5:7, bold = TRUE, background = "#f2f0f7") %>%
  
    add_footnote(c("Obesity is defined as percent of adults who reported a height and weight that yield a body mass index of 30 or greater.", "Diabetes is defined as percent of adults who reported that a doctor, nurse or other health professional has diagnosed them with diabetes.", "Asthma is defined as percent of adults who reported that a doctor, nurse or other health professional has diagnosed them with asthma.", "Low birthweight is defined as percent of births with a birthweight less than 2500 grams (5.5 pounds).",     "Source: Chicago Health Atlas, Chicago Department of Public Health (2022–2023) and Illinois Department of Public Health (2017-2021).", "Tabulations by Great Cities Institute"))


table_4a

```




## 4B: Number and Percentages without Health Insurance by Race/Ethnicity, 2018-2022:


```{r}
# analysis
## select variables, filter for year 2022
df <- usa_data %>%  filter(year == 2022) %>% select(race_ethnicity, hcovany, perwt)

# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)
  
  
# Calculate numerator: weighted count of individuals without health insurance by race_ethnicity
numerator_without_insurance <- survey_design %>%
  filter(hcovany == 1) %>% 
  group_by(race_ethnicity) %>%
  summarise(weighted_n_without_insurance = survey_total(vartype = "se"))

# Calculate denominator: total population by race_ethnicity
denominator <- survey_design %>%
  group_by(race_ethnicity) %>%
  summarise(total_pop = survey_total(vartype = "se")) 
  
# calculate rate

# Join numerator and denominator
df_rates <- numerator_without_insurance %>%
  left_join(denominator, by = "race_ethnicity") %>%
  mutate(
    no_insurance_rate = (weighted_n_without_insurance / total_pop) * 100,
    no_insurance_rate = paste0(format(round(no_insurance_rate, 1), nsmall = 1), "%"), 
      unininsured_sum = sum(weighted_n_without_insurance),
    no_insurance_rate_2 = (weighted_n_without_insurance / unininsured_sum) * 100,
    no_insurance_rate_2 = paste0(format(round(no_insurance_rate_2, 1), nsmall = 1), "%")

  )
  

```


```{r}
# visuals
table_4b <- df_rates  %>% select(race_ethnicity, weighted_n_without_insurance, no_insurance_rate, no_insurance_rate_2)   %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Race/Ethnicity", "Number without health insurance", "No Insurance Rate", "Percentage of Uninsured Individuals Without Health Insurance"),
      align = c("l",rep("c",3)),
      caption = "No Health Insurance by Race/Ethnicity, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "8.6em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

  footnote(general = "Source: IPUMS-USA database (2018-2022 American Community Survey 5-Year Estimates). Tabulations by Great Cities Institute",
           general_title = "")   


table_4b

```



## 4C: Number and Percentages of Mexicans born in U.S. versus Mexicans born in Mexico without job provided Health Insurance, 2018-2022:

```{r}
## select variables, filter for year 2022
df <- usa_data %>%  filter(year == 2022) %>% select(mex_foreign_born, hinsemp, perwt) %>% filter(!is.na(mex_foreign_born))

# Create survey design object
survey_design <- df %>%
  as_survey_design(weights = perwt)
  
  
# Calculate numerator: weighted count of individuals without job provided health insurance by race_ethnicity
numerator_without_insurance <- survey_design %>%
  filter(hinsemp == 1) %>% 
  group_by(mex_foreign_born) %>%
  summarise(weighted_n_without_insurance = survey_total(vartype = "se"))

# Calculate denominator: total population by race_ethnicity
denominator <- survey_design %>%
  group_by(mex_foreign_born) %>%
  summarise(total_pop = survey_total(vartype = "se")) 
  
# calculate rate

# Join numerator and denominator
df_rates <- numerator_without_insurance %>%
  left_join(denominator, by = "mex_foreign_born") %>%
  mutate(
    no_insurance_rate = (weighted_n_without_insurance / total_pop) * 100,
    no_insurance_rate = paste0(format(round(no_insurance_rate, 1), nsmall = 1), "%")


  )
  
```

```{r}
# visuals
table_4b.2 <- df_rates  %>% select(mex_foreign_born, weighted_n_without_insurance, no_insurance_rate)   %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Mexican Nativity", "Number without Job provided health insurance", "No Job provided health Insurance"),
      align = c("l",rep("c",3)),
      caption = "No Health Insurance, 2018-2022",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:3, width = "8.6em") %>%  

  row_spec(0, bold = T,
             #background = "#f2f0f7",
            font_size = 8) %>%

  footnote(general = "Source: IPUMS-USA database (2018-2022 American Community Survey 5-Year Estimates). Tabulations by Great Cities Institute",
           general_title = "")   


table_4b.2

```


## 4E: Mental Health  in Top 4 Community Areas with the Largest Mexican Population Share, 2022-2023:


```{r}
# analysis
mexican_top <- df_health %>% arrange(desc(prc_mxc)) %>% top_n(4, prc_mxc) %>% select(commnty, psych_distress_rate, unmet_mental_health_rate, loneliness_rate)   %>%
 mutate(
    psych_distress_rate = paste0(round(psych_distress_rate, 1), "%"),
    unmet_mental_health_rate = paste0(round(unmet_mental_health_rate, 1), "%"),
    loneliness_rate = paste0(round(loneliness_rate, 1), "%")
)



df_white_city <- data.frame(
  commnty = "White City Mean",
 psych_distress_rate = 7.5,
  unmet_mental_health_rate = 57.2,
  loneliness_rate = 26.5
)

df_black_city <- data.frame(
  commnty = "Black City Mean",
  psych_distress_rate = 9.9,
  unmet_mental_health_rate = 81.4,
  loneliness_rate = 30.2
)

# Create a data frame
df_latino_city <- data.frame(
  commnty = "Hispanic/Latino City Mean",
  psych_distress_rate = 14.9,
  unmet_mental_health_rate = 80.0,
  loneliness_rate = 29.9
)

df_city_avgs <- rbind(df_white_city, df_black_city, df_latino_city)  %>%
  mutate(
    psych_distress_rate = paste0(sprintf("%.1f", psych_distress_rate), "%"),
    unmet_mental_health_rate = paste0(sprintf("%.1f", unmet_mental_health_rate), "%"),
    loneliness_rate = paste0(sprintf("%.1f", loneliness_rate), "%")
  )


df_final <- rbind(mexican_top, df_city_avgs)
```


```{r}
# visuals
table_4e <- df_final %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Top 4 Community Areas with the Largest Mexican Population Share", "Adult Serious Psych Distress Rate(22-23)", "Adult Unmet Mental Health Need Rate(22-23)", "Adult Loneliness Rate(22-23)"),
      align = c("l",rep("c",3)),
      caption = "Mental Health, 2022-2023",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "9.2em") %>%  
row_spec(0, bold = TRUE, font_size = 8) %>%
  row_spec(5:7, bold = TRUE, background = "#f2f0f7") %>%

    add_footnote(c("Serious Psychological Distress is the percent of adults who experienced feelings of nervousness, hopelessness, and depression, in the past 30 days.", "Unmet mental health need is the percent of adults with psychological distress who are not receiving treatment or medication.", "Loneliness rate is defined as the percent of adults who reported feeling left out or felt alone.", "Source:  Chicago Health Atlas, Chicago Department of Public Health (2022-2023). Tabulations by Great Cities Institute"))



table_4e



```


## 4F: Chronic Health Indicators in top 4 Mexican Community Areas,(2017-2021, 2022-2023):

```{r}
# Analysis
mexican_top <- df_health %>% 
  arrange(desc(prc_mxc)) %>% 
  top_n(4, prc_mxc) %>% 
  select(commnty, cancer_diagnosis_rate, hyper_tension_rate, lung_cancer_rate) %>% 
mutate(
    cancer_diagnosis_rate = format(round(cancer_diagnosis_rate, 1), nsmall = 1),
    hyper_tension_rate = paste0(format(round(hyper_tension_rate, 1), nsmall = 1), "%"),
    lung_cancer_rate = format(round(lung_cancer_rate, 1), nsmall = 1)
)

df_white_city <- data.frame(
  commnty = "White City Mean",
  cancer_diagnosis_rate = 676.16, 
  hyper_tension_rate = 29.8, 
  lung_cancer_rate = 83.41
)

df_black_city <- data.frame(
  commnty = "Black City Mean",
  cancer_diagnosis_rate = 529.94, 
  hyper_tension_rate = 43.9, 
  lung_cancer_rate = 77.39
)

df_latino_city <- data.frame(
  commnty = "Hispanic/Latino City Mean",
  cancer_diagnosis_rate = 253.29, 
  hyper_tension_rate = 25.7, 
  lung_cancer_rate = 13.93
)

df_city_avgs <- rbind(df_white_city, df_black_city, df_latino_city) %>% 
 mutate(
    cancer_diagnosis_rate = format(round(cancer_diagnosis_rate, 1), nsmall = 1),
    hyper_tension_rate = paste0(format(round(hyper_tension_rate, 1), nsmall = 1), "%"),
    lung_cancer_rate = format(round(lung_cancer_rate, 1), nsmall = 1)
)


df_final <- rbind(mexican_top, df_city_avgs)

```



```{r}
# visuals
table_4f <- df_final %>% 
 
   # using kable 
  
  kbl(, booktabs = T,
      col.names = c("Top 4 Community Areas with the Largest Mexican Population Share", "Adult Cancer Diagnosis Rate per 100,000 (17-21)", "Adult Hypertension Rate (22-23)", "Adult Lung Cancer Diagnosis Rate per 100,000 (17-21)"),
      align = c("l",rep("c",3)),
      caption = "Chronic Health Indicators, (2017-2021, 2022-2023)",
      format.args = list(big.mark = ","),
      linesep = "")  %>% 
  kable_classic(full_width = F,  
                font_size = 8,
                html_font = "Arial",
                position = "center",
                latex_options = "HOLD_position") %>% 
  column_spec(1, width = "14.2em") %>%
  column_spec(2:4, width = "9.2em") %>%  
row_spec(0, bold = TRUE, font_size = 8) %>%
  row_spec(5:7, bold = TRUE, background = "#f2f0f7") %>%     add_footnote(c(
  "Cancer diagnosis is defined as the annual diagnosis rate for all invasive cancers, excluding pre-cancerous conditions.  All ages, risk-adjusted.",
  "Hypertension is the percentage of adults diagnosed with high blood pressure, excluding borderline or pregnancy-related cases.",
  "Lung cancer is defined as lung and bronchus cancer for ages 15 and over, risk-adjusted.",
  "Source: Chicago Health Atlas, Chicago Department of Public Health (2022-2023) and Illinois Department of Public Health, (2022-2023).", "Tabulations by Great Cities Institute"
))



table_4f
```


